### 2023-11-17 paper/relationaltypes.tex
add initial rough outline/draft for paper

% ICFP - two-column (sigplan) format
\documentclass[sigplan,screen,review]{acmart}
% \documentclass[letterpaper]{llncs}
% \usepackage[letterpaper, margin=1.5in]{geometry}

\usepackage{multicol}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{nccmath}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage[scaled]{beramono}
\usepackage[T1]{fontenc}

\usepackage{graphicx}
\graphicspath{ {./images/} }

\usepackage{url}

\makeatletter % allow us to mention @-commands
\def\arcr{\@arraycr}
\makeatother

\lstset{
    % identifierstyle=\color{violet},
    % textcolor=blue,
    % keywordstyle=\color{blue},
    % keywordstyle=\text,
    basicstyle=\ttfamily\small,
    % mathescape=true,
    % showspaces=false,
    % morekeywords={let, fix, in}
}

\title{Completing Programs with Relational Types}
% T1 <: T2 could be (x, y) : T1 |- (x, y) : T2 
% \author{}
% \date{}

\begin{document}

% \newcommand{\code}[1]{\texttt{\small #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\claim}{\ \vdash\ }
\newcommand{\hastype}{\ \ddagger\ }
\newcommand{\subtypes}{\ \sqsubseteq\ }
\newcommand{\I}{\hspace{4mm}}
\newcommand{\B}[1]{\textbf{#1}}
\newcommand{\F}[1]{\text{#1}}
\newcommand{\bigand}{\bigwedge\nolimits}
\newcommand{\C}[1]{\color{teal} \rhd\ \emph{#1}}
% \newcommand{\fig}[1]{Fig. {\color{red} \fig{#1}}}
\newcommand{\FIG}[1]{Fig. {\color{red} \ref{#1}}}
\newcommand{\TODO}[1]{\noindent \textbf{\color{red} TODO: #1}}

\maketitle

\section{Introduction}

\subsection{Context}
\TODO{blah blah blah}
- attribute grammars/rules
- bidirectional typing
- horn clauses 
- predicate abstraction

\subsection{Gap}
\TODO{blah blah blah}

\subsection{Innovation}
This work introduces a technique for guiding the completion of untyped programs
by inferring the expectation of next part of the program from the
portion completed thus far. The expectation may represent the next terminal, symbol, or 
nonterminal necessary to construct a correct program. If the expectation is an expression 
nonterminal, then an expected type is also generated. The expected type represents an upper bound
on the value of the next portion to be constructed. The types are expressive enough to represent relations 
between values rather than merely the shape of values. Thus, a \textit{relational type} can be viewed 
as a structured representation of a predicate defined by horn clauses.

This work offers the first technique for inferring relations from incomplete untyped programs. 
The the object under analysis is a incomplete program constructed by a human or a human-like agent in concrete syntax. Thus, 
the analysis must be able to operate without a abstract syntax tree. Instead, we implement the analysis 
as \textit{attribute rules}, interwoven with top-down parsing rules to the form an \textit{attribute grammar}. 
Due to the left-recursion problem in top-down parsing, the parser consolidates information in right-associative manner.
Thus, one difference of analysis in a top-down parsing stack vs an recursion over an AST is re-associating program information 
to the left on the fly. One example of this would be the association of a chain of record selections, e.g.  
\[ 
x.uno.dos.tres 
\]

\[ 
x(.uno(.dos(.tres)))
\]

\[ 
((x.uno).dos).tres
\]

The automatic analysis of incomplete programs generating expected relations during top-down parsing is novel. 
Our technique leverages bidirectional typing along with intersection and union types. Union types are necessary in
the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \text{duality} method
of solving horn clauses.

We have designed a structured language of types in order to express relations and shapes succinctly.
Although horn clauses would be sufficient to express the same semantics, relational types make type annotations easier to write
and inferred types easier to read.

The analysis handles the semantics at two levels: typing and subtyping. Typing is used when the value of an expression
is syntactically represented by an expression. When types or expressions cannot be decomposed easily, more advanced
reasoning techniques are necessary, and the problem is lifted to subtyping. To handle subtyping in a rich language that includes  
relations, our technique encodes relational types into horn clauses and leverages the capabilities of established solvers.

\TODO{blah blah blah}

\TODO{more detail on innovation but still at a high level}

\section{Overview}

\TODO{blah blah blah}

\section{Language}
The programming language is pure and functional. Its syntax and dynamic semantics 
are fairly standard. The main departure from tradition is that its function
and application rules subsume pattern matching. This departure enables a more direct  
correspondence between the structures of programs and their types, but it is 
not a necessary condition.
The syntax is given in \FIG{fig:syntax}.
It includes functions with pattern matching, records, a fixed point combinator, let binding, 
tags for discriminating cases, and application.
A function consists of a sequence of paths, where each path maps a pattern to an expression.
A record consists of a sequence of fields, where each field maps a unique label to an expression. 
The type language includes tag types, field types, implications, unions, intersections, 
inductions, existentials, universals, top, and bottom. 
The existential type consists of multiple bound variables, a payload containing the bound variables, 
and a subtyping constraint over the bound variables. 
If the bound variables aren't indicated, then all variables in the payload are
assumed to be bound variables. If the subtyping constraint isn't indicated, then it is assumed to 
be a tautology, such as (\code{unit<:unit}).
The universal type consists of a bound variable, the variable's upper bound, and a payload. If the upper
bound is not indicated, it is assumed to be the top type (\code{top}).
The typing semantics rely on a typing environment for keeping track of typings of term variables.
The subtyping semantics rely on subtyping environment for keeping track of and constraints on type variables. 
Note that the syntax of the subtyping environment allows an upper bound constraint over a type
rather than merely a type variable to allow for relational constraints.

\TODO{update tag syntax: \code{cons;cons;e:cons//cons//T}\ \ \code{nil;():nil//unit}}

\subsection{Types}

\paragraph{Quantified types}
A universal type is a second order type with universally quantified type variables. 
An existential type is a second order type with existentially quantified type variables.

System F-style. Parameterization of types indexed by types (i.e. second order).

\TODO{mention somwhere that the second order quantification serves two distinct purposes; 1. polymorphism as in System-F. 2. refinement as in first-order quantification of liquid types.
Relational types is able to leverage second-order quantification for refinement, eschewing the first-order quantification used in other systems.
}

\paragraph{Combination types.}
One of the advantages of untyped programs is that they may be written in a flexible manner.
Subtyping is necessary safely reflect the flexibility of compositions in programs, without too many false failures.
Another main advantage of untyped programs is that users don't have to provide type specifications.
Thus, a general way of constructing types from compositions encountered in the the program is necessary.
Some compositions indicate that a type should strengthen, and some compositions indicate that a type should weaken.
To this end, the type language uses intersection and union combinators, 
whose semantics are degenerate versions of those in set-theory.

For instance, when inferring the type of a function, 
the system's goal is to infer the weakest valid parameter type and the strongest valid return type for a function definition.
It strengthens the parameter type with intersection and weakens the return type with union according to the function body,
to arrive at a valid type for the function.

By contrast, the liquid type language relies on the less flexible tagged unions of ML datatypes, 
which is sufficient in its setting since those types are specified by the user. 
Likewise, it does not rely on union to weaken to a valid return type. 
Instead, it weakens to the strongest valid return type by dropping conjunctions from 
the return type's qualifiers until a valid return type is found.

\paragraph{Inductive types.} Similar to ML datatypes.

\paragraph{Qualified types.}
In addition to expressing the shapes of terms, the system should be able express relations between terms,
such as "a list has the length of some natural number".
Rather than using a distinct syntax for relational predicates, 
the type language treats relations as just another type thereby reusing machinery already 
available for types, such as existential types, union types, and inductive types.
Since parametric types are second order, constraining relations requires subtyping.
Thus, parametric types are extended with constraints in the form of subtyping.

\subsection{Expressions}
\TODO{blah blah blah}

\subsection{Typing}
The typing is given in \FIG{fig:typing}.
Most of the rules are fairly standard. 
The rule for function typing is a bit special in that
it treats a function as a sequence of paths whose type is an intersection of implications,
rather than having a separate pattern matching rule. 
Likewise, the type of a record is an intersection of field types.
The let-binding rule allows for prenex polymorphism by generalizing via subtyping.

\subsection{Subtyping}
The subtyping is given in \FIG{fig:subtyping}.
Some of the rules are fairly standard, including implication, the union rules, and intersection rules.
Note that in addition to left and right rules, union and intersection each have rules for 
interacting with implication's antecedent and consequent, respectively.
The constraint rule checks that a subtyping relation exists as a constraint in the subtyping environment.
The right induction rule is standard and simply unrolls the induction.
The left induction rule relies on the induction principle to construct an 
inductive constraint hypothesis.  
The field and tag rules simply check that the labels match and subtyping holds for their constituent types.
The existential rules are quite special, as they involve a subtyping constraint as part of 
a second-order comprehension. 
The left existential rule checks that subtyping holds for all variations of the payload 
where the subtyping constraint holds. 
The right existential rule checks that subtyping holds for some variation of the payload where the constraint holds.
The left universal rule checks that subtyping holds for some variation of the payload consistent with
the variable's upper bound. 
The right universal rule checks that subtyping holds for all variations of the payload consistent with
the variable's upper bound.

\section{Analysis}

The analysis consists of two main parts. The top level is type inference, which corresponds to typing
and generates a type for an expression. When type inference encounters constraints that its types must adhere to,
it calls unification to solve these constraints. Note that since the types are expressive enough to represent constraints,
an alternative approach of generating constraints and solving them in separate stages could also be designed 
using the same structures. Additional structures for the analysis are given in \FIG{fig:internal}. 
Inference generates a solution set $T$, which contains triples, each with a type variable set, a subtyping environment, and a type.  
Unification generates a solution set $C$, which contains subtyping environments.

\subsection{Type Inference}
\TODO{blah blah blah}

\subsection{Subtype Solving}
\TODO{type unification}
\TODO{econding to/decoding from horn clauses}

\section{Experiments}

\TODO{develop 12 tree/list experiments}

\section{Related work}

\TODO{blah blah blah}

\end{document}

### 2023-11-17 paper/relationaltypes.tex
update paper

@@
 \subsection{Context}
-\TODO{blah blah blah}
-- attribute grammars/rules
-- bidirectional typing
-- horn clauses 
-- predicate abstraction
+Ensuring the correctness of programs is a persistent challenge that has been 
+mitigated by formal specifications and automated reasoning tools.
+
+Much research has been devoted to automatically verifying the correctness of programs with simple types.
+One of the main tricks for handling the challenge of correctness has been
+to design languages to make verifying correctness easier. Languages, such as Java and ML, 
+are \textit{intrinsically typed}, 
+requiring nearly all terms to be associated with some type specified by the user. 
+The clever design of ML allows annotations to be fairly sparse by 
+having types specified at constructor definitions and relying on type inference elsewhere.
+
+Despite the advantages of intrinsically typed languages,
+untyped or \textit{extrinsically typed} languages, 
+such as Javascript/Typescript and Python, have surged in popularity. 
+Untyped languages place less initial burden on the programmer to define the upper bounds  
+of various parts of their program and allow reusing code in more flexible ways.
+
+The more expressive the static properties the better the stronger the notion of correctness. 
+Simple types found in ML or Java cannot express relations between values.
+To this end, researchers have extended the simple types found in Java and ML into 
+\textit{refinement types}, \textit{predicate subtyping}, and \textit{dependent types}. 
+Correctness of a program can also be lifted into a purely logical form, such as \textit{horn clauses}.
+
+Refinement types offer greater precision than simple types, but still rely on intrinsic type specifications.
+Dependent types can express detailed relations, but may require users to provide proofs along with detailed annotations.
+Predicate subtyping offers some of the expressivity of dependent types, but with the automatic subtyping of refinement types.
+All of these techniques are based on intrinsic typing and therefore require users to provide additional specifications 
+beyond the runtime behavior of their programs.
+
+Another challenge is guiding the construction of correct programs.
+Subparts of a program can be guided locally with \textit{bidirectional typing}, which decomposes 
+types and propagates expected types down the syntax tree to guide local subparts of the program. 
+Another issue in incomplete programs are the scenarios where the abstract syntax tree cannot be built, 
+do to an incomplete expression represented concretely. In such a case, the analysis can be combined with
+a top-down parser, forming an \textit{attribute grammar}.
 
 \subsection{Gap}
-\TODO{blah blah blah}
+To the best of our knowledge, the problem of guiding the completion of incomplete and untyped programs with relational information  
+has not yet been tackled. The
+
+\textit{NSG} \cite{} offers a technique for guiding program completion by extracting information 
+from incomplete Java programs to guide the next portion. The analysis for extracting
+information was implemented in the form \textit{attribute rules} mixed in with parsing rules
+to form an \textit{attribute grammar}. However, the information extracted is quite 
+simple and directly available in the syntax of programs due to the intrinsic typing and required
+type annotations of Java. There is no guidance based on relations.
+
+Synquid \cite{} tackles the problem of synthesis from a formal specification, rather than completion from the left. 
+Since Synquid can construct its program from scratch, its analysis works directly on an AST and it can avoid
+using a complex attribute grammar.
+
+Synquid's guidance can express relations but it requires an intrinsically typed language with 
+an universe of logical qualifiers provided, which it leverages to perform a combination of 
+\textit{Hindley-Milner type inference} and \textit{predicate abstraction}.
 
 \subsection{Innovation}
-This work introduces a technique for guiding the completion of untyped programs
-by inferring the expectation of next part of the program from the
-portion completed thus far. The expectation may represent the next terminal, symbol, or 
-nonterminal necessary to construct a correct program. If the expectation is an expression 
+This work introduces the first technique for guiding the completion of untyped programs with relational information.
+
+It is novel in inferring relations without intrinsically typed or provided universes of relational qualifiers 
+in conjunction with performing such an analysis during top-down parsing.
+
+The guidance produces a description of the expectation of the next part of the program.
+The expectation may represent a terminal, a symbol, or 
+a nonterminal necessary to construct a correct program. If the expectation is an expression 
 nonterminal, then an expected type is also generated. The expected type represents an upper bound
@@
 
-This work offers the first technique for inferring relations from incomplete untyped programs. 
-The the object under analysis is a incomplete program constructed by a human or a human-like agent in concrete syntax. Thus, 
-the analysis must be able to operate without a abstract syntax tree. Instead, we implement the analysis 
-as \textit{attribute rules}, interwoven with top-down parsing rules to the form an \textit{attribute grammar}. 
-Due to the left-recursion problem in top-down parsing, the parser consolidates information in right-associative manner.
-Thus, one difference of analysis in a top-down parsing stack vs an recursion over an AST is re-associating program information 
-to the left on the fly. One example of this would be the association of a chain of record selections, e.g.  
+The the object under analysis is a incomplete program constructed by a human or a human-like agent in concrete syntax. 
+Thus, the analysis must be able to operate without a abstract syntax tree. Instead, we implement the analysis 
+as \textit{attribute rules}, interwoven with top-down parsing rules to form an \textit{attribute grammar}.
+
+Using attribute rules during parsing adds many complications that would not be present for an
+analysis over an AST. 
+The analysis must be separated rules for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
+Additionally, due to the left-recursion problem in top-down parsing, 
+the parser consolidates information in a right-associative manner.
+Thus, one difference of the analysis over a top-down parsing stack compare to recursion over an AST is 
+the re-associating of program information from right to the left on the fly. 
+One example of this would be the association of a chain of record selections, e.g.  
 \[ 
@@
 \]
+
+\noindent is parsed as
 
@@
 
+\noindent but represents the expression
+
 \[ 
@@
 
-The automatic analysis of incomplete programs generating expected relations during top-down parsing is novel. 
-Our technique leverages bidirectional typing along with intersection and union types. Union types are necessary in
+Our technique leverages bidirectional typing along with intersection and union types. Union types are necessary due to
 the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
@@
 
-\TODO{blah blah blah}
-
-\TODO{more detail on innovation but still at a high level}
-
 \section{Overview}
@@
 \section{Language}
-The programming language is pure and functional. Its syntax and dynamic semantics 
+The language consists of pure expressions and optional type annotations. 
+Its syntax and dynamic semantics 
 are fairly standard. The main departure from tradition is that its function
@@
 not a necessary condition.
-The syntax is given in \FIG{fig:syntax}.
-It includes functions with pattern matching, records, a fixed point combinator, let binding, 
-tags for discriminating cases, and application.
-A function consists of a sequence of paths, where each path maps a pattern to an expression.
-A record consists of a sequence of fields, where each field maps a unique label to an expression. 
-The type language includes tag types, field types, implications, unions, intersections, 
-inductions, existentials, universals, top, and bottom. 
-The existential type consists of multiple bound variables, a payload containing the bound variables, 
-and a subtyping constraint over the bound variables. 
-If the bound variables aren't indicated, then all variables in the payload are
-assumed to be bound variables. If the subtyping constraint isn't indicated, then it is assumed to 
-be a tautology, such as (\code{unit<:unit}).
-The universal type consists of a bound variable, the variable's upper bound, and a payload. If the upper
-bound is not indicated, it is assumed to be the top type (\code{top}).
-The typing semantics rely on a typing environment for keeping track of typings of term variables.
-The subtyping semantics rely on subtyping environment for keeping track of and constraints on type variables. 
-Note that the syntax of the subtyping environment allows an upper bound constraint over a type
-rather than merely a type variable to allow for relational constraints.
-
-\TODO{update tag syntax: \code{cons;cons;e:cons//cons//T}\ \ \code{nil;():nil//unit}}
 

### 2023-11-18 paper/relationaltypes.tex
some small changes to paper

@@
 \subsection{Context}
-Ensuring the correctness of programs is a persistent challenge that has been 
-mitigated by formal specifications and automated reasoning tools.
-
 Much research has been devoted to automatically verifying the correctness of programs with simple types.
@@
 
-The more expressive the static properties the better the stronger the notion of correctness. 
+The more expressive the static properties the stronger the notion of correctness. 
 Simple types found in ML or Java cannot express relations between values.
@@
 Subparts of a program can be guided locally with \textit{bidirectional typing}, which decomposes 
-types and propagates expected types down the syntax tree to guide local subparts of the program. 
+types and propagates expected types down the syntax tree to guide subparts of the program. 
 Another issue in incomplete programs are the scenarios where the abstract syntax tree cannot be built, 
-do to an incomplete expression represented concretely. In such a case, the analysis can be combined with
+due to an incomplete expression represented concretely. In such a case, the analysis can be combined with
 a top-down parser, forming an \textit{attribute grammar}.
@@
 To the best of our knowledge, the problem of guiding the completion of incomplete and untyped programs with relational information  
-has not yet been tackled. The
+has not yet been tackled.
 
@@
 from incomplete Java programs to guide the next portion. The analysis for extracting
-information was implemented in the form \textit{attribute rules} mixed in with parsing rules
+information takes the form of \textit{attribute rules} mixed in with parsing rules
 to form an \textit{attribute grammar}. However, the information extracted is quite 
@@
 
-Synquid \cite{} tackles the problem of synthesis from a formal specification, rather than completion from the left. 
+\textit{Synquid} \cite{} tackles the problem of synthesis from a formal specification, rather than completion from the left. 
 Since Synquid can construct its program from scratch, its analysis works directly on an AST and it can avoid
@@
 Synquid's guidance can express relations but it requires an intrinsically typed language with 
-an universe of logical qualifiers provided, which it leverages to perform a combination of 
+a universe of logical qualifiers provided, which it leverages to perform a combination of 
 \textit{Hindley-Milner type inference} and \textit{predicate abstraction}.
@@
 This work introduces the first technique for guiding the completion of untyped programs with relational information. 
-
-It is novel in inferring relations without intrinsically typed or provided universes of relational qualifiers 
+It is novel in inferring relations without intrinsic types or declared universes of relational qualifiers 
 in conjunction with performing such an analysis during top-down parsing.
 
-The guidance produces a description of the expectation of the next part of the program.
+The analysis produces a description of the expectation for the next part of the program.
 The expectation may represent a terminal, a symbol, or 
@@
 nonterminal, then an expected type is also generated. The expected type represents an upper bound
-on the value of the next portion to be constructed. The types are expressive enough to represent relations 
+on the possible values of the next portion to be constructed. The types are expressive enough to represent relations 
 between values rather than merely the shape of values. Thus, a \textit{relational type} can be viewed 
@@
 
-The the object under analysis is a incomplete program constructed by a human or a human-like agent in concrete syntax. 
-Thus, the analysis must be able to operate without a abstract syntax tree. Instead, we implement the analysis 
+The object under analysis is a incomplete program in concrete syntax constructed by a human or a human-like agent. 
+Thus, the analysis must be able to operate without an abstract syntax tree. Instead, we implement the analysis 
 as \textit{attribute rules}, interwoven with top-down parsing rules to form an \textit{attribute grammar}.
@@
 analysis over an AST. 
-The analysis must be separated rules for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
+The analysis must be separate for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
 Additionally, due to the left-recursion problem in top-down parsing, 
 the parser consolidates information in a right-associative manner.
-Thus, one difference of the analysis over a top-down parsing stack compare to recursion over an AST is 
+Thus, one difference of an analysis over a top-down parsing stack as opposed to recursion over an AST is 
 the re-associating of program information from right to the left on the fly. 
-One example of this would be the association of a chain of record selections, e.g.  
+Consider a chain of record selections:  
 \[ 
@@
 the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
-Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \text{duality} method
+Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \textit{duality} method
 of solving horn clauses.

### 2023-11-20 paper/relationaltypes.tex
update paper on type decomposition

@@
 analysis over an AST. 
-The analysis must be separate for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
+The analysis must be separated into rules for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
 Additionally, due to the left-recursion problem in top-down parsing, 
@@
 
-Our technique leverages bidirectional typing along with intersection and union types. Union types are necessary due to
-the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
-Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \textit{duality} method
-of solving horn clauses.
+In addition to intersection types, seen in refinement type systems, our system also requires union types
+due to the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
+That is, we must be able to start with a strong type and weaken as more use cases are uncovered in the program. 
+In contrast, intrinsic typing explicitly declares a reasonable weakest type, which a refinement type system  
+can strengthen via intersections (or conjunctions).
+
+Additionally, our technique leverages bidirectional typing.
+The inheriting attribute rules include the downward propagation of types. 
+The synthesizing attribute rules include the upward propagation of types.
+
+For introduction rules, our technique propagates types both downward and upward.
+The original bidirectional typing and Synquid's variation \textit{roundtrip typing} only propagate downward for introduction rules. 
+Our problem requires the additional upward propagation because the specification may need to be extracted from the program,
+whereas previous problems assumed a provided specification (e.g. for a function definition).
+For elimination rules, our technique propagates types both upward and downward.
+The original bidirectional typing only propagates upward for elimination rules, but Synquid's variation \textit{roundtrip typing} 
+introduced propagating in both directions for elimination rules.
+Consider application of a function to an argument. Using downward propagation, 
+type propagation can guide the construction of the applicator's function body,
+rather than waiting for the body's full construction and only checking the correctness of the body 
+once the result of the application is used somewhere else.
 
@@
 relations, our technique encodes relational types into horn clauses and leverages the capabilities of established solvers.
+
+Decomposing types into its constituent parts may be necessary when propagating
+types upward from elimination rules or when propagating types downward over introduction rules. 
+These type propagation rules call the solver to find a decomposition that satisfies a subtyping. 
+Thus, solving subtyping provides a generic interface that type propagation can use for type decomposition
+without having to specify the type syntax restrictions directly. 
+In contrast, Synquid's typing rules specify the syntactic forms of types directly.
+Additionally, Synquid is more limited in the syntactic forms of types that it allows.
+For example, our system allows the type of a function to be constructed as an intersection of implications,
+while Synquid requires the function's type to be a single implication (i.e. arrow) type.
+Our system must be more flexible than Synquid due to the specification being extracted from the program,
+rather than specified by the user (as is the case in Synquid).
+
+Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \textit{duality} method
+of solving horn clauses.
 

### 2023-11-24 src/tapas/slim/SlimParser.py
update tests, add piping of exceptions into guidance

@@ -322,7 +322,7 @@ class SlimParser ( Parser ):
                 self.state = 17
                 localctx._ID = self.match(SlimParser.ID)
 
-                localctx.combo = self.collect(ExprAttr(self._solver, plate).combine_id, (None if localctx._ID is None else localctx._ID.text))
+                localctx.combo = self.collect(ExprAttr(self._solver, plate).combine_var, (None if localctx._ID is None else localctx._ID.text))
 
                 pass
 
@@ -390,7 +390,7 @@ class SlimParser ( Parser ):
                 plate_cator = self.guide_nonterm('expr', ExprAttr(self._solver, plate).distill_projmulti_cator)
 
                 self.state = 40
-                localctx.cator = localctx._expr = self.expr(plate_expr)
+                localctx.cator = localctx._expr = self.expr(plate_cator)
 
                 self.guide_symbol(')')
 
@@ -652,7 +652,7 @@ class SlimParser ( Parser ):
                 self.state = 100
                 localctx._ID = self.match(SlimParser.ID)
 
-                localctx.combo = self.collect(PatternAttr(self._solver, plate).combine_id, (None if localctx._ID is None else localctx._ID.text))
+                localctx.combo = self.collect(PatternAttr(self._solver, plate).combine_var, (None if localctx._ID is None else localctx._ID.text))
 
                 pass
 
@@ -1001,7 +1001,7 @@ class SlimParser ( Parser ):
                 plate_body = self.guide_nonterm('expr', RecordAttr(self._solver, plate).distill_single_body, (None if localctx._ID is None else localctx._ID.text))
 
                 self.state = 169
-                localctx.body = self.expr(plate_expr)
+                localctx.body = self.expr(plate_body)
 
                 localctx.combo = self.collect(RecordAttr(self._solver, plate).combine_single, (None if localctx._ID is None else localctx._ID.text), localctx.body.combo)
 

### 2023-11-24 src/tapas/slim/analysis.py

@@ -81,19 +81,19 @@ class ECombo:
 
 
 
-"""
-Pat data types
-"""
+# """
+# Pat data types
+# """
 
-@dataclass(frozen=True, eq=True)
-class PVar:
-    id : str
+# @dataclass(frozen=True, eq=True)
+# class PVar:
+#     id : str
 
-@dataclass(frozen=True, eq=True)
-class PUnit:
-    pass
+# @dataclass(frozen=True, eq=True)
+# class PUnit:
+#     pass
 
-Expr = Union[PVar, PUnit]
+# Expr = Union[PVar, PUnit]
 
 @dataclass(frozen=True, eq=True)
 class PCombo:
@@ -161,8 +161,8 @@ class Attr:
 
 class ExprAttr(Attr):
 
-    def combine_id(self, text : str) -> ECombo:
-        return ECombo(self.plate.enviro[text])
+    def combine_var(self, id : str) -> ECombo:
+        return ECombo(self.plate.enviro[id])
 
     def combine_unit(self) -> ECombo:
         return ECombo(TUnit())
@@ -250,7 +250,7 @@ end ExprAttr
 
 class PatternAttr(Attr):
 
-    def combine_id(self, id : str) -> PCombo:
+    def combine_var(self, id : str) -> PCombo:
         descrip = self.solver.fresh_type_var()
         enviro = m().set(id, descrip)
         interp = self.solver.solve(self.plate.interp, descrip, self.plate.prescrip)

### 2023-11-24 src/tapas/slim/server.py

@@ -29,10 +29,9 @@ class Done:
     pass
 
 I = Union[Kill, str]
-O = Union[Guidance, AttributeError, Killed, Done]
+O = Union[Guidance, Exception, Killed, Done]
 
 async def _mk_task(parser : SlimParser, input : Queue[I], output : Queue[O]) -> Optional[SlimParser.ExprContext]:
-    none : Any = None
     parser.init()
     # parser.buildParseTrees = False
     code = ''
@@ -62,8 +61,8 @@ async def _mk_task(parser : SlimParser, input : Queue[I], output : Queue[O]) -> 
             else:
                 await output.put(parser.getGuidance())
 
-        except AttributeError as e : 
-            print(f"OOGA EXCEPTION: {type(e)} // {e.args}")
+        except Exception as e : 
+            print(f"EXCEPTION: {type(e)} // {e.args}")
             await output.put(e)
             ctx = None
 

### 2023-11-27 paper/relationaltypes.tex
update tests, add piping of exceptions into guidance
add parsing for if-then-else and funnel pipeline

@@
 
-The object under analysis is a incomplete program in concrete syntax constructed by a human or a human-like agent. 
-Thus, the analysis must be able to operate without an abstract syntax tree. Instead, we implement the analysis 
-as \textit{attribute rules}, interwoven with top-down parsing rules to form an \textit{attribute grammar}.
-
-Using attribute rules during parsing adds many complications that would not be present for an
-analysis over an AST. 
-The analysis must be separated into rules for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
-Additionally, due to the left-recursion problem in top-down parsing, 
-the parser consolidates information in a right-associative manner.
-Thus, one difference of an analysis over a top-down parsing stack as opposed to recursion over an AST is 
-the re-associating of program information from right to the left on the fly. 
-Consider a chain of record selections:  
-\[ 
-x.uno.dos.tres 
-\]
-
-\noindent is parsed as
-
-\[ 
-x(.uno(.dos(.tres)))
-\]
-
-\noindent but represents the expression
-
-\[ 
-((x.uno).dos).tres
-\]
-
 In addition to intersection types, seen in refinement type systems, our system also requires union types
@@
 rather than specified by the user (as is the case in Synquid).
+
+The object under analysis is a incomplete program in concrete syntax constructed by a human or a human-like agent. 
+Thus, the analysis must be able to operate without an abstract syntax tree. Instead, we implement the analysis 
+as \textit{attribute rules}, interwoven with top-down parsing rules to form an \textit{attribute grammar}.
+
+Using attribute rules during parsing adds many complications that would not be present for an
+analysis over an AST. 
+The analysis must be separated into rules for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
+Additionally, due to the left-recursion problem in top-down parsing, 
+the parser consolidates information in a right-associative manner.
+Thus, one difference of an analysis over a top-down parsing stack as opposed to recursion over an AST is 
+the re-associating of program information from right to the left on the fly. 
+Consider a chain of record projections:  
+\[ 
+x.uno.dos.tres 
+\]
+
+\noindent is parsed as
+
+\[ 
+x(.uno(.dos(.tres)))
+\]
+
+\noindent but represents the expression
+
+\[ 
+((x.uno).dos).tres
+\]
+
+Propagating types downward over a top-down parse-tree results in a varying interpretations of the downwardly propagated type used for local guidance.
+For syntax that represents complete expressions, e.g. \[ :uno @ \], the inherited type represents the expected type for that expression.
+However, for syntax that cannot represent a complete expression, the inherited type must represent something else. 
+For instance, in the parse tree for a chain of projections, a syntactic subtree of projections, such as \[ .dos.tres \] does not constitute a complete expression.   
+There is no type for this syntax. Thus, the inherited type represents the type that \[ .dos \] is expected to project from (or cut into), 
+e.g. the descriptive type of \[ x.uno \].
 

### 2023-11-27 src/tapas/slim/.antlr/SlimLexer.java

@@ -29,7 +29,8 @@ public class SlimLexer extends Lexer {
 		new PredictionContextCache();
 	public static final int
 		T__0=1, T__1=2, T__2=3, T__3=4, T__4=5, T__5=6, T__6=7, T__7=8, T__8=9, 
-		T__9=10, T__10=11, T__11=12, ID=13, INT=14, WS=15;
+		T__9=10, T__10=11, T__11=12, T__12=13, T__13=14, T__14=15, T__15=16, ID=17, 
+		INT=18, WS=19;
 	public static String[] channelNames = {
 		"DEFAULT_TOKEN_CHANNEL", "HIDDEN"
 	};
@@ -41,22 +42,23 @@ public class SlimLexer extends Lexer {
 	private static String[] makeRuleNames() {
 		return new String[] {
 			"T__0", "T__1", "T__2", "T__3", "T__4", "T__5", "T__6", "T__7", "T__8", 
-			"T__9", "T__10", "T__11", "ID", "INT", "WS"
+			"T__9", "T__10", "T__11", "T__12", "T__13", "T__14", "T__15", "ID", "INT", 
+			"WS"
 		};
 	}
 	public static final String[] ruleNames = makeRuleNames();
 
 	private static String[] makeLiteralNames() {
 		return new String[] {
-			null, "','", "'let'", "';'", "'fix'", "'('", "')'", "'@'", "':'", "'case'", 
-			"'=>'", "'='", "'.'"
+			null, "','", "'if'", "'then'", "'else'", "'let'", "';'", "'fix'", "'('", 
+			"')'", "'@'", "':'", "'case'", "'=>'", "'='", "'|>'", "'.'"
 		};
 	}
 	private static final String[] _LITERAL_NAMES = makeLiteralNames();
 	private static String[] makeSymbolicNames() {
 		return new String[] {
 			null, null, null, null, null, null, null, null, null, null, null, null, 
-			null, "ID", "INT", "WS"
[…]

### 2023-11-27 src/tapas/slim/SlimListener.py

@@ -65,6 +65,15 @@ class SlimListener(ParseTreeListener):
         pass
 
 
+    # Enter a parse tree produced by SlimParser#pipeline.
+    def enterPipeline(self, ctx:SlimParser.PipelineContext):
+        pass
+
+    # Exit a parse tree produced by SlimParser#pipeline.
+    def exitPipeline(self, ctx:SlimParser.PipelineContext):
+        pass
+
+
     # Enter a parse tree produced by SlimParser#keychain.
     def enterKeychain(self, ctx:SlimParser.KeychainContext):
         pass

### 2023-11-27 src/tapas/slim/SlimParser.py

@@ -22,85 +22,98 @@ from pyrsistent.typing import PMap
 
 def serializedATN():
     return [
-        4,1,15,237,2,0,7,0,2,1,7,1,2,2,7,2,2,3,7,3,2,4,7,4,2,5,7,5,2,6,7,
-        6,2,7,7,7,2,8,7,8,2,9,7,9,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,
-        0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,
-        0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,
-        0,1,0,1,0,3,0,65,8,0,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,
-        1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,3,1,92,8,
-        1,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,2,1,
-        2,1,2,1,2,1,2,1,2,1,2,3,2,115,8,2,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,
-        3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,1,3,3,3,138,8,
-        3,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,4,1,
-        4,1,4,3,4,157,8,4,1,5,1,5,1,5,1,5,1,5,1,5,1,5,1,5,1,5,1,5,1,5,1,
-        5,3,5,171,8,5,1,6,1,6,1,6,1,6,1,6,1,6,3,6,179,8,6,1,7,1,7,1,7,1,
-        7,1,7,1,7,1,7,1,7,1,7,1,7,1,7,1,7,3,7,193,8,7,1,8,1,8,1,8,1,8,1,
-        8,1,8,1,8,1,8,1,8,1,8,1,8,1,8,1,8,1,8,1,8,1,8,1,8,3,8,212,8,8,1,
[…]

### 2023-11-27 src/tapas/slim/analysis.py

@@ -80,7 +80,7 @@ Typ = Union[TVar, TUnit, TTag, TField, Inter, Imp, Exis, Induc, Top, Bot]
 
 @dataclass(frozen=True, eq=True)
 class ECombo:
-    descrip : Typ    
+    typ : Typ    
 
 def mk_stack_machine(
     mk_plate : Callable[[T], tuple[list[T], Callable, list[R]]], 
@@ -147,7 +147,7 @@ def concretize_type(typ : Typ) -> str:
 @dataclass(frozen=True, eq=True)
 class PCombo:
     enviro : PMap[str, Typ]
-    descrip : Typ    
+    typ : Typ    
 
 
 """
@@ -177,7 +177,7 @@ Enviro = PMap[str, Typ]
 class Distillation: 
     interp : Interp
     enviro : Enviro 
-    prescrip : Typ
+    typ : Typ
 
 Guidance = Union[Symbol, Terminal, Nonterm]
 
@@ -199,62 +199,83 @@ class Solver:
         '''
         TODO
         '''
+
+        '''
+        TODO:
+        solve cut elimination with rewriting LOWER solving (LOWER <: ARG -> RESULT) for RESULT 
+        e.g. lower = A -> B & C -> D becomes lower = X -> ({B with X <: A} | {D with X <: C} ) 
+        '''
         return interp 
 
 class Attr:
-    def __init__(self, solver : Solver, plate : Distillation):
+    def __init__(self, solver : Solver, distillation : Distillation):
         self.solver = solver
-        self.plate = plate 
+        self.distillation = distillation 
 
 
 
 class BaseAttr(Attr):
 
     def combine_var(self, id : str) -> ECombo:
-        return ECombo(self.plate.enviro[id])
+        return ECombo(self.distillation.enviro[id])
 
     def combine_unit(self) -> ECombo:
         return ECombo(TUnit())
 
     def distill_tag_body(self, id : str) -> Distillation:
-        prescrip = self.solver.fresh_type_var()
-        interp = self.solver.solve(self.plate.interp, TTag(id, prescrip), self.plate.prescrip)
-        return Distillation(interp, self.plate.enviro, prescrip)
+        typ = self.solver.fresh_type_var()
+        interp = self.solver.solve(self.distillation.interp, TTag(id, typ), self.distillation.typ)
+        return Distillation(interp, self.distillation.enviro, typ)
 
     def combine_tag(self, label : str, body : ECombo) -> ECombo:
[…]

### 2023-11-28 paper/relationaltypes.tex
edit a little on spelling and wording about inherited types

@@
 Propagating types downward over a top-down parse-tree results in a varying interpretations of the downwardly propagated type used for local guidance.
-For syntax that represents complete expressions, e.g. \[ :uno @ \], the inherited type represents the expected type for that expression.
+For syntax that represents complete expressions, e.g. \[ x \], the inherited type represents the prescriptive type for that expression.
 However, for syntax that cannot represent a complete expression, the inherited type must represent something else. 
 For instance, in the parse tree for a chain of projections, a syntactic subtree of projections, such as \[ .dos.tres \] does not constitute a complete expression.   
-There is no type for this syntax. Thus, the inherited type represents the type that \[ .dos \] is expected to project from (or cut into), 
+There is no prescriptive type for this syntax. Thus, the inherited type represents the type that \[ .dos \] is expected to project from (or cut into), 
 e.g. the descriptive type of \[ x.uno \].
@@
 \TODO{type unification}
-\TODO{econding to/decoding from horn clauses}
+\TODO{encoding to/decoding from horn clauses}
 

### 2024-02-19 src/tapas/slim/.antlr/SlimLexer.java
add subtyping solving

@@ -31,7 +31,7 @@ public class SlimLexer extends Lexer {
 		T__0=1, T__1=2, T__2=3, T__3=4, T__4=5, T__5=6, T__6=7, T__7=8, T__8=9, 
 		T__9=10, T__10=11, T__11=12, T__12=13, T__13=14, T__14=15, T__15=16, T__16=17, 
 		T__17=18, T__18=19, T__19=20, T__20=21, T__21=22, T__22=23, T__23=24, 
-		T__24=25, T__25=26, T__26=27, T__27=28, ID=29, INT=30, WS=31;
+		T__24=25, T__25=26, T__26=27, T__27=28, T__28=29, ID=30, INT=31, WS=32;
 	public static String[] channelNames = {
 		"DEFAULT_TOKEN_CHANNEL", "HIDDEN"
 	};
@@ -45,17 +45,17 @@ public class SlimLexer extends Lexer {
 			"T__0", "T__1", "T__2", "T__3", "T__4", "T__5", "T__6", "T__7", "T__8", 
 			"T__9", "T__10", "T__11", "T__12", "T__13", "T__14", "T__15", "T__16", 
 			"T__17", "T__18", "T__19", "T__20", "T__21", "T__22", "T__23", "T__24", 
-			"T__25", "T__26", "T__27", "ID", "INT", "WS"
+			"T__25", "T__26", "T__27", "T__28", "ID", "INT", "WS"
 		};
 	}
 	public static final String[] ruleNames = makeRuleNames();
 
 	private static String[] makeLiteralNames() {
 		return new String[] {
-			null, "'@'", "':'", "'('", "')'", "'|'", "'&'", "'->'", "','", "'{'", 
-			"'.'", "'}'", "'['", "']'", "'least'", "'with'", "'greatest'", "'of'", 
-			"'<:'", "'if'", "'then'", "'else'", "'let'", "';'", "'fix'", "'case'", 
-			"'=>'", "'='", "'|>'"
+			null, "'top'", "'bot'", "'@'", "'~'", "':'", "'('", "')'", "'|'", "'&'", 
[…]

### 2024-02-19 src/tapas/slim/.antlr/SlimParser.java

@@ -31,27 +31,27 @@ public class SlimParser extends Parser {
 		T__0=1, T__1=2, T__2=3, T__3=4, T__4=5, T__5=6, T__6=7, T__7=8, T__8=9, 
 		T__9=10, T__10=11, T__11=12, T__12=13, T__13=14, T__14=15, T__15=16, T__16=17, 
 		T__17=18, T__18=19, T__19=20, T__20=21, T__21=22, T__22=23, T__23=24, 
-		T__24=25, T__25=26, T__26=27, T__27=28, ID=29, INT=30, WS=31;
+		T__24=25, T__25=26, T__26=27, T__27=28, T__28=29, ID=30, INT=31, WS=32;
 	public static final int
-		RULE_ids = 0, RULE_typ_base = 1, RULE_typ = 2, RULE_qualification = 3, 
-		RULE_subtyping = 4, RULE_expr = 5, RULE_base = 6, RULE_function = 7, RULE_record = 8, 
-		RULE_argchain = 9, RULE_pipeline = 10, RULE_keychain = 11, RULE_target = 12, 
-		RULE_pattern = 13, RULE_pattern_base = 14, RULE_pattern_record = 15;
+		RULE_ids = 0, RULE_typ_base = 1, RULE_typ = 2, RULE_negchain = 3, RULE_qualification = 4, 
+		RULE_subtyping = 5, RULE_expr = 6, RULE_base = 7, RULE_function = 8, RULE_record = 9, 
+		RULE_argchain = 10, RULE_pipeline = 11, RULE_keychain = 12, RULE_target = 13, 
+		RULE_pattern = 14, RULE_pattern_base = 15, RULE_pattern_record = 16;
 	private static String[] makeRuleNames() {
 		return new String[] {
-			"ids", "typ_base", "typ", "qualification", "subtyping", "expr", "base", 
-			"function", "record", "argchain", "pipeline", "keychain", "target", "pattern", 
-			"pattern_base", "pattern_record"
+			"ids", "typ_base", "typ", "negchain", "qualification", "subtyping", "expr", 
+			"base", "function", "record", "argchain", "pipeline", "keychain", "target", 
+			"pattern", "pattern_base", "pattern_record"
 		};
 	}
 	public static final String[] ruleNames = makeRuleNames();
 
 	private static String[] makeLiteralNames() {
 		return new String[] {
[…]

### 2024-02-19 src/tapas/slim/SlimLexer.py

@@ -22,64 +22,64 @@ from pyrsistent.typing import PMap
 
 def serializedATN():
     return [
-        4,0,31,169,6,-1,2,0,7,0,2,1,7,1,2,2,7,2,2,3,7,3,2,4,7,4,2,5,7,5,
+        4,0,32,171,6,-1,2,0,7,0,2,1,7,1,2,2,7,2,2,3,7,3,2,4,7,4,2,5,7,5,
         2,6,7,6,2,7,7,7,2,8,7,8,2,9,7,9,2,10,7,10,2,11,7,11,2,12,7,12,2,
         13,7,13,2,14,7,14,2,15,7,15,2,16,7,16,2,17,7,17,2,18,7,18,2,19,7,
         19,2,20,7,20,2,21,7,21,2,22,7,22,2,23,7,23,2,24,7,24,2,25,7,25,2,
-        26,7,26,2,27,7,27,2,28,7,28,2,29,7,29,2,30,7,30,1,0,1,0,1,1,1,1,
-        1,2,1,2,1,3,1,3,1,4,1,4,1,5,1,5,1,6,1,6,1,6,1,7,1,7,1,8,1,8,1,9,
-        1,9,1,10,1,10,1,11,1,11,1,12,1,12,1,13,1,13,1,13,1,13,1,13,1,13,
-        1,14,1,14,1,14,1,14,1,14,1,15,1,15,1,15,1,15,1,15,1,15,1,15,1,15,
-        1,15,1,16,1,16,1,16,1,17,1,17,1,17,1,18,1,18,1,18,1,19,1,19,1,19,
-        1,19,1,19,1,20,1,20,1,20,1,20,1,20,1,21,1,21,1,21,1,21,1,22,1,22,
-        1,23,1,23,1,23,1,23,1,24,1,24,1,24,1,24,1,24,1,25,1,25,1,25,1,26,
-        1,26,1,27,1,27,1,27,1,28,4,28,154,8,28,11,28,12,28,155,1,29,4,29,
-        159,8,29,11,29,12,29,160,1,30,4,30,164,8,30,11,30,12,30,165,1,30,
-        1,30,0,0,31,1,1,3,2,5,3,7,4,9,5,11,6,13,7,15,8,17,9,19,10,21,11,
-        23,12,25,13,27,14,29,15,31,16,33,17,35,18,37,19,39,20,41,21,43,22,
-        45,23,47,24,49,25,51,26,53,27,55,28,57,29,59,30,61,31,1,0,3,2,0,
[…]

### 2024-02-19 src/tapas/slim/SlimListener.py

@@ -47,6 +47,15 @@ class SlimListener(ParseTreeListener):
         pass
 
 
+    # Enter a parse tree produced by SlimParser#negchain.
+    def enterNegchain(self, ctx:SlimParser.NegchainContext):
+        pass
+
+    # Exit a parse tree produced by SlimParser#negchain.
+    def exitNegchain(self, ctx:SlimParser.NegchainContext):
+        pass
+
+
     # Enter a parse tree produced by SlimParser#qualification.
     def enterQualification(self, ctx:SlimParser.QualificationContext):
         pass

### 2024-02-19 src/tapas/slim/analyzer.py
add strong var rules

@@ -872,6 +872,16 @@ def inhabitable(t : Typ) -> bool:
         # TODO
         return True
 
+def selective(t : Typ) -> bool:
+    t = simplify_typ(t)
+    if False:
+        pass
+    elif isinstance(t, Top):
+        return False
+    else:
+        # TODO
+        return True
+
 class Solver:
     _type_id : int = 0 
     _battery  : int = -1
@@ -1065,7 +1075,7 @@ class Solver:
                 return self.solve(model, weakest_strong, weak)
             else:
                 strongest = extract_strongest(model, strong.id)
-                if isinstance(strongest, Bot):
+                if not inhabitable(strongest):
                     return [Model(
                         model.constraints.add(Subtyping(strong, weak)),
                         model.freezer
@@ -1100,7 +1110,7 @@ class Solver:
                 return self.solve(model, strong, strongest_weak)
             else:
                 weakest = extract_weakest(model, weak.id)
-                if isinstance(weakest, Top):
+                if not selective(weakest):
                     return [Model(
                         model.constraints.add(Subtyping(strong, weak)),
                         model.freezer

### 2024-02-20 src/tapas/slim/analyzer.py
update subtyping solving rules

@@ -1160,6 +1160,7 @@ class Solver:
                     return solution
 
         elif isinstance(weak, Imp) and isinstance(weak.antec, Unio):
+            # TODO: do we need this rule? 
             '''
             antecedent union: strong <: ((T1 | T2) -> TR)
             A -> Q & B -> Q ~~~ A | B -> Q

### 2024-03-18 src/tapas/slim/.antlr/SlimParser.java
update paper

@@ -1,4 +1,4 @@
-// Generated from /Users/thomas/tlogan/lightweight-tapas/src/tapas/slim/Slim.g4 by ANTLR 4.13.1
+// Generated from /Users/thomas/tlogan@github.com/lightweight-tapas/src/tapas/slim/Slim.g4 by ANTLR 4.13.1
 
 from dataclasses import dataclass
 from typing import *

### 2024-03-25 src/tapas/slim/analyzer.py
update insert_at_path and related procedures

@@ -24,6 +24,16 @@ R = TypeVar('R')
 
 Op = Optional
 
+@dataclass(frozen=True, eq=True)
+class RNode:
+    content : PMap[str, RTree] 
+
+@dataclass(frozen=True, eq=True)
+class RLeaf:
+    content : Typ 
+
+RTree = Union[RNode, RLeaf]
+
 """
 Typ data types
 """
@@ -516,27 +526,53 @@ def factor_path(path : tuple[str, ...], least : LeastFP) -> Typ:
     column = extract_column(path, least.id, choices)
     return column 
 
-def insert_at_path(m : PMap, path : tuple[str, ...], o):
-    if path:
-        key = path[0]  
-        if key in m:
-            assert path[1:]
-            n = m[key] 
-            return m.add({key : insert_at_path(n, path[1:], o)})
+
+def insert_at_path(rnode : RNode, path : tuple[str, ...], t : Typ) -> RNode:
+    assert path
+    key = path[0]  
+    remainder = path[1:]
+
+#     print(f"""
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# DEBUG insert_at_path 
+# ~~ path: {path}
+# ~~ key in rnode.content: {key in rnode.content}
+# ~~ remainder: {remainder}
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+#     """)
+
+    if remainder:
+        if key in rnode.content:
+            assert remainder
+            n = rnode.content[key] 
+            assert isinstance(n, RNode)
         else:
-            return m.add({key : insert_at_path(pmap(), path[1:], o)})
+            n = RNode(m())
+
+        o = insert_at_path(n, remainder, t)
+        content = rnode.content.set(key, o) 
+        return RNode(content)
     else:
-        return o
+        o = RLeaf(t) 
+        content = rnode.content.set(key, o) 
+        return RNode(content)
 
-
-def to_record_typ(m) -> Typ:
+def to_record_typ(rnode : RNode) -> Typ:
     result = Top()
-    for key in m:
-        v = m[key]
-         
-        if isinstance(v, Typ):
-            field = TField(key, v)
+    for key in rnode.content:
+        v = rnode.content[key]
+
+        print(f"""
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
[…]

### 2024-03-25 src/test.py

@@ -955,7 +955,7 @@ let less_equal = {less_equal} ;
 less_equal(~zero @, ~succ ~zero @)
     ''')
     pieces = [max]
-    (combo, guides, parsetree) = analyze(pieces)
+    (combo, guides, parsetree) = analyze(pieces, True)
     print(parsetree)
     # raise_guide(guides)
     # assert combo
@@ -963,14 +963,15 @@ less_equal(~zero @, ~succ ~zero @)
 
 
 if __name__ == '__main__':
+    # test_two_less_equal_one_query()
+    # test_app_less_equal_two_one()
+    test_max()
+    ########################
     # p(less_equal_rel)
     # test_less_equal_imp_subs_two_one_imp_query()
     # test_weak_diff()
     # test_weak_diff_in_pair()
     #########################
-    # test_two_less_equal_one_query()
-    # test_app_less_equal_two_one()
-    ########################
     # test_less_equal()
     # test_fix()
     # test_nil_query_subs_list_nat_diff()
@@ -988,7 +989,7 @@ if __name__ == '__main__':
     # test_two_single_subs_nat_list()
     #####################
     # test_even_list_subs_nat_list()
-    test_nat_list_subs_even_list()
+    # test_nat_list_subs_even_list()
     # test_one_query_subs_nat_list()
     pass
 

### 2024-06-03 src/tapas/slim/analyzer.py
update variable rules and notes on inhabitable check

@@ -1437,9 +1437,9 @@ class Solver:
             return (rev_aliasing, t)
 
 
-    def interpret_lower_id(self, world : World, id : str) -> tuple[Typ, PSet[Subtyping]]:
+    def interpret_as_weakest_lower(self, world : World, id : str) -> tuple[Typ, PSet[Subtyping]]:
         '''
-        for constraints X <: T, X <: U; find weakest type stronger than T, stronger than U
+        for constraints X <: T, X <: U; find weakest type lower than T, lower than U
         which is T & U.
         NOTE: related to weakest precondition concept
         '''
@@ -1454,12 +1454,12 @@ class Solver:
         
         return (simplify_typ(result), pset(constraints))
 
-    def interpret_upper_id(self, world : World, id : str) -> tuple[Typ, PSet[Subtyping]]:
+    def interpret_as_strongest_upper(self, world : World, id : str) -> tuple[Typ, PSet[Subtyping]]:
         # if id in self.aliasing:
         #     return (self.aliasing[id], s())
 
         '''
-        for constraints T <: X, U <: X; find strongest type weaker than T, weaker than U
+        for constraints T <: X, U <: X; find strongest type upper than T, upper than U
         which is T | U.
         NOTE: related to strongest postcondition concept
         '''
@@ -1530,9 +1530,9 @@ class Solver:
     def interpret_with_polarity(self, polarity : bool, world : World, typ : Typ, ignored_ids : PSet[str]) -> tuple[Typ, PSet[Subtyping]]:
         def interpret_id(polarity : bool, id : str): 
             if polarity:
-                return self.interpret_upper_id(world, id)
+                return self.interpret_as_strongest_upper(world, id)
             else:
-                return self.interpret_lower_id(world, id)
+                return self.interpret_as_weakest_lower(world, id)
 
         if False:
             assert False
@@ -1954,7 +1954,9 @@ class Solver:
         result = (
             self.is_inhabitable(world, target) and
             all(
-                self.is_inhabitable(world, legacy) and
+                # NOTE: no need to check legacy; e.g. self.is_inhabitable(world, legacy)
+                # - we know it's inhabitable since it is in the world
+                # - and only the solver adds constraints to the world
                 self.is_intersection_inhabitable(world, legacy, target)
                 for legacy in self.extract_uppers(world, id)[0]
             ) 
[…]

### 2024-06-04 src/test_analysis.py
update

@@ -1686,11 +1686,11 @@ if __name__ == '__main__':
     # test_single_shape()
     # test_implication_unification()
     # test_lted_wrapper()
-    # test_max()
+    test_max()
     # test_max_annotated()
     # test_max_subtyping()
     # test_max_subtyping_fail()
-    test_constrained_universal_subtyping_fail()
+    # test_constrained_universal_subtyping_fail()
     # test_constrained_universal_subtyping_record_pass()
     # test_constrained_universal_subtyping_record_fail()
     # test_constrained_universal_subtyping_function_fail()

### 2024-06-04 src/tapas/slim/analyzer.py
update rules in paper including lower leastFP

@@ -1919,8 +1919,6 @@ class Solver:
                 self.is_intersection_inhabitable(world, left, right.right)
             )
 
-
-
         elif isinstance(left, TField) and isinstance(right, TField): 
             return left.label != right.label or (
                 self.is_intersection_inhabitable(world, left.body, right.body)
@@ -2061,6 +2059,7 @@ upper:
                 strong_body = sub_typ(renaming, lower.body)
                 return self.solve(world, strong_body, upper)
             else:
+                # TODO: consider case non-relational least fixed point
                 '''
                 rewrite into existential making shape of relation visible
                 '''

### 2024-06-10 src/tapas/slim/.antlr/SlimListener.java
update least fixed point typ denotation

@@ -1,4 +1,4 @@
-// Generated from /Users/thomas/tlogan/lightweight-tapas/src/tapas/slim/Slim.g4 by ANTLR 4.13.1
+// Generated from /Users/thomas/tlogan@github.com/ERST/src/tapas/slim/Slim.g4 by ANTLR 4.13.1
 
 from dataclasses import dataclass
 from typing import *
@@ -7,8 +7,8 @@ from contextlib import contextmanager
 
 from tapas.slim.analyzer import * 
 
-from pyrsistent import m, pmap, v
-from pyrsistent.typing import PMap 
+from pyrsistent.typing import PMap, PSet 
+from pyrsistent import m, s, pmap, pset
 
 
 import org.antlr.v4.runtime.tree.ParseTreeListener;
@@ -28,6 +28,26 @@ public interface SlimListener extends ParseTreeListener {
 	 * @param ctx the parse tree
 	 */
 	void exitIds(SlimParser.IdsContext ctx);
+	/**
+	 * Enter a parse tree produced by {@link SlimParser#preamble}.
+	 * @param ctx the parse tree
+	 */
+	void enterPreamble(SlimParser.PreambleContext ctx);
+	/**
+	 * Exit a parse tree produced by {@link SlimParser#preamble}.
+	 * @param ctx the parse tree
+	 */
+	void exitPreamble(SlimParser.PreambleContext ctx);
+	/**
+	 * Enter a parse tree produced by {@link SlimParser#program}.
+	 * @param ctx the parse tree
+	 */
+	void enterProgram(SlimParser.ProgramContext ctx);
+	/**
+	 * Exit a parse tree produced by {@link SlimParser#program}.
+	 * @param ctx the parse tree
+	 */
+	void exitProgram(SlimParser.ProgramContext ctx);
 	/**
 	 * Enter a parse tree produced by {@link SlimParser#typ_base}.
 	 * @param ctx the parse tree
@@ -99,6 +119,16 @@ public interface SlimListener extends ParseTreeListener {
 	 */
 	void exitBase(SlimParser.BaseContext ctx);
 	/**
+	 * Enter a parse tree produced by {@link SlimParser#record}.
+	 * @param ctx the parse tree
+	 */
+	void enterRecord(SlimParser.RecordContext ctx);
+	/**
+	 * Exit a parse tree produced by {@link SlimParser#record}.
+	 * @param ctx the parse tree
+	 */
+	void exitRecord(SlimParser.RecordContext ctx);
+	/**
[…]

### 2024-08-01 src/tapas/slim/.antlr/SlimLexer.java
revert concrete syntax from ANY to EXI; the quantification of "any" in English is ambiguous

@@ -54,7 +54,7 @@ public class SlimLexer extends Lexer {
 	private static String[] makeLiteralNames() {
 		return new String[] {
 			null, "'alias'", "'='", "'TOP'", "'BOT'", "'@'", "'~'", "':'", "'('", 
-			"')'", "'|'", "'&'", "'->'", "','", "'ANY'", "'['", "']'", "'ALL'", "'FX'", 
+			"')'", "'|'", "'&'", "'->'", "','", "'EXI'", "'['", "']'", "'ALL'", "'FX'", 
 			"'\\'", "';'", "'<:'", "'if'", "'then'", "'else'", "'let'", "'in'", "'fix'", 
 			"'case'", "'=>'", "'.'", "'|>'"
 		};
@@ -201,8 +201,8 @@ public class SlimLexer extends Lexer {
 		"\u0000\u0000\u0000_`\u0005|\u0000\u0000`\u0014\u0001\u0000\u0000\u0000"+
 		"ab\u0005&\u0000\u0000b\u0016\u0001\u0000\u0000\u0000cd\u0005-\u0000\u0000"+
 		"de\u0005>\u0000\u0000e\u0018\u0001\u0000\u0000\u0000fg\u0005,\u0000\u0000"+
-		"g\u001a\u0001\u0000\u0000\u0000hi\u0005A\u0000\u0000ij\u0005N\u0000\u0000"+
-		"jk\u0005Y\u0000\u0000k\u001c\u0001\u0000\u0000\u0000lm\u0005[\u0000\u0000"+
+		"g\u001a\u0001\u0000\u0000\u0000hi\u0005E\u0000\u0000ij\u0005X\u0000\u0000"+
+		"jk\u0005I\u0000\u0000k\u001c\u0001\u0000\u0000\u0000lm\u0005[\u0000\u0000"+
 		"m\u001e\u0001\u0000\u0000\u0000no\u0005]\u0000\u0000o \u0001\u0000\u0000"+
 		"\u0000pq\u0005A\u0000\u0000qr\u0005L\u0000\u0000rs\u0005L\u0000\u0000"+
 		"s\"\u0001\u0000\u0000\u0000tu\u0005F\u0000\u0000uv\u0005X\u0000\u0000"+

### 2024-08-01 src/tapas/slim/SlimLexer.py

@@ -62,8 +62,8 @@ def serializedATN():
         0,88,12,1,0,0,0,89,90,5,58,0,0,90,14,1,0,0,0,91,92,5,40,0,0,92,16,
         1,0,0,0,93,94,5,41,0,0,94,18,1,0,0,0,95,96,5,124,0,0,96,20,1,0,0,
         0,97,98,5,38,0,0,98,22,1,0,0,0,99,100,5,45,0,0,100,101,5,62,0,0,
-        101,24,1,0,0,0,102,103,5,44,0,0,103,26,1,0,0,0,104,105,5,65,0,0,
-        105,106,5,78,0,0,106,107,5,89,0,0,107,28,1,0,0,0,108,109,5,91,0,
+        101,24,1,0,0,0,102,103,5,44,0,0,103,26,1,0,0,0,104,105,5,69,0,0,
+        105,106,5,88,0,0,106,107,5,73,0,0,107,28,1,0,0,0,108,109,5,91,0,
         0,109,30,1,0,0,0,110,111,5,93,0,0,111,32,1,0,0,0,112,113,5,65,0,
         0,113,114,5,76,0,0,114,115,5,76,0,0,115,34,1,0,0,0,116,117,5,70,
         0,0,117,118,5,88,0,0,118,36,1,0,0,0,119,120,5,92,0,0,120,38,1,0,
@@ -133,7 +133,7 @@ class SlimLexer(Lexer):
 
     literalNames = [ "<INVALID>",
             "'alias'", "'='", "'TOP'", "'BOT'", "'@'", "'~'", "':'", "'('", 
-            "')'", "'|'", "'&'", "'->'", "','", "'ANY'", "'['", "']'", "'ALL'", 
+            "')'", "'|'", "'&'", "'->'", "','", "'EXI'", "'['", "']'", "'ALL'", 
             "'FX'", "'\\'", "';'", "'<:'", "'if'", "'then'", "'else'", "'let'", 
             "'in'", "'fix'", "'case'", "'=>'", "'.'", "'|>'" ]
 

### 2024-08-01 src/tapas/slim/SlimParser.py

@@ -196,7 +196,7 @@ class SlimParser ( Parser ):
 
     literalNames = [ "<INVALID>", "'alias'", "'='", "'TOP'", "'BOT'", "'@'", 
                      "'~'", "':'", "'('", "')'", "'|'", "'&'", "'->'", "','", 
-                     "'ANY'", "'['", "']'", "'ALL'", "'FX'", "'\\'", "';'", 
+                     "'EXI'", "'['", "']'", "'ALL'", "'FX'", "'\\'", "';'", 
                      "'<:'", "'if'", "'then'", "'else'", "'let'", "'in'", 
                      "'fix'", "'case'", "'=>'", "'.'", "'|>'" ]
 

### 2024-08-01 src/tapas/slim/typlib.py

@@ -43,14 +43,14 @@ even = (f"""
 nat_list = (f"""
 (FX NL 
     | (~zero @, ~nil @) 
-    | (ANY [N L ; (N, L) <: NL] (~succ N, ~cons L))
+    | (EXI [N L ; (N, L) <: NL] (~succ N, ~cons L))
 )
 """.strip())
 
 even_list = (f"""
 (FX NL 
     | (~zero @, ~nil @) 
-    | ANY [N L ; (N, L) <: NL] (~succ ~succ N, ~cons ~cons L)  
+    | EXI [N L ; (N, L) <: NL] (~succ ~succ N, ~cons ~cons L)  
 )
 """.strip())
 
@@ -59,50 +59,50 @@ even_list = (f"""
 nat_equal = (f"""
 (FX SELF 
     | (~zero @, ~zero @) 
-    | ANY [A B ; (A, B) <: SELF] (~succ A, ~succ B)  
+    | EXI [A B ; (A, B) <: SELF] (~succ A, ~succ B)  
 )
 """.strip())
 
 addition = (f'''
 FX AR 
-    | (ANY [Y Z ; (Y, Z) <: ({nat_equal})] (x : ~zero @ & y : Y & z : Z))
-    | (ANY [X Y Z ; (x : X & y : Y & z : Z) <: AR] (x : ~succ X & y : Y & z : ~succ Z))
+    | (EXI [Y Z ; (Y, Z) <: ({nat_equal})] (x : ~zero @ & y : Y & z : Z))
+    | (EXI [X Y Z ; (x : X & y : Y & z : Z) <: AR] (x : ~succ X & y : Y & z : ~succ Z))
 ''')
 
 lte = (f"""
 (FX SELF 
-    | (ANY [x ; x <: ({nat})] (~zero @, x))
-    | (ANY [a b ; (a,b) <: SELF] (~succ a, ~succ b))
+    | (EXI [x ; x <: ({nat})] (~zero @, x))
+    | (EXI [a b ; (a,b) <: SELF] (~succ a, ~succ b))
 )
 """)
 
 open_lte = (f"""
 (FX SELF 
-    | (ANY [x] (~zero @, x))
-    | (ANY [a b ; (a,b) <: SELF] (~succ a, ~succ b))
+    | (EXI [x] (~zero @, x))
+    | (EXI [a b ; (a,b) <: SELF] (~succ a, ~succ b))
 )
 """)
 
 
 lted = (f"""
 (FX self 
[…]

### 2024-08-28 src/tapas/slim/analyzer.py
update flex-l rule

@@ -1935,15 +1935,39 @@ class Solver:
     def extract_transitive_upper_bounds(self, world : World, id : str) -> PSet[Subtyping]:
         return pset( 
             Subtyping(st0.lower, st1.upper) 
-            for st0 in self.extract_upper_bounds(world, id).union(
-                self.extract_rel_upper_bounds(world, id)
-            )
+            for st0 in self.extract_upper_bounds(world, id)
             for st1 in (
                 self.extract_transitive_upper_bounds(world, st0.upper.id)
                 if isinstance(st0.upper, TVar) and st0.upper.id in world.skolems else
                 [st0]
             )
         )
+
+    def extract_rel_transitive_upper_bounds(self, world : World, id : str) -> PSet[Subtyping]:
+        return pset(
+            Subtyping(st0.lower, st1.upper) 
+            for st0 in self.extract_rel_upper_bounds(world, id)
+            for st1 in (
+                self.extract_transitive_upper_bounds(world, st0.upper.id)
+                if isinstance(st0.upper, TVar) and st0.upper.id in world.skolems else
+                [st0]
+            ) 
+        )
+    
+    def extract_lower_skolems(self, world : World, id : str) -> PSet[str]:
+        return pset(
+            st0.lower.id
+            for st0 in self.extract_lower_bounds(world, id)
+            if isinstance(st0.lower, TVar) and st0.lower.id in world.skolems
+        )
+
+    def extract_transitive_lower_skolems(self, world : World, id : str) -> PSet[str]:
+        lids0 = self.extract_lower_skolems(world, id)
+        return lids0.union(pset( 
+            lid1
+            for lid0 in lids0 
+            for lid1 in self.extract_transitive_lower_skolems(world, lid0)
+        ))
 
     def extract_lower_bounds(self, world : World, id : str) -> PSet[Subtyping]:
         return pset(
@@ -1964,21 +1988,6 @@ class Solver:
         )
 
 
-    def extract_lower_skolems(self, world : World, id : str) -> PSet[str]:
-        return pset(
-            st0.lower.id
-            for st0 in self.extract_lower_bounds(world, id)
-            if isinstance(st0.lower, TVar) and st0.lower.id in world.skolems
-        )
-
-
-    def extract_transitive_lower_skolems(self, world : World, id : str) -> PSet[str]:
-        lids0 = self.extract_lower_skolems(world, id)
-        return lids0.union(pset( 
-            lid1
[…]

### 2024-09-12 src/tapas/slim/analyzer.py
update diff rules

@@ -2515,7 +2515,8 @@ class Solver:
         #######################################
 
         elif isinstance(upper, Diff): 
-            if diff_well_formed(upper):
+
+            if diff_well_formed(upper): # TODO: change to: DF(lower) and DF(upper)
                 # TODO: need a sound/safe/conservative inhabitable check
                 # only works if we assume T is not empty
                 '''
@@ -2530,7 +2531,9 @@ class Solver:
                     m
                     for m in context_worlds 
                     if (
-                        not self.is_inhabitable(world, lower) or 
+                        # not self.is_inhabitable(world, lower) or 
+                        # just fail (incompletely) if lower is empty
+                        # we don't care to reason about empty types on the LHS.
                         self.solve(m, lower, upper.negation) == []
                     )
                 ]   

### 2024-09-26 src/tapas/slim/analyzer.py
update syntax for rule for relational type construction

@@ -3163,9 +3163,10 @@ class ExprRule(Rule):
         for outer_world in nt.worlds:
             outer_worlds.append(outer_world)
             inner_worlds = self.solver.solve(outer_world, body_typ, Imp(self_typ, Imp(in_typ, out_typ)))
-
+            # fix(a => (b0 => c0),(b1 => c1),(b2 => c2),...) 
+            # has type: 
+            # FIX[A](intersect worlds: ALL[A <: ...]B -> C)
             induc_body = Bot()
-
             ###########################
             for i, inner_world in enumerate(reversed(inner_worlds)):
                 lefts = pset()

### 2024-09-30 src/tapas/slim/analyzer.py
update notes on relation of skolems to bound ids in combine_fix

@@ -3171,18 +3171,16 @@ class ExprRule(Rule):
             for i, inner_world in enumerate(reversed(inner_worlds)):
                 lefts = pset()
                 left_typ = in_typ  
+                # assert in_typ.id not in inner_world.relids
+                ###### TODO: ensure that the assertion is invariant
                 if in_typ.id not in inner_world.relids:
                     lefts = self.solver.extract_upper_bounds(inner_world, in_typ.id)
                     left_typ = make_inter(list(lefts))
                 # end if
+                ###########################
                 rights = self.solver.extract_lower_bounds(inner_world, out_typ.id)
                 right_typ = make_unio(list(rights))
 
-                left_bound_ids = extract_free_vars_from_typ(s(), left_typ)
-                right_bound_ids = extract_free_vars_from_typ(s(), right_typ)
-                bound_ids = left_bound_ids.union(right_bound_ids).union(
-                    inner_world.skolems.intersection(extract_free_vars_from_constraints(s(), inner_world.constraints))
-                )
                 rel_pattern = make_pair_typ(left_typ, right_typ)
                 fixpoint_interps = self.solver.extract_upper_bounds(inner_world, self_typ.id)
                 IH_rel_constraints = s()
@@ -3213,16 +3211,28 @@ class ExprRule(Rule):
                     [IH_typ.id] 
                 ) 
 
-                rel_reachable_constraints = pset(
+                influential_constraints = pset(
                     st
                     # for st in extract_reachable_constraints_from_typ(inner_world, rel_pattern)
                     for st in inner_world.constraints
+                    # NOTE: contains at least one influential variable
                     if not bool(extract_free_vars_from_constraints(s(), [st]).difference(influential_vars))
                     # if not bool(pset([self_typ.id, in_typ.id, out_typ.id])
                     #     .intersection(extract_free_vars_from_typ(s(), st.lower).union(extract_free_vars_from_typ(s(), st.upper)))
                     # )
                 )
-                rel_constraints = IH_rel_constraints.union(rel_reachable_constraints)
+
+                # TODO: make sure skolems are considered properly
+                # TODO: perhaps skolems should be universally bound on the far outside.
+                bound_ids = extract_free_vars_from_typ(s(), rel_pattern).union(
+                    extract_free_vars_from_constraints(s(), influential_constraints)
+                    # .intersection(inner_world.skolems)
+                    # .difference(inner_world.skolems)
+                )
[…]

### 2024-10-01 src/tapas/slim/analyzer.py
update rules for typing recursive function

@@ -3204,14 +3204,10 @@ class ExprRule(Rule):
             induc_body = Bot()
             ###########################
             for i, inner_world in enumerate(reversed(inner_worlds)):
-                lefts = pset()
-                left_typ = in_typ  
-                # assert in_typ.id not in inner_world.relids
                 ###### TODO: ensure that the assertion is invariant
-                if in_typ.id not in inner_world.relids:
-                    lefts = self.solver.extract_upper_bounds(inner_world, in_typ.id)
-                    left_typ = make_inter(list(lefts))
-                # end if
+                assert in_typ.id not in inner_world.relids
+                lefts = self.solver.extract_upper_bounds(inner_world, in_typ.id)
+                left_typ = make_inter(list(lefts))
                 ###########################
                 rights = self.solver.extract_lower_bounds(inner_world, out_typ.id)
                 right_typ = make_unio(list(rights))

### 2024-10-02 src/tapas/slim/analyzer.py
update rules for typing of function

@@ -2973,6 +2973,7 @@ class BaseRule(Rule):
             for t in nt.enviro.values()
             for id in extract_free_vars_from_typ(s(), t)
         )
+        # TODO: move augmentation outside
         augmented_branches = self.solver.augment_branches_with_diff(branches)
         generalized_branches = []
         for branch in augmented_branches: 
@@ -2981,6 +2982,7 @@ class BaseRule(Rule):
             imp = Imp(param_typ, body_typ)
             for branch_world in branch.worlds:
                 influential_vars = rigids.union(
+                    # TODO: reconsider if skolems should be considered influential
                     branch_world.skolems
                 ).union(
                     extract_free_vars_from_typ(s(), imp)
@@ -2991,6 +2993,7 @@ class BaseRule(Rule):
                     for st in branch_world.constraints
                     if not bool(extract_free_vars_from_constraints(s(), [st]).difference(influential_vars))
                 )
+                #TODO: replace with make_constraint function
                 universal_constraints = extract_universal_constraints(branch_world.skolems, reachable_constraints)
                 reachable_ids = extract_free_vars_from_constraints(s(), reachable_constraints).union(extract_free_vars_from_typ(s(), imp))
                 universal_bound_ids = tuple(sorted(reachable_ids.difference(branch_world.skolems).difference(rigids)))

### 2024-10-03 src/tapas/slim/analyzer.py
update combine_function to use make_constraint_typ

@@ -1292,24 +1292,6 @@ def extract_reachable_constraints_from_typ(world : World, typ : Typ, debug = Fal
             constraints = constraints.union(new_constraints)
     return constraints
 
-def extract_existential_constraints(skolems: PSet[str], constraints : PSet[Subtyping]) -> PSet[Subtyping]:
-    return pset( 
-        st
-        for st in constraints
-        for lower_fvs in [extract_free_vars_from_typ(s(), st.lower)]
-        for upper_fvs in [extract_free_vars_from_typ(s(), st.upper)]
-        if bool(skolems.intersection(lower_fvs.union(upper_fvs))) 
-    )
-
-def extract_universal_constraints(skolems: PSet[str], constraints : PSet[Subtyping]) -> PSet[Subtyping]:
-    return pset( 
-        st
-        for st in constraints
-        for lower_fvs in [extract_free_vars_from_typ(s(), st.lower)]
-        for upper_fvs in [extract_free_vars_from_typ(s(), st.upper)]
-        if bool(lower_fvs.union(upper_fvs).difference(skolems)) 
-    )
-
 # def package_typ(world : World, typ : Typ) -> Typ:
 #     constraints = extract_reachable_constraints_from_typ(world, typ)
 #     existential_constraints = extract_existential_constraints(world.skolems, constraints)
@@ -3000,36 +2982,7 @@ class BaseRule(Rule):
                 for st in branch.world.constraints
                 if not bool(extract_free_vars_from_constraints(s(), [st]).difference(influential_vars))
             )
-            #TODO: replace with make_constraint function
-            universal_constraints = extract_universal_constraints(branch.world.skolems, reachable_constraints)
-            reachable_ids = extract_free_vars_from_constraints(s(), reachable_constraints).union(extract_free_vars_from_typ(s(), imp))
-            universal_bound_ids = tuple(sorted(reachable_ids.difference(branch.world.skolems).difference(rigids)))
-
-            if not universal_bound_ids:
-                assert not universal_bound_ids
-                body = imp 
-            else:
-                body = All(universal_bound_ids, tuple(sorted(universal_constraints)), imp)
-            #end if-else
-
-            ######## NOTE: generalization #############
-
-            constraints = (
-                tuple(sorted(reachable_constraints.difference(universal_constraints)))
-            )
-
-            fvs = sorted(
-                extract_free_vars_from_typ(s(), body).union(
-                    extract_free_vars_from_constraints(s(), constraints)
-                ).intersection(branch.world.skolems)
-            )
[…]

### 2024-12-31 src/test_analysis.py
clean up skolem rules

@@ -1980,9 +1980,9 @@ def test_skolem_subtyping_A():
 BOT -> TOP 
     """.strip())
 
-    lower = (f"""
-@ -> @
-    """.strip())
+#     lower = (f"""
+# @ -> @
+#     """.strip())
 
     upper = (f"""
 ALL[X] X -> X
@@ -2000,9 +2000,9 @@ TOP -> BOT
 ALL[X] X -> X
     """.strip())
 
-    upper = (f"""
-@ -> @
-    """.strip())
+#     upper = (f"""
+# @ -> @
+#     """.strip())
     solver = analyzer.Solver(m())
     worlds = solve(solver, lower, upper)
     assert worlds

### 2025-01-06 src/tapas/slim/analyzer.py
clean up code and rules for learning constraints on single variables

@@ -1946,44 +1946,6 @@ class Solver:
     def extract_factored_upper_bounds(self, world : World, id : str) -> PSet[Typ]:
         return find_factors(world, TVar(id))
 
-    def extract_rel_upper_bounds(self, world : World, id : str) -> PSet[Subtyping]:
-        return pset(
-            st 
-            for st in world.constraints
-            if not isinstance(st.lower, TVar) and id in extract_free_vars_from_typ(s(), st.lower)
-        )
-
-    def extract_upper_skolems(self, world : World, id : str) -> PSet[str]:
-        return pset(
-            t0.id
-            for t0 in self.extract_upper_bounds(world, id)
-            if isinstance(t0, TVar) and t0.id in world.skolems
-        )
-
-    def extract_rel_upper_skolems(self, world : World, id : str) -> PSet[tuple[Typ, str]]:
-        return pset(
-            (st.lower, st.upper.id) 
-            for st in self.extract_rel_upper_bounds(world, id)
-            if isinstance(st.upper, TVar) and st.upper.id in world.skolems
-        )
-
-
-    def extract_transitive_upper_skolems(self, world : World, id : str) -> PSet[str]:
-        uids0 = self.extract_upper_skolems(world, id)
-        return uids0.union(pset( 
-            uid1
-            for uid0 in uids0 
-            for uid1 in self.extract_transitive_upper_skolems(world, uid0)
-        ))
-
-    def extract_rel_transitive_upper_skolems(self, world : World, id : str) -> PSet[tuple[Typ, str]]:
-        sts0 = self.extract_rel_upper_skolems(world, id)
-        return sts0.union(pset( 
-            (st0[0], skol)
-            for st0 in sts0 
-            for skol in self.extract_transitive_upper_skolems(world, st0[1])
-        ))
-
     def extract_transitive_upper_bounds(self, world : World, id : str) -> PSet[Typ]:
         return pset( 
             t1 
@@ -1994,34 +1956,6 @@ class Solver:
                 [t0]
             )
         )
-
-    def extract_rel_transitive_upper_bounds(self, world : World, id : str) -> PSet[Subtyping]:
-        return pset(
-            Subtyping(st.lower, upper) 
-            for st in self.extract_rel_upper_bounds(world, id)
-            for upper in (
-                self.extract_transitive_upper_bounds(world, st.upper.id)
-                if isinstance(st.upper, TVar) and st.upper.id in world.skolems else
-                [st.upper]
[…]

### 2025-01-18 src/tapas/slim/analyzer.py
add notes on simplifiying proof typing with a single world as input

@@ -2998,6 +2998,7 @@ class BaseRule(Rule):
         # return Result(nt.worlds, simplify_typ(result))
         return Result(nt.worlds, result)
 
+    # TODO: redo Context such that it only has a single world as input
     def combine_function(self, nt : Context, branches : list[Branch]) -> Result:
         '''
         Example

### 2025-02-07 src/tapas/slim/exprlib.py
remove notes about modifying constraint learning

@@ -63,9 +63,6 @@ case (x, y) => (
 # inner constraints: there is a variable that is open and local
 # 
 
-# TODO: need to update safety subproblems when learning constraints on variables
-# need to consider variables contained in other type forms.
-
 # IMPORTANT NOTE: closed variables can be influenced by open variables
 # as new information on X and Y arrives, the constraint on Z is specialized.
 # as opposed to Z being strict and preventing learning constraints on X and Y.

### 2025-02-14 src/tapas/slim/analyzer.py
update fixed point duality rules in paper with negative and positive interpretations

@@ -2076,11 +2076,18 @@ class Solver:
             assert not bool(foreignids.intersection(closedids))
 
             influential_ids = foreignids.union(closedids).union(payload_ids)
+#             print(f"""
+# ~~~~~~~ DEBUG influential_ids: {influential_ids} ~~~~~~~~~~~~~~~
+#             """)
             constraints = pset(
                 st
                 for st in raw_constraints
-                if not bool(extract_free_vars_from_constraints(s(), [st]).difference(influential_ids))
+
                 # if all fids are influential
+                # if not bool(extract_free_vars_from_constraints(s(), [st]).difference(influential_ids))
+
+                # if at least one fid is influential
+                if bool(extract_free_vars_from_constraints(s(), [st]).intersection(influential_ids))
             )
 
             outer_constraints = pset(
@@ -3093,6 +3100,27 @@ class BaseRule(Rule):
 
     # TODO: redo Context such that it only has a single world as input
     def combine_function(self, pid : int, enviro : Enviro, world : World, function_branches : list[NDBranch]) -> Result:
+        # for i, branch in enumerate(function_branches): 
+#             print(f"""
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# DEBUG combine_function NDBranch
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# pattern {i}: {concretize_typ(branch.pattern)}
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+#             """)
+#             for j, result in enumerate(branch.results):
+#                 print(f"""
+#     ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+#     result {i},{j}: 
+
+#     closedids: {result.world.closedids}
+    
+#     constraints:
+#     {concretize_constraints(result.world.constraints)}
+
+#     typ: {concretize_typ(result.typ)}
+#     ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+#                 """)
         '''
         Example
         ==============
@@ -3259,6 +3287,19 @@ class ExprRule(Rule):
             IH_rel_constraints = s()
             IH_left_constraints = s()
[…]

### 2025-02-14 src/test_analysis.py
update strategies for interpretation and cleaning up constraints

@@ -874,6 +874,8 @@ def test_fix_body():
     print("answer:\n" + analyzer.concretize_typ(result))
 
 def test_fix():
+
+    # TODO: update negative and positive interpretation to remove constraints
     code = '''
 fix(case self => (
     case ~nil @ => ~zero @ 
@@ -882,14 +884,16 @@ fix(case self => (
 ))
     '''
 
-    code = '''
-let f = (case ~cons x => x case ~hans x => x) in
-fix(case self => (
-    case ~nil @ => ~zero @ 
-    case y => 
-        ~succ (self(f(y)))
-))
-    '''
+    # TODO: update interpretation to get transitive interpretations 
+
+#     code = '''
+# let f = (case ~cons x => x) in
+# fix(case self => (
+#     case ~nil @ => ~zero @ 
+#     case y => 
+#         ~succ (self(f(y)))
+# ))
+#     '''
 
 #     code = '''
 # let f = (case a => a) in

### 2025-02-17 src/test_analysis.py
update notions of foreign vs local variables along with some notation

@@ -2117,9 +2117,9 @@ if __name__ == '__main__':
     # test_lted_wrapper()
     # test_max_parts_disjoint()
     ######## TODO: update type construction to replace skolems and variables in payload and in relational constraints. 
-    test_max()
+    # test_max()
 
-    # test_fix()
+    test_fix()
     # test_max_annotated()
     # test_max_subtyping()
     # test_max_subtyping_fail()

### 2025-02-19 src/tapas/slim/analyzer.py
update rules

@@ -2392,8 +2392,8 @@ class Solver:
             #     self.extract_factored_upper_bounds(world, lower.id)
             # )
             # return self.solve(world, make_inter(list(bounds)), upper)
-            new_constraints, lower_interp = self.prune_interpret_negative_id(world.closedids, world.constraints, lower.id)
-            if self.solve(World(new_constraints, world.closedids, world.relids), lower_interp, upper):
+            ignore_constraints, lower_interp = self.prune_interpret_negative_id(world.closedids, world.constraints, lower.id)
+            if bool(self.solve(world, lower_interp, upper)):
                 return [world]
             else:
                 return []
@@ -2454,8 +2454,8 @@ class Solver:
             return worlds
 
         elif isinstance(upper, TVar) and upper.id in world.closedids: 
-            new_constraints, upper_interp = self.prune_interpret_positive_id(world.closedids, world.constraints, upper.id)
-            if self.solve(World(new_constraints, world.closedids, world.relids), upper_interp, upper):
+            ignore_constraints, upper_interp = self.prune_interpret_positive_id(world.closedids, world.constraints, upper.id)
+            if self.solve(world, upper_interp, upper):
                 return [world]
             else:
                 return []

### 2025-02-19 src/tapas/slim/analyzer.py
update how rules are grouped with additional conditions

@@ -2197,30 +2197,29 @@ class Solver:
             return self.solve(world, self.aliasing[lower.id], upper)
 
         #######################################
+        #### Implication Rewriting ############
         #######################################
+        # elif isinstance(upper, Imp) and isinstance(upper.antec, Unio):
+        #     return self.solve(world, lower, Inter(
+        #         Imp(upper.antec.left, upper.consq), 
+        #         Imp(upper.antec.right, upper.consq)
+        #     ))
 
-        elif isinstance(upper, Imp) and isinstance(upper.antec, Unio):
-            return self.solve(world, lower, Inter(
-                Imp(upper.antec.left, upper.consq), 
-                Imp(upper.antec.right, upper.consq)
-            ))
+        # elif isinstance(upper, Imp) and isinstance(upper.consq, Inter):
+        #     return self.solve(world, lower, Inter(
+        #         Imp(upper.antec, upper.consq.left), 
+        #         Imp(upper.antec, upper.consq.right)
+        #     ))
 
-        elif isinstance(upper, Imp) and isinstance(upper.consq, Inter):
-            return self.solve(world, lower, Inter(
-                Imp(upper.antec, upper.consq.left), 
-                Imp(upper.antec, upper.consq.right)
-            ))
-
-        elif isinstance(upper, TField) and isinstance(upper.body, Inter):
-            return [
-                m1
-                for m0 in self.solve(world, lower, TField(upper.label, upper.body.left))
-                for m1 in self.solve(m0, lower, TField(upper.label, upper.body.right))
-            ]
-
-
+        # elif isinstance(upper, TField) and isinstance(upper.body, Inter):
+        #     return [
+        #         m1
+        #         for m0 in self.solve(world, lower, TField(upper.label, upper.body.left))
+        #         for m1 in self.solve(m0, lower, TField(upper.label, upper.body.right))
+        #     ]
 
         #######################################
+        #### Base Preservation ################
         #######################################
 
         elif isinstance(lower, TUnit) and isinstance(upper, TUnit): 
@@ -2247,14 +2246,12 @@ class Solver:
             return worlds
 
         #######################################
-        ####  ####
+        ####  Abstraction Elimination #########
         #######################################
 
         elif isinstance(lower, Bot): 
             return [world] 
 
-        elif isinstance(upper, Top): 
-            return [world] 
 
         elif isinstance(lower, Unio):
             return [
@@ -2285,12 +2282,84 @@ class Solver:
                 for m2 in self.solve(m1, lower_body, upper)
             ]
 
[…]

### 2025-02-21 src/tapas/slim/analyzer.py
update type and qualification interpretation

@@ -1842,39 +1842,38 @@ class Solver:
         elif isinstance(src, TUnit):  
             return constraints, TUnit()
         elif isinstance(src, TTag):  
-            constraints, body = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.body)
-            return constraints, TTag(src.label, body)
+            body_constraints, body = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.body)
+            return body_constraints, TTag(src.label, body)
         elif isinstance(src, TField):  
-            constraints, body = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.body)
-            return constraints, TField(src.label, body)
+            body_constraints, body = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.body)
+            return body_constraints, TField(src.label, body)
         elif isinstance(src, Unio):  
-            constraints, left = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.left)
-            constraints, right = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.right)
-            return constraints, Unio(left, right)
+            left_constraints, left = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.left)
+            right_constraints, right = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.right)
+            return (left_constraints).intersection(right_constraints), Unio(left, right)
         elif isinstance(src, Inter):  
-            constraints, left = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.left)
-            constraints, right = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.right)
-            return constraints, Inter(left, right)
+            left_constraints, left = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.left)
+            right_constraints, right = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.right)
+            return (left_constraints).intersection(right_constraints), Inter(left, right)
         elif isinstance(src, Diff):  
-            constraints, context = self.prune_interpret_polar_typ(ignore, positive, closedids, constraints, src.context)
-            constraints, negation = self.prune_interpret_polar_typ(ignore, not positive, closedids, constraints, src.negation)
-            return constraints, Diff(context, negation)
[…]

### 2025-03-07 src/tapas/slim/analyzer.py
update loop typing rules for simple and relational typing

@@ -2179,6 +2179,17 @@ class Solver:
         if self.count > self._limit:
             return []
 
+        print(f"""
+DEBUG SOLVE:
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+{concretize_typ(lower)}
+<:
+{concretize_typ(upper)}
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+              
+
+        """)
+
         #######################################
         #### Reflection ####
         #######################################
@@ -2239,8 +2250,14 @@ class Solver:
         elif isinstance(lower, Imp) and isinstance(upper, Imp): 
             worlds = [
                 m1
+                # TODO: consider switching order
+                # NOTE: to add relational constraint in consequent before checking antecedent
+
                 for m0 in self.solve(world, upper.antec, lower.antec) 
                 for m1 in self.solve(m0, lower.consq, upper.consq) 
+
+                # for m0 in self.solve(world, lower.consq, upper.consq) 
+                # for m1 in self.solve(m0, upper.antec, lower.antec) 
             ]
             return worlds
 

### 2025-03-07 src/tapas/slim/.antlr/SlimParser.java
add back implication rewriting

@@ -177,6 +177,14 @@ public class SlimParser extends Parser {
 			super(parent, invokingState);
 		}
 		@Override public int getRuleIndex() { return RULE_ids; }
+		@Override
+		public void enterRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).enterIds(this);
+		}
+		@Override
+		public void exitRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).exitIds(this);
+		}
 	}
 
 	public final IdsContext ids() throws RecognitionException {
@@ -243,6 +251,14 @@ public class SlimParser extends Parser {
 			super(parent, invokingState);
 		}
 		@Override public int getRuleIndex() { return RULE_preamble; }
+		@Override
+		public void enterRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).enterPreamble(this);
+		}
+		@Override
+		public void exitRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).exitPreamble(this);
+		}
 	}
 
 	public final PreambleContext preamble() throws RecognitionException {
@@ -322,6 +338,14 @@ public class SlimParser extends Parser {
 			this.contexts = contexts;
 		}
 		@Override public int getRuleIndex() { return RULE_program; }
+		@Override
+		public void enterRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).enterProgram(this);
+		}
+		@Override
+		public void exitRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).exitProgram(this);
+		}
 	}
 
 	public final ProgramContext program(list[Context] contexts) throws RecognitionException {
@@ -392,6 +416,14 @@ public class SlimParser extends Parser {
 			super(parent, invokingState);
 		}
 		@Override public int getRuleIndex() { return RULE_typ_base; }
+		@Override
+		public void enterRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).enterTyp_base(this);
+		}
+		@Override
+		public void exitRule(ParseTreeListener listener) {
+			if ( listener instanceof SlimListener ) ((SlimListener)listener).exitTyp_base(this);
+		}
 	}
 
[…]

### 2025-03-07 src/tapas/slim/analyzer.py

@@ -2209,24 +2209,24 @@ DEBUG SOLVE:
         #######################################
         #### Implication Rewriting ############
         #######################################
-        # elif isinstance(upper, Imp) and isinstance(upper.antec, Unio):
-        #     return self.solve(world, lower, Inter(
-        #         Imp(upper.antec.left, upper.consq), 
-        #         Imp(upper.antec.right, upper.consq)
-        #     ))
+        elif isinstance(upper, Imp) and isinstance(upper.antec, Unio):
+            return self.solve(world, lower, Inter(
+                Imp(upper.antec.left, upper.consq), 
+                Imp(upper.antec.right, upper.consq)
+            ))
 
-        # elif isinstance(upper, Imp) and isinstance(upper.consq, Inter):
-        #     return self.solve(world, lower, Inter(
-        #         Imp(upper.antec, upper.consq.left), 
-        #         Imp(upper.antec, upper.consq.right)
-        #     ))
+        elif isinstance(upper, Imp) and isinstance(upper.consq, Inter):
+            return self.solve(world, lower, Inter(
+                Imp(upper.antec, upper.consq.left), 
+                Imp(upper.antec, upper.consq.right)
+            ))
 
-        # elif isinstance(upper, TField) and isinstance(upper.body, Inter):
-        #     return [
-        #         m1
-        #         for m0 in self.solve(world, lower, TField(upper.label, upper.body.left))
-        #         for m1 in self.solve(m0, lower, TField(upper.label, upper.body.right))
-        #     ]
+        elif isinstance(upper, TField) and isinstance(upper.body, Inter):
+            return [
+                m1
+                for m0 in self.solve(world, lower, TField(upper.label, upper.body.left))
+                for m1 in self.solve(m0, lower, TField(upper.label, upper.body.right))
+            ]
 
         #######################################
         #### Base Preservation ################

### 2025-03-10 src/tapas/slim/analyzer.py
update subtyping to better handle relational types

@@ -1027,6 +1027,14 @@ def find_path(assumed_key : Typ, search_target : Typ) -> Optional[tuple[str, ...
             return k
     return None
 
+def find_paths(assumed_key : Typ, search_target : Typ) -> PSet[tuple[str, ...]]:
+    results : PSet[tuple[str, ...]] = s()
+    ordered_path_target_pairs = extract_ordered_path_target_pairs(assumed_key)
+    for (k,v) in ordered_path_target_pairs:
+        if v == search_target:
+            results = results.add(k)
+    return results 
+
 def factorize_choice(induc_id : str, choice : Typ, path : tuple[str, ...]) -> Typ:
     if isinstance(choice, Exi):
 
@@ -1064,6 +1072,16 @@ def find_factors(world : World, search_target : Typ) -> PSet[Typ]:
             if isinstance(constraint.upper, Fixpoint):
                 result = factorize_least_fp(constraint.upper, path)
                 results = results.add(result)
+    return results
+
+def find_factors_from_constraint(constraint : Subtyping, search_target : Typ) -> PSet[Typ]:
+    results = s()
+    paths = find_paths(constraint.lower, search_target)
+    # paths = [find_path(constraint.lower, search_target)]
+    for path in paths:
+        if isinstance(constraint.upper, Fixpoint) and path != None:
+            result = factorize_least_fp(constraint.upper, path)
+            results = results.add(result)
     return results
 
 
@@ -2140,6 +2158,8 @@ class Solver:
     def is_pattern_typ(self, t : Typ) -> bool:
         if isinstance(t, TVar):
             return True
+        if isinstance(t, TUnit):
+            return True
         elif isinstance(t, TTag):
             return self.is_pattern_typ(t.body)
         elif isinstance(t, TField):
@@ -2179,16 +2199,14 @@ class Solver:
         if self.count > self._limit:
             return []
 
-        print(f"""
-DEBUG SOLVE:
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-{concretize_typ(lower)}
-<:
-{concretize_typ(upper)}
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-              
-
-        """)
+#         print(f"""
+# DEBUG SOLVE:
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# {concretize_typ(lower)}
+# <:
+# {concretize_typ(upper)}
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
[…]

### 2025-03-10 src/tapas/slim/analyzer.py
update closed variable safety checks

@@ -2435,8 +2435,25 @@ class Solver:
 
         # elif isinstance(lower, TVar) and lower.id in world.closedids and (not isinstance(upper, TVar) or upper.id in world.closedids): 
         elif isinstance(lower, TVar) and lower.id in world.closedids and (not isinstance(upper, TVar)): 
-            ignore_constraints, lower_interp = self.prune_interpret_negative_id(world.closedids, world.constraints, lower.id)
-            if bool(self.solve(world, lower_interp, upper)):
+            # ignore_constraints, lower_interp = self.prune_interpret_negative_id(world.closedids, world.constraints, lower.id)
+            # if bool(self.solve(world, lower_interp, upper)):
+            #     return [world]
+            # else:
+            #     return []
+
+            closed_parts = pset(
+                st.upper
+                for st in world.constraints
+                if st.lower == lower 
+            ).union(self.extract_factored_upper_bounds(world, lower.id))
+
+            one_part_consistent = any(
+                bool(self.solve(world, closed_part, upper))
+                for closed_part in closed_parts
+                if not isinstance(closed_part, TVar) or closed_part.id in world.closedids 
+            )
+
+            if one_part_consistent: 
                 return [world]
             else:
                 return []
@@ -2450,8 +2467,25 @@ class Solver:
 
         # elif isinstance(upper, TVar) and upper.id in world.closedids and (not isinstance(lower, TVar) or lower.id in world.closedids): 
         elif isinstance(upper, TVar) and upper.id in world.closedids and (not isinstance(lower, TVar)): 
-            ignore_constraints, upper_interp = self.prune_interpret_positive_id(world.closedids, world.constraints, upper.id)
-            if self.solve(world, upper_interp, upper):
+            # ignore_constraints, upper_interp = self.prune_interpret_positive_id(world.closedids, world.constraints, upper.id)
+            # if self.solve(world, upper_interp, upper):
+            #     return [world]
+            # else:
+            #     return []
+
+            closed_parts = pset(
+                st.lower
+                for st in world.constraints
+                if st.upper == upper 
+            )
+
+            one_part_consistent = any(
+                bool(self.solve(world, upper, closed_part))
+                for closed_part in closed_parts
+                if not isinstance(closed_part, TVar) or closed_part.id in world.closedids 
+            )
+
+            if one_part_consistent: 
                 return [world]
             else:
                 return []

### 2025-03-10 src/test_analysis.py

@@ -2239,7 +2239,7 @@ if __name__ == '__main__':
     # test_lted_wrapper()
     # test_max_parts_disjoint()
     ######## TODO: update type construction to replace skolems and variables in payload and in relational constraints. 
-    # test_max()
+    test_max()
 
     # test_fix()
     # test_max_annotated()
@@ -2290,7 +2290,7 @@ if __name__ == '__main__':
     # test_pair_subs_relational_constraint_false()
     # test_top_subs_relational_constraint_false()
     #####################################
-    test_relational_implication_subtyping_pass_A()
+    # test_relational_implication_subtyping_pass_A()
     # test_relational_implication_subtyping_pass_B()
     # test_relational_implication_subtyping_fail()
     #####################################

### 2025-03-22 src/tapas/slim/analyzer.py
update order of subtyping rules

@@ -2247,6 +2247,16 @@ class Solver:
             return [world] 
 
         #######################################
+        #### Reflection ####
+        #######################################
+
+        elif isinstance(lower, Bot): 
+            return [world] 
+
+        elif isinstance(upper, Top): 
+            return [world] 
+
+        #######################################
         #### Dealiasing ####
         #######################################
         elif isinstance(upper, TVar) and upper.id in self.aliasing: 
@@ -2324,9 +2334,6 @@ class Solver:
         ####  Abstraction Elimination #########
         #######################################
 
-        elif isinstance(lower, Bot): 
-            return [world] 
-
 
         elif isinstance(lower, Unio):
             return [
@@ -2388,9 +2395,6 @@ class Solver:
         #######################################
         #### Refinement Introduction ##########
         #######################################
-
-        elif isinstance(upper, Top): 
-            return [world] 
 
         elif isinstance(upper, Inter):
             return [
@@ -2510,54 +2514,41 @@ class Solver:
             else:
                 return []
 
+        #######################################
+        #### Diff Introduction #############
+        #######################################
+
+        elif isinstance(upper, Diff): 
+
+            # if diff_well_formed(upper): # TODO: change to: DF(lower) and DF(upper)
+            if True:
+                # TODO: need a sound/safe/conservative inhabitable check
+                # only works if we assume T is not empty
+                '''
+                T <: A \\ B === (T <: A) and (T is inhabitable --> ~(T <: B))
+                ----
+                T <: A \\ B === (T <: A) and ((T <: B) --> T is empty)
+                ----
+                T <: A \\ B === (T <: A) and (~(T <: B) or T is empty)
+                '''
+                context_worlds = self.solve(world, lower, upper.context)
+                return [
+                    m
+                    for m in context_worlds 
+                    # if (
+                    #     # not self.is_inhabitable(world, lower) or 
+                    #     # Fail (incompletely) if lower is empty
+                    #     self.solve(m, lower, upper.negation) == []
+                    # )
+                ]   
+            else:
+                return []
 
 
         #######################################
-        #### Refinement Elimination ###########
+        #### Fixpoint Elimination #############
         #######################################
 
-
-        elif isinstance(lower, Inter) and self.is_base_typ(upper) : 
-            return self.solve(world, lower.left, upper) + self.solve(world, lower.right, upper)
-
[…]

### 2025-03-22 src/tapas/slim/.antlr/SlimParser.java
update such that records subsume all tagged data constructors

@@ -456,7 +456,7 @@ public class SlimParser extends Parser {
 				setState(85);
 				((Typ_baseContext)_localctx).typ_base = typ_base();
 
-				_localctx.combo = TTag((((Typ_baseContext)_localctx).ID!=null?((Typ_baseContext)_localctx).ID.getText():null), ((Typ_baseContext)_localctx).typ_base.combo) 
+				_localctx.combo = TEntry((((Typ_baseContext)_localctx).ID!=null?((Typ_baseContext)_localctx).ID.getText():null), ((Typ_baseContext)_localctx).typ_base.combo) 
 
 				}
 				break;
@@ -470,7 +470,7 @@ public class SlimParser extends Parser {
 				setState(90);
 				((Typ_baseContext)_localctx).typ_base = typ_base();
 
-				_localctx.combo = TField((((Typ_baseContext)_localctx).ID!=null?((Typ_baseContext)_localctx).ID.getText():null), ((Typ_baseContext)_localctx).typ_base.combo) 
+				_localctx.combo = TEntry((((Typ_baseContext)_localctx).ID!=null?((Typ_baseContext)_localctx).ID.getText():null), ((Typ_baseContext)_localctx).typ_base.combo) 
 
 				}
 				break;
@@ -619,7 +619,7 @@ public class SlimParser extends Parser {
 				setState(125);
 				((TypContext)_localctx).typ = typ();
 
-				_localctx.combo = Inter(TField('head', ((TypContext)_localctx).typ_base.combo), TField('tail', ((TypContext)_localctx).typ.combo)) 
+				_localctx.combo = Inter(TEntry('head', ((TypContext)_localctx).typ_base.combo), TEntry('tail', ((TypContext)_localctx).typ.combo)) 
 
 				}
 				break;

### 2025-03-22 src/tapas/slim/analyzer.py

@@ -199,13 +199,13 @@ class TVar:
 class TUnit:
     pass
 
-@dataclass(frozen=True, eq=True)
-class TTag:
-    label : str
-    body : Typ 
+# @dataclass(frozen=True, eq=True)
+# class TTag:
+#     label : str
+#     body : Typ 
 
 @dataclass(frozen=True, eq=True)
-class TField:
+class TEntry:
     label : str
     body : Typ 
 
@@ -254,7 +254,8 @@ class Top:
 class Bot:
     pass
 
-Typ = Union[TVar, TUnit, TTag, TField, Unio, Inter, Diff, Imp, Exi, All, Fixpoint, Top, Bot]
+# Typ = Union[TVar, TUnit, TTag, TField, Unio, Inter, Diff, Imp, Exi, All, Fixpoint, Top, Bot]
+Typ = Union[TVar, TUnit, TEntry, Unio, Inter, Diff, Imp, Exi, All, Fixpoint, Top, Bot]
 
 
 @dataclass(frozen=True, eq=True)
@@ -298,10 +299,10 @@ Nameless Type
 class BVar:
     id : int 
 
-@dataclass(frozen=True, eq=True)
-class TTagNL:
-    label : str
-    body : NL 
+# @dataclass(frozen=True, eq=True)
+# class TTagNL:
+#     label : str
+#     body : NL 
 
 @dataclass(frozen=True, eq=True)
 class TFieldNL:
@@ -349,7 +350,8 @@ class SubtypingNL:
     strong : NL 
     weak : NL 
 
-NL = Union[TVar, BVar, TUnit, TTagNL, TFieldNL, UnioNL, InterNL, DiffNL, ImpNL, ExiNL, AllNL, LeastFPNL, Top, Bot]
+# NL = Union[TVar, BVar, TUnit, TTagNL, TFieldNL, UnioNL, InterNL, DiffNL, ImpNL, ExiNL, AllNL, LeastFPNL, Top, Bot]
+NL = Union[TVar, BVar, TUnit, TFieldNL, UnioNL, InterNL, DiffNL, ImpNL, ExiNL, AllNL, LeastFPNL, Top, Bot]
 
 def to_nameless(bound_ids : tuple[str, ...], typ : Typ) -> NL:
     assert isinstance(bound_ids, tuple)
@@ -363,9 +365,9 @@ def to_nameless(bound_ids : tuple[str, ...], typ : Typ) -> NL:
             return typ
[…]

### 2025-03-25 src/tapas/slim/analyzer.py
update variable rules

@@ -2498,7 +2498,8 @@ class Solver:
             return worlds 
 
         # elif isinstance(lower, TVar) and lower.id in world.closedids and (not isinstance(upper, TVar) or upper.id in world.closedids): 
-        elif isinstance(lower, TVar) and lower.id in world.closedids and (not isinstance(upper, TVar)): 
+        # elif isinstance(lower, TVar) and lower.id in world.closedids and (not isinstance(upper, TVar)): 
+        elif isinstance(lower, TVar) and lower.id in world.closedids: 
             # ignore_constraints, lower_interp = self.prune_interpret_negative_id(world.closedids, world.constraints, lower.id)
             # if bool(self.solve(world, lower_interp, upper)):
             #     return [world]

### 2025-03-28 src/test_analysis.py
update variable rules and descriptions

@@ -2300,11 +2300,11 @@ if __name__ == '__main__':
     # test_two_cons_query_subs_nat_list()
     # test_plus_equals_two_query()
     #####################################
-    # test_even_list_subs_nat_list_pass()
-    # test_nat_list_subs_even_list_fail()
+    test_even_list_subs_nat_list_pass()
+    test_nat_list_subs_even_list_fail()
     # #####################################
-    # test_even_subs_nat_pass()
-    # test_nat_subs_even_fail()
+    test_even_subs_nat_pass()
+    test_nat_subs_even_fail()
     #####################################
     # test_one_plus_equals_two_query()
     #####################################
@@ -2313,7 +2313,7 @@ if __name__ == '__main__':
     # test_lted_wrapper()
     # test_max_parts_disjoint()
     ######## TODO: update type construction to replace skolems and variables in payload and in relational constraints. 
-    # test_max()
+    test_max()
 
     # test_fix()
     # test_max_annotated()

### 2025-03-28 src/tapas/slim/analyzer.py
update closed variable safety code

@@ -2211,11 +2211,14 @@ class Solver:
             if not isinstance(st.upper, TVar) or st.upper.id not in world.closedids
         )
 
-        new_world = replace(world, constraints = world.constraints.union(closed_constraints))
-        return [
-            replace(w0, constraints = w0.constraints.add(Subtyping(lower, upper))) 
-            for w0 in self.solve_multi(new_world, unsolved_constraints)
-        ]
+        if self.closed_constraints_safe(world, closed_constraints):
+            new_world = replace(world, constraints = world.constraints.union(closed_constraints))
+            return [
+                replace(w0, constraints = w0.constraints.add(Subtyping(lower, upper))) 
+                for w0 in self.solve_multi(new_world, unsolved_constraints)
+            ]
+        else:
+            return []
 
     def is_pattern_typ(self, t : Typ) -> bool:
         if isinstance(t, TVar):
@@ -2260,6 +2263,65 @@ class Solver:
             isinstance(t, Unio) or
             isinstance(t, Fixpoint) or
             isinstance(t, Exi)
+        )
+
+
+    def closed_variable_elimination_safe(self, world : World, lower : TVar, upper : Typ) -> bool:
+        upper_parts = pset(
+            st.upper
+            for st in world.constraints
+            if st.lower == lower 
+        ).union(self.extract_factored_upper_bounds(world, lower.id))
+
+        some_parts_consistent = any(
+            (
+                (isinstance(upper_part, TVar) and upper_part.id not in world.closedids) 
+                or bool(self.solve(world, upper_part, upper))
+            )
+            for upper_part in upper_parts
+            # if not isinstance(upper_part, TVar) or upper_part.id in world.closedids  else
+        )
+
+        all_parts_consistent = all(
+            bool(self.solve(world, st.lower, upper))
+            for st in world.constraints
+            if st.upper == lower 
+        )
+
+        return some_parts_consistent and all_parts_consistent
+
+    def closed_variable_introduction_safe(self, world : World, lower : Typ, upper : TVar) -> bool:
+        some_parts_consistent = any(
+            (
+                (isinstance(st.lower, TVar) and st.lower.id not in world.closedids)
+                or bool(self.solve(world, lower, st.lower))
+            )
+            for st in world.constraints
+            if st.upper == upper 
+        )
+
+        all_parts_consistent = all(
+            bool(self.solve(world, lower, st.upper))
+            for st in world.constraints
+            if st.lower == upper 
[…]

### 2025-05-04 src/tapas/slim/analyzer.py
update a little

@@ -3072,6 +3072,15 @@ class ExprRule(Rule):
 
 
     def combine_application(self, pid : int, world : World, cator_typ : Typ, arg_typs : list[Typ]) -> List[Result]: 
+#         print(f"""
+# ~~~~~~~~~~~~~~~~~~~~~~~
+# DEBUG combine_application ASSUME
+# ~~~~~~~~~~~~~~~~~~~~~~~
+# world.closedids: {world.closedids}
+# cator_typ: {concretize_typ(cator_typ)}
+# arg_typs: {[concretize_typ(arg_typ) for arg_typ in arg_typs]}
+# ~~~~~~~~~~~~~~~~~~~~~~~
+#         """)
         worlds = [world] 
         current_cator_typ = cator_typ
         for arg_typ in arg_typs:
@@ -3081,6 +3090,27 @@ class ExprRule(Rule):
                 new_worlds.extend(self.solver.solve(world, current_cator_typ, Imp(arg_typ, result_var)))
             worlds = new_worlds
             current_cator_typ = result_var
+
+#         print(f"""
+# ~~~~~~~~~~~~~~~~~~~~~~~
+# DEBUG combine_application GUARANTEE 
+# ~~~~~~~~~~~~~~~~~~~~~~~
+# cator_typ: {concretize_typ(cator_typ)}
+# arg_typs: {[concretize_typ(arg_typ) for arg_typ in arg_typs]}
+
+# result: {result_var}
+# ~~~~~~~~~~~~~~~~~~~~~~~
+#         """)
+#         for i,world in enumerate(worlds):
+#             print(f"""
+# ~~~~~~~~~~~~~~~~~~~~~~~
+# DEBUG combine_application - World {i} 
+# ~~~~~~~~~~~~~~~~~~~~~~~
+# world.closedids: {world.closedids}
+# world.constraints: 
+# {concretize_constraints(world.constraints)}
+# ~~~~~~~~~~~~~~~~~~~~~~~
+#             """)
 
         return [
             Result(pid, world, result_var)

### 2025-05-21 src/tapas/slim/analyzer.py
update lemmas and definition for fixed point combinator loop case

@@ -1110,6 +1110,7 @@ def factorize_least_fp(t : Fixpoint, path : tuple[str, ...]) -> Fixpoint:
     return Fixpoint(t.id, factorized_body)
 
 
+# TODO: replace with find_paths plural
 def find_factors(world : World, search_target : Typ) -> PSet[Typ]:
     results = s()
     for constraint in world.constraints:

### 2025-05-27 src/test_analysis.py
update duality and fixed point lemmas

@@ -2300,11 +2300,11 @@ if __name__ == '__main__':
     # test_two_cons_query_subs_nat_list()
     # test_plus_equals_two_query()
     #####################################
-    test_even_list_subs_nat_list_pass()
-    test_nat_list_subs_even_list_fail()
+    # test_even_list_subs_nat_list_pass()
+    # test_nat_list_subs_even_list_fail()
     # #####################################
-    test_even_subs_nat_pass()
-    test_nat_subs_even_fail()
+    # test_even_subs_nat_pass()
+    # test_nat_subs_even_fail()
     #####################################
     # test_one_plus_equals_two_query()
     #####################################

### 2025-05-28 src/tapas/slim/analyzer.py
update implementation by removing pruning and factoring influential filter out of making constraint type

@@ -1265,15 +1265,26 @@ end extract_free_vars_from_typ
 '''
 
 def extract_free_vars_from_constraints(bound_vars : PSet[str], constraints : Iterable[Subtyping]) -> PSet[str]:
-    result = s()
-    for st in constraints:
-        result = (
-            result
-            .union(extract_free_vars_from_typ(bound_vars, st.lower))
-            .union(extract_free_vars_from_typ(bound_vars, st.upper))
-        )
+    return pset(
+        id
+        for st in constraints
+        for id in extract_free_vars_from_typ(bound_vars, st.lower).union(extract_free_vars_from_typ(bound_vars, st.upper))
+    )
 
-    return result
+def extract_free_vars_from_enviro(enviro : PMap[str, Typ]) -> PSet[str]:
+    return pset( 
+        id
+        for t in enviro.values()
+        for id in extract_free_vars_from_typ(s(), t)
+    )
+
+def filter_constraints_by_all_variables(constraints : Iterable[Subtyping], influential_ids) -> PSet[Subtyping]:
+    return pset( 
+        st
+        for st in constraints 
+        # if all free type variabls in st constraint are influential
+        if not bool(extract_free_vars_from_constraints(s(), [st]).difference(influential_ids))
+    )
 
 def is_variable_unassigned(world : World, id : str) -> bool:
     return (
@@ -1693,8 +1704,8 @@ class Solver:
         big_typ = base
         for result in results:
 
-            new_constraints, t = self.prune_interpret_polar_typ(polarity, result.world.closedids, result.world.constraints, result.typ)
-            ctyp = self.make_constraint_typ(polarity)(s(), result.world.closedids, new_constraints, t)
+            t = self.interpret_polar_typ(polarity, result.world.closedids, result.world.constraints, result.typ)
+            ctyp = self.make_constraint_typ(polarity)(s(), result.world.closedids, result.world.constraints, t)
             big_typ = operator(big_typ, ctyp)
         return big_typ
 
@@ -1785,15 +1796,6 @@ class Solver:
         (outer_con, inner_con) = ((Exi, All) if positive else (All, Exi))
 
         def make(foreignids : PSet[str], closedids : PSet[str], constraints : PSet[Subtyping], payload : Typ):
-
-            influential_ids = foreignids.union(closedids).union(extract_free_vars_from_typ(s(), payload))
-            constraints = pset( 
-                st
-                for st in constraints 
-                # if all free type variabls in st constraint are influential
[…]

### 2025-05-31 src/tapas/slim/analyzer.py
add outline for proof subtyping soundness

@@ -2437,6 +2437,8 @@ class Solver:
                     for m1 in self.solve(m0, constraint.lower, constraint.upper)
                 ]  
 
+            # TODO: modify so worlds has exactly one world?
+
             return [
                 m2
                 for m0 in worlds

### 2025-06-09 src/tapas/slim/analyzer.py
simplify type variable rules

@@ -2165,33 +2165,11 @@ class Solver:
 
 
     def solve_open_variable_introduction(self, world : World, lower : Typ, upper : TVar) -> list[World]:
-        new_constraints = self.sub_polar_constraints(False, world.constraints, upper.id, lower).difference(world.constraints)
-
-        closed_constraints = pset(
-            st
-            for st in new_constraints 
-            if (
-                False
-                or isinstance(st.lower, TVar) and st.lower.id in world.closedids
-                or isinstance(st.upper, TVar) and st.upper.id in world.closedids
-            )
-        )
-
-        unsolved_constraints = pset(
-            st
-            for st in new_constraints 
-            if not isinstance(st.lower, TVar) or st.lower.id not in world.closedids
-            if not isinstance(st.upper, TVar) or st.upper.id not in world.closedids
-        )
-
-        if self.closed_constraints_safe(world, closed_constraints):
-            new_world = replace(world, constraints = world.constraints.union(closed_constraints))
-            return [
-                replace(w0, constraints = w0.constraints.add(Subtyping(lower, upper))) 
-                for w0 in self.solve_multi(new_world, unsolved_constraints)
-            ]
-        else:
-            return []
+        subbed_constraints = self.sub_polar_constraints(False, world.constraints, upper.id, lower).difference(world.constraints)
+        return [
+            replace(w0, constraints = w0.constraints.add(Subtyping(lower, upper))) 
+            for w0 in self.solve_multi(world, subbed_constraints)
+        ]
 
     def is_pattern_typ(self, t : Typ) -> bool:
         if isinstance(t, TVar):
@@ -2513,52 +2491,26 @@ class Solver:
         #######################################
 
         elif isinstance(lower, TVar) and lower.id not in world.closedids:
-            new_constraints = self.sub_polar_constraints(True, world.constraints, lower.id, upper).difference(world.constraints)
+            subbed_constraints = self.sub_polar_constraints(True, world.constraints, lower.id, upper).difference(world.constraints)
 
-            closed_constraints = pset(
-                st
-                for st in new_constraints 
-                if (
-                    False
-                    or (isinstance(st.lower, TVar) and st.lower.id in world.closedids)
-                    or (isinstance(st.upper, TVar) and st.upper.id in world.closedids)
-                )
-            )
+            worlds = [
+                replace(w0, constraints = w0.constraints.add(Subtyping(lower, upper))) 
+                for w0 in self.solve_multi(world, subbed_constraints)
+            ]
 
[…]

### 2025-06-19 src/tapas/slim/exprlib.py
add packing of type when checking against annotation

@@ -11,7 +11,7 @@ from tapas.slim.typlib import *
 
 def ctx(names : list[str], code : str):
     base = code
-    for name in reversed(names):
+    for name in (names):
         base = (f"""
 let zzz_{name} : ({context_map.get(name)}) -> TOP = 
 {{ {name} => {base} }} 
@@ -39,7 +39,7 @@ context_map = {
 "map" : f"ALL[A B] (A -> B) -> {List_('A')} -> {List_('B')}",
 "app" : f"ALL[A B] (A -> B) -> A -> B",
 "revapp" : f"ALL[A B] A -> (A -> B) -> B",
-"runState" : f"ALL[A] (ALL[S] {State('S', 'A')}) -> B",
+"runState" : f"ALL[A B] (ALL[S] {State('S', 'A')}) -> B",
 "argState" : f"ALL[S] {State('S', Nat)}",
 "zero" : Church,
 "succ" : f"{Church} -> {Church}",

### 2025-06-26 src/tapas/slim/SlimLexer.py
update indexed type syntax; update skolem variable rules

@@ -22,73 +22,72 @@ from pyrsistent import m, s, pmap, pset
 
 def serializedATN():
     return [
-        4,0,38,197,6,-1,2,0,7,0,2,1,7,1,2,2,7,2,2,3,7,3,2,4,7,4,2,5,7,5,
+        4,0,37,192,6,-1,2,0,7,0,2,1,7,1,2,2,7,2,2,3,7,3,2,4,7,4,2,5,7,5,
         2,6,7,6,2,7,7,7,2,8,7,8,2,9,7,9,2,10,7,10,2,11,7,11,2,12,7,12,2,
         13,7,13,2,14,7,14,2,15,7,15,2,16,7,16,2,17,7,17,2,18,7,18,2,19,7,
         19,2,20,7,20,2,21,7,21,2,22,7,22,2,23,7,23,2,24,7,24,2,25,7,25,2,
         26,7,26,2,27,7,27,2,28,7,28,2,29,7,29,2,30,7,30,2,31,7,31,2,32,7,
-        32,2,33,7,33,2,34,7,34,2,35,7,35,2,36,7,36,2,37,7,37,1,0,1,0,1,0,
-        1,0,1,0,1,0,1,1,1,1,1,2,1,2,1,2,1,2,1,3,1,3,1,3,1,3,1,4,1,4,1,5,
-        1,5,1,6,1,6,1,7,1,7,1,8,1,8,1,9,1,9,1,10,1,10,1,11,1,11,1,11,1,12,
-        1,12,1,13,1,13,1,13,1,13,1,14,1,14,1,15,1,15,1,16,1,16,1,16,1,16,
-        1,17,1,17,1,17,1,17,1,18,1,18,1,19,1,19,1,19,1,20,1,20,1,20,1,21,
-        1,21,1,22,1,22,1,22,1,23,1,23,1,23,1,23,1,23,1,24,1,24,1,24,1,24,
-        1,24,1,25,1,25,1,25,1,25,1,26,1,26,1,26,1,27,1,27,1,27,1,27,1,28,
-        1,28,1,29,1,29,1,30,1,30,1,30,1,31,1,31,1,32,1,32,1,33,1,33,1,33,
-        1,34,1,34,1,35,1,35,5,35,181,8,35,10,35,12,35,184,9,35,1,36,4,36,
-        187,8,36,11,36,12,36,188,1,37,4,37,192,8,37,11,37,12,37,193,1,37,
[…]

### 2025-06-30 src/tapas/slim/.antlr/SlimLexer.java
update structural typing tests

@@ -57,7 +57,7 @@ public class SlimLexer extends Lexer {
 			null, "'alias'", "'='", "'TOP'", "'BOT'", "'@'", "'<'", "'>'", "'('", 
 			"')'", "'|'", "'&'", "'->'", "'*'", "'EXI'", "'['", "']'", "'ALL'", "'LFP'", 
 			"'\\'", "':'", "'<:'", "','", "'if'", "'then'", "'else'", "'let'", "'in'", 
-			"'fix'", "';'", "'{'", "'=>'", "'}'", "'.'", "'|>'"
+			"'loop'", "';'", "'{'", "'=>'", "'}'", "'.'", "'|>'"
 		};
 	}
 	private static final String[] _LITERAL_NAMES = makeLiteralNames();
@@ -128,7 +128,7 @@ public class SlimLexer extends Lexer {
 	public ATN getATN() { return _ATN; }
 
 	public static final String _serializedATN =
-		"\u0004\u0000%\u00c0\u0006\uffff\uffff\u0002\u0000\u0007\u0000\u0002\u0001"+
+		"\u0004\u0000%\u00c1\u0006\uffff\uffff\u0002\u0000\u0007\u0000\u0002\u0001"+
 		"\u0007\u0001\u0002\u0002\u0007\u0002\u0002\u0003\u0007\u0003\u0002\u0004"+
 		"\u0007\u0004\u0002\u0005\u0007\u0005\u0002\u0006\u0007\u0006\u0002\u0007"+
 		"\u0007\u0007\u0002\b\u0007\b\u0002\t\u0007\t\u0002\n\u0007\n\u0002\u000b"+
@@ -153,16 +153,16 @@ public class SlimLexer extends Lexer {
 		"\u0017\u0001\u0017\u0001\u0017\u0001\u0018\u0001\u0018\u0001\u0018\u0001"+
 		"\u0018\u0001\u0018\u0001\u0019\u0001\u0019\u0001\u0019\u0001\u0019\u0001"+
 		"\u001a\u0001\u001a\u0001\u001a\u0001\u001b\u0001\u001b\u0001\u001b\u0001"+
[…]

### 2025-06-30 src/tapas/slim/.antlr/SlimParser.java

@@ -53,7 +53,7 @@ public class SlimParser extends Parser {
 			null, "'alias'", "'='", "'TOP'", "'BOT'", "'@'", "'<'", "'>'", "'('", 
 			"')'", "'|'", "'&'", "'->'", "'*'", "'EXI'", "'['", "']'", "'ALL'", "'LFP'", 
 			"'\\'", "':'", "'<:'", "','", "'if'", "'then'", "'else'", "'let'", "'in'", 
-			"'fix'", "';'", "'{'", "'=>'", "'}'", "'.'", "'|>'"
+			"'loop'", "';'", "'{'", "'=>'", "'}'", "'.'", "'|>'"
 		};
 	}
 	private static final String[] _LITERAL_NAMES = makeLiteralNames();

### 2025-06-30 src/tapas/slim/exprlib.py

@@ -49,6 +49,9 @@ context_map = {
 "h" : f"{Nat} -> ALL[A] A -> A",
 "l" : List_(f"ALL[A] {Nat} -> A -> A"),
 "r" : f"(ALL[A] A -> ALL[B] B -> B) -> {Nat}",
+"sort" : f"(ALL[A] (A * A -> {Bool}) -> {List_('A')} -> {List_('A')})",
+"scalarCmp" : f"{Nat} * {Nat} -> {Bool}",
+"lexicoCmp" : f"{List_(Nat)} * {List_(Nat)} -> {Bool}",
 }
 
 
@@ -98,7 +101,54 @@ single = (f"""
     case x => (cons;(x, <nil> @)) 
 """.strip())
 
-append = (f"""
+double = (f"""
+fix({{ self => 
+    {{ zero;@ => zero;@ }}
+    {{ succ;n => succ;succ;(self(n)) }}
+}})
+""".strip())
+
+halve = (f"""
+fix({{ self => 
+    {{ zero;@ => zero;@ }}
+    {{ succ;succ;n => succ;(self(n)) }}
+}})
+""".strip())
+
+
+both = (f"""
+(
+{{ (true;@),(true;@) => true;@  }}
+{{ (false;@),(true;@) => false;@  }}
+{{ (false;@),(false;@) => false;@  }}
+{{ (true;@),(false;@) => false;@  }}
+)
+""".strip())
+
+isNat = (f"""
+fix({{ self => 
+    {{ zero;@ => true;@ }}
+    {{ succ;n => self(n) }}
+    {{ ignore => false;@ }}
+}})
+""".strip())
+
+isNatList = (f"""
+fix({{ self => 
+    {{ nil;@ => true;@ }}
+    {{ cons;(n, xs) => both(({isNat}(n)), self(xs)) }}
+    {{ ignore => false;@ }}
+}})
+""".strip())
+
+stdCmp = (f"""
+({{a,b =>
+    if both(({isNat})(a), ({isNat})(b)) then
+        scalarCmp(a,b)
+    else (both(({isNatList})(a), ({isNatList})(b))) |> ({{true;@ =>
+        lexicoCmp(a,b)
+    }})
+}})
 """.strip())
 
 ############################################################

### 2025-06-30 src/tapas/slim/language.py

@@ -169,6 +169,9 @@ def analyze(code : str) -> tuple[Optional[analyzer.Typ], str, analyzer.Solver]:
             result = tc.results[0]
             return (result.typ, tc.toStringTree(recog=parser), parser._solver)
         else: 
+            print("!!!!!!!!!!!!!!!")
+            print("result size: " + str(len(tc.results)))
+            print("!!!!!!!!!!!!!!!")
             return (None, tc.toStringTree(recog=parser), parser._solver)
     # except Exception as e:
     #     raise e

### 2025-06-30 src/tapas/slim/typlib.py

@@ -16,17 +16,24 @@ Bool = (f"""
 """.strip())
 
 Nat = (f"""
-(LFP[Self] BOT
+(LFP[R] BOT
 | (<zero> @) 
-| (<succ> Self)
+| (<succ> R)
+)
+""".strip())
+
+Even = (f"""
+(LFP[R] BOT
+| (<zero> @) 
+| (<succ> <succ> R)
 )
 """.strip())
 
 def List_(T): 
     return (f"""
-(LFP[Self] BOT
+(LFP[R] BOT
 | (<nil> @) 
-| (<cons> (({T}) * Self))
+| (<cons> (({T}) * R))
 )
     """.strip())
 

### 2025-06-30 src/test_analysis_new.py

@@ -451,12 +451,12 @@ c
 def test_typing_G9():
     #TODO: fail
     code = (f"""
-fix({{ loop =>
+loop({{ self =>
     {{ x =>
         if <true> @ then
             x
         else 
-            loop(loop)(x)
+            self(self)(x)
     }}
 }})
     """)
@@ -555,6 +555,58 @@ let foo = {{f =>
 #     assert solve_subtyping(lower, upper)
 
 ###############################################################
+##### Structural Typing 
+###############################################################
+def test_typing_structures_1():
+    code = (f"""
+(
+{{ nil;@ => false;@}}
+{{ zero;@ => false;@}}
+{{ succ;n => true;@}}
+{{ cons;(x,xs) => true;@}}
+)
+    """)
+    print(code)
+    assert infer_typ(code)
+
+# def test_typing_structures_debug():
+
+#     lower = (f"""
+#     """)
+
+#     upper = (f"""
+#     """)
+
+#     assert solve_subtyping(lower, upper)
+
+
+def test_typing_structures_2():
+    # double
+    code = (f"""
+let double : (
+   {tl.Nat} -> {tl.Even} 
+) = loop({{self => (
+    {{ (zero;@) => zero;@ }}
+    {{ (succ;n) => succ;succ;((self)(n)) }}
+)}}) in double)
+    """)
+    print(code)
+    assert infer_typ(code)
+
+def test_typing_structures_3():
+    # halve 
+    code = (f"""
+let halve : (
+    {tl.Even} -> {tl.Nat} 
+) = loop({{self => (
+    {{ (zero;@) => zero;@ }}
+    {{ (succ;succ;n) => succ;((self)(n)) }}
+)}}) in halve)
+    """)
+    print(code)
+    assert infer_typ(code)
+
+###############################################################
 ###############################################################
 
 if __name__ == '__main__':
@@ -562,5 +614,6 @@ if __name__ == '__main__':
     # SCRATCH WORK
     # test_typing_A9()
     # test_max()
+    test_typing_structures_3()
 
 #######################################################################

### 2025-07-01 src/tapas/slim/exprlib.py
update examples and remove some syntactic sugar

@@ -125,29 +125,14 @@ both = (f"""
 )
 """.strip())
 
-isNat = (f"""
-loop({{ self => 
-    {{ zero;@ => true;@ }}
-    {{ succ;n => self(n) }}
-    {{ ignore => false;@ }}
-}})
-""".strip())
-
-isNatList = (f"""
-loop({{ self => 
-    {{ nil;@ => true;@ }}
-    {{ cons;(n, xs) => both(({isNat})(n), (self)(xs)) }}
-    {{ ignore => false;@ }}
-}})
-""".strip())
-
 stdCmp = (f"""
 ({{a,b =>
-    if both(({isNat})(a), ({isNat})(b)) then
-        scalarCmp(a,b)
-    else (both(({isNatList})(a), ({isNatList})(b))) |> ({{true;@ =>
-        lexicoCmp(a,b)
-    }})
+    a |> (
+        {{zero;@ => scalarCmp(a,b)}}
+        {{succ;n => scalarCmp(a,b)}}
+        {{nil;@ => lexicoCmp(a,b)}}
+        {{cons;(x,xs) => lexicoCmp(a,b)}}
+    )
 }})
 """.strip())
 

### 2025-07-02 src/test_analysis_new.py
update related work on extrusion and distributivity

@@ -611,6 +611,7 @@ if __name__ == '__main__':
     # SCRATCH WORK
     # test_typing_A9()
     # test_max()
-    test_typing_structures_4()
+    # test_typing_structures_4()
+    test_typing_sanity_inter_nesting_constraint()
 
 #######################################################################

### 2025-07-03 src/tapas/slim/exprlib.py
update experiments to assume context; update failure to include TOP

@@ -4,20 +4,14 @@ from typing import *
 from dataclasses import dataclass
 from tapas.slim.typlib import *
 
+from pyrsistent.typing import PMap, PSet 
+from pyrsistent import m, s, pmap, pset
+
 
 ############################################################
 #### Context ####
 ############################################################
 
-def ctx(names : list[str], code : str):
-    base = code
-    for name in (names):
-        base = (f"""
-let zzz_{name} : ({context_map.get(name)}) -> TOP = 
-{{ {name} => {base} }} 
-in zzz_{name} 
-        """.strip())
-    return base
 
 context_map = {
 "head" : f"ALL[A] {List_('A')} -> A",
@@ -53,6 +47,21 @@ context_map = {
 "scalarCmp" : f"{Nat} * {Nat} -> {Bool}",
 "lexicoCmp" : f"{List_(Nat)} * {List_(Nat)} -> {Bool}",
 }
+
+def ctx(names : list[str]) -> PMap[str, str]:
+    return pmap({
+        n : context_map[n]
+        for n in names
+    })
+# def ctx(names : list[str], code : str):
+#     base = code
+#     for name in (names):
+#         base = (f"""
+# let zzz_{name} : ({context_map.get(name)}) -> TOP = 
+# {{ {name} => {base} }} 
+# in zzz_{name} 
+#         """.strip())
+#     return base
 
 
 ############################################################

### 2025-07-03 src/tapas/slim/language.py

@@ -169,15 +169,12 @@ def analyze(code : str, typing_context = m()) -> tuple[Optional[analyzer.Typ], s
             raise Exception("Parsing Error")
         elif len(tc.results) == 1:
             result = tc.results[0]
-            print(f"""
-~~~~~~~~~~~~~
-constraints:
-~~~~~~~~~~~~~
-{analyzer.concretize_constraints(result.world.constraints)}
-~~~~~~~~~~~~~
-constraints:
-            """)
-            return (result.typ, tc.toStringTree(recog=parser), parser._solver)
+            solver = parser._solver
+            t1 = solver.interpret_polar_typ(True, result.world.closedids, result.world.constraints, result.typ) 
+            influential_ids = result.world.closedids.union(analyzer.extract_free_vars_from_typ(s(), t1))
+            influential_constraints = analyzer.filter_constraints_by_all_variables(result.world.constraints, influential_ids)
+            t2 = solver.make_constraint_typ(True)(s(), result.world.closedids, influential_constraints, t1)
+            return (t2, tc.toStringTree(recog=parser), parser._solver)
         else: 
             print("!!!!!!!!!!!!!!!")
             print("result size: " + str(len(tc.results)))
@@ -208,6 +205,9 @@ def infer_typ(code : str, context = {}) -> str:
     end = time.time()
     print(f"TIME: {end - start}")
     if result == None:
+        return ""
+    if result == analyzer.Top:
+        print("~~~~~~~~~~~TOP!!!!!!!")
         return ""
     else:
         answer = analyzer.concretize_typ(result)

### 2025-07-03 src/tapas/slim/analyzer.py
try out some ideas to debug

@@ -2350,6 +2350,46 @@ class Solver:
 
         return all_parts_consistent() and some_parts_consistent()
 
+    def check_closed_variable_elimination_new(self, world : World, lower : TVar, upper : Typ) -> list[World]:
+        upper_parts = pset(
+            st.upper
+            for st in world.constraints
+            if st.lower == lower 
+        ).union(self.extract_factored_upper_bounds(world, lower.id))
+
+        if any(
+            isinstance(upper_part, TVar) and upper_part.id not in world.closedids
+            for upper_part in upper_parts
+        ):
+            worlds = [world]
+            last_constraints = [Subtyping(lower, upper)]
+        else:
+            closed_upper_parts = [
+                upper_part
+                for upper_part in upper_parts
+                if not isinstance(upper_part, TVar) or upper_part.id in world.closedids
+            ]
+            worlds = []
+            for upper_part in closed_upper_parts:
+                worlds += self.solve(world, upper_part, upper)
+            last_constraints = []
+
+              
+        constraints = self.sub_polar_constraints(True, world.constraints, lower.id, upper)
+        subbed_constraints = list(constraints.difference(world.constraints))
+        print(f"""
+~~~~~~~~~~~~~~~~~~~~~~~~~~
+DEBUG lower TVar Skolem
+~~~~~~~~~~~~~~~~~~~~~~~~~~
+{concretize_typ(lower)} <: {concretize_typ(upper)}
+~~~~~~~~~~~~~~~~~~~~~~~~~~
+        """)
+        return [
+            replace(w1, constraints = w1.constraints.union(last_constraints)) 
+            for w0 in worlds
+            for w1 in self.solve_multi(w0, subbed_constraints)
+        ]
+
     def check_closed_variable_introduction(self, world : World, lower : Typ, upper : TVar) -> bool:
 
         lower_parts = pset(
@@ -2371,6 +2411,41 @@ class Solver:
         all_parts_consistent = lambda : bool(self.solve_multi(world, subbed_constraints))
 
         return all_parts_consistent() and some_parts_consistent()
+
+    def check_closed_variable_introduction_new(self, world : World, lower : Typ, upper : TVar) -> list[World]:
+
+        lower_parts = pset(
+            st.lower
+            for st in world.constraints
+            if st.upper == upper 
[…]

### 2025-07-03 src/tapas/slim/typlib.py

@@ -16,8 +16,8 @@ Bool = (f"""
 """.strip())
 
 Nat = (f"""
-(LFP[R] BOT
-| (<zero> @) 
+(LFP[R] 
+ (<zero> @) 
 | (<succ> R)
 )
 """.strip())
@@ -31,8 +31,8 @@ Even = (f"""
 
 def List_(T): 
     return (f"""
-(LFP[R] BOT
-| (<nil> @) 
+(LFP[R] 
+ (<nil> @) 
 | (<cons> (({T}) * R))
 )
     """.strip())

### 2025-07-03 src/tapas/slim/exprlib.py
update with annotation consolidation

@@ -43,9 +43,12 @@ context_map = {
 "h" : f"{Nat} -> ALL[A] A -> A",
 "l" : List_(f"ALL[A] {Nat} -> A -> A"),
 "r" : f"(ALL[A] A -> ALL[B] B -> B) -> {Nat}",
-"sort" : f"(ALL[A] (A * A -> {Bool}) -> {List_('A')} -> {List_('A')})",
-"scalarCmp" : f"{Nat} * {Nat} -> {Bool}",
-"lexicoCmp" : f"{List_(Nat)} * {List_(Nat)} -> {Bool}",
+"sort" : f"(ALL[A] ((A * A) -> {Bool}) -> {List_('A')} -> {List_('A')})",
+"scalarCmp" : f"({Nat} * {Nat}) -> ({Bool})",
+"lexicoCmp" : f"({List_(Nat)} * {List_(Nat)}) -> ({Bool})",
+"hof" : f"ALL[A B C] ((A * B) -> C) -> (<thing> A) -> (<thing> B)",
+"mkpath" : f"ALL[A C] (A -> C) -> (A) -> (C)",
+"mkpair" : f"ALL[A B] (A -> B) -> (A * B)",
 }
 
 def ctx(names : list[str]) -> PMap[str, str]:
@@ -192,10 +195,18 @@ let lted = {lted} in
 # ''')
 
 length = (f"""
-loop(case self => (
-    case ~nil @ => ~zero @ 
-    case ~cons (x, xs) => ~succ (self(xs)) 
-)) 
+loop({{ self => 
+    {{ nil;@ => zero;@  }}
+    {{ cons;(x,xs) => succ;(self(xs)) }}
+}}) 
+""".strip())
+
+lengthy = (f"""
+let id = {{x => x}} in
+loop(id({{ self => 
+    {{ nil;@ => zero;@  }}
+    {{ cons;(x,xs) => succ;(self(xs)) }}
+}})) 
 """.strip())
 
 add = (f"""

### 2025-07-03 src/tapas/slim/language.py

@@ -167,14 +167,13 @@ def analyze(code : str, typing_context = m()) -> tuple[Optional[analyzer.Typ], s
         tc = parser.program([context])
         if tc.results == None:
             raise Exception("Parsing Error")
-        elif len(tc.results) == 1:
-            result = tc.results[0]
+        elif len(tc.results) > 0:
+            # result = tc.results[0]
             solver = parser._solver
-
-
-            print(f"""
+            for i, result in enumerate(tc.results):
+                print(f"""
 =======================
-analysis result
+analysis multi result {i}
 =======================
 typ:
 {analyzer.concretize_typ(result.typ)}
@@ -185,18 +184,16 @@ skolems:
 constraints:
 {analyzer.concretize_constraints(result.world.constraints)}
 =======================
-            """)
+                """)
 
-
-
-
-
-            t1 = solver.interpret_polar_typ(True, result.world.closedids, result.world.constraints, result.typ) 
-            influential_ids = result.world.closedids.union(analyzer.extract_free_vars_from_typ(s(), t1))
-            influential_constraints = analyzer.filter_constraints_by_all_variables(result.world.constraints, influential_ids)
-            t2 = solver.make_constraint_typ(True)(s(), result.world.closedids, influential_constraints, t1)
+            # t1 = solver.interpret_polar_typ(True, result.world.closedids, result.world.constraints, result.typ) 
+            # influential_ids = result.world.closedids.union(analyzer.extract_free_vars_from_typ(s(), t1))
+            # influential_constraints = analyzer.filter_constraints_by_all_variables(result.world.constraints, influential_ids)
+            # t2 = solver.make_constraint_typ(True)(s(), result.world.closedids, influential_constraints, t1)
+            t2 = solver.decode_polarity_typ(tc.results, True)
             return (t2, tc.toStringTree(recog=parser), parser._solver)
         else: 
+            # In what scenario would we expect be multiple results at the top level?
             print("!!!!!!!!!!!!!!!")
             print("result size: " + str(len(tc.results)))
             print("!!!!!!!!!!!!!!!")

### 2025-07-03 src/test_analysis_new.py

@@ -26,7 +26,17 @@ from tapas.slim import exprlib as el, typlib as tl
 from tapas.slim.exprlib import ctx 
 
 def test_max():
-    assert infer_typ(el.max) != "" 
+    assert infer_typ(el.max) 
+
+
+
+def test_length():
+    #TODO: why is cons distributed across the pair?
+    assert infer_typ(el.length)
+
+def test_lengthy():
+    # TODO: is this type due to lack of consolidation?  
+    assert infer_typ(el.lengthy)
 
 
 ###############################################################
@@ -671,14 +681,69 @@ def test_typing_structures_1():
     print(code)
     assert infer_typ(code, ctx(["scalarCmp", "lexicoCmp"]))
 
+
+
+def test_typing_structures_sanity():
+    #TODO: find way to speed up; parsing might be slow
+    code = f"""
+let f = {{ ((one;@), (two;@)) => three;@ }} in
+let g = {{ ((uno;@), (dos;@)) => tres;@ }} in
+let h = {{(a,b) => 
+    (
+    {{one;@ => f(a,b)}}
+    {{uno;@ => g(a,b)}}
+    ) (a)
+}} in
+h
+    """
+    print(code)
+    assert infer_typ(code, ctx(["hof"]))
+
+def test_typing_structures_sanity_path():
+    code = f"""
+let h = (
+    {{one;@ => three;@}}
+    {{uno;@ => tres;@}}
+)
+in
+let result : (
+    ((<uno> @) -> (<tres> @)) &
+    ((<one> @) -> (<three> @))
+) = mkpath(h) in
+result
+    """
+    print(code)
+    assert infer_typ(code, ctx(["mkpath"]))
+
+def test_identity_application():
+    # TODO: why is there so much extra clutter?
+    code = f"""
+let h = ({{self =>
+    {{zero;@ => nil;@}}
+    {{succ;n => cons;n}}
+}}
+) in
+h
+{{x => x }}(h)
+    """
+    print(code)
+    assert infer_typ(code, ctx(["mkpath"]))
+
+
+
 def test_typing_structures_2():
     code = f"""
 let stdCmp = {el.stdCmp} in
 let stdSort : (
     ({tl.List_(tl.Nat)} -> {tl.List_(tl.Nat)}) &
     ({tl.List_(tl.List_(tl.Nat))} -> {tl.List_(tl.List_(tl.Nat))})
-) = sort(stdCmp) in @
[…]

### 2025-07-03 src/tapas/slim/analyzer.py
move rewriting rules after variable rules

@@ -2498,24 +2498,24 @@ DEBUG lower TVar Skolem
         if self.count > self._limit:
             return []
 
-        print(f"""
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-DEBUG SOLVE:
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-count: {self.count}
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-closed:
-{world.closedids}
+#         print(f"""
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# DEBUG SOLVE:
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# count: {self.count}
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# closed:
+# {world.closedids}
 
-constraints:
-{concretize_constraints(world.constraints)}
+# constraints:
+# {concretize_constraints(world.constraints)}
               
-|-
-{concretize_typ(lower)}
-<:
-{concretize_typ(upper)}
-~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-        """)
+# |-
+# {concretize_typ(lower)}
+# <:
+# {concretize_typ(upper)}
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+#         """)
 
         #######################################
         #### Reflection ####
@@ -2543,38 +2543,6 @@ constraints:
 
         elif isinstance(lower, TVar) and lower.id in self.aliasing: 
             return self.solve(world, self.aliasing[lower.id], upper)
-
-        #######################################
-        #### Implication Rewriting ############
-        #######################################
-        elif isinstance(upper, Imp) and isinstance(upper.antec, Unio):
-            return self.solve(world, lower, Inter(
-                Imp(upper.antec.left, upper.consq), 
-                Imp(upper.antec.right, upper.consq)
-            ))
-
-        elif isinstance(upper, Imp) and isinstance(upper.consq, Inter):
-            return self.solve(world, lower, Inter(
-                Imp(upper.antec, upper.consq.left), 
-                Imp(upper.antec, upper.consq.right)
-            ))
-
-        elif isinstance(upper, TEntry) and isinstance(upper.body, Inter):
-            return [
-                m1
-                for m0 in self.solve(world, lower, TEntry(upper.label, upper.body.left))
-                for m1 in self.solve(m0, lower, TEntry(upper.label, upper.body.right))
-            ]
-
[…]

### 2025-07-03 src/tapas/slim/language.py

@@ -224,19 +224,21 @@ def infer_typ(code : str, context = {}) -> str:
     print(f"TIME: {end - start}")
     if result == None:
         return ""
-    if result == analyzer.Top():
-        print("~~~~~~~~~~~TOP!!!!!!!")
-        return ""
     else:
-        answer = analyzer.concretize_typ(result)
-        print(f"""
-=========================================================================================================================
-INFERENCE: 
-~~~~~~~
-{answer}
-=========================================================================================================================
-        """)
-        return answer
+        simpres = solver.simplify_typ(result)
+        if simpres == analyzer.Top():
+            print("~~~~~~~~~~~TOP!!!!!!!")
+            return ""
+        else:
+            answer = analyzer.concretize_typ(simpres)
+            print(f"""
+    =========================================================================================================================
+    INFERENCE: 
+    ~~~~~~~
+    {answer}
+    =========================================================================================================================
+            """)
+            return answer
 
 def solve_subtyping(a : str, b : str) -> list[analyzer.World]:
     x = parse_typ(a)

### 2025-07-04 src/tapas/slim/analyzer.py
update upper diff rule in implementation  with syntax restriction

@@ -2326,6 +2326,19 @@ class Solver:
         else:
             return False
 
+    def is_subtractable_typ(self, t : Typ) -> bool:
+        if self.is_pattern_typ(t):
+            return True
+        else:
+            return (
+                isinstance(t, Exi) and
+                self.is_pattern_typ(t.body) and
+                not bool(t.constraints) and 
+                not bool(extract_free_vars_from_typ(s(), t)) and
+                True
+            )
+
+
     def is_base_typ(self, t : Typ) -> bool:
         return (
             # isinstance(t, TTag) or
@@ -2815,28 +2828,11 @@ DEBUG lower TVar Skolem
         #######################################
 
         elif isinstance(upper, Diff): 
-
-            # if diff_well_formed(upper): # TODO: change to: DF(lower) and DF(upper)
-            if True:
-                # TODO: need a sound/safe/conservative inhabitable check
-                # only works if we assume T is not empty
-                '''
-                T <: A \\ B === (T <: A) and (T is inhabitable --> ~(T <: B))
-                ----
-                T <: A \\ B === (T <: A) and ((T <: B) --> T is empty)
-                ----
-                T <: A \\ B === (T <: A) and (~(T <: B) or T is empty)
-                '''
-                context_worlds = self.solve(world, lower, upper.context)
-                return [
-                    m
-                    for m in context_worlds 
-                    # if (
-                    #     # not self.is_inhabitable(world, lower) or 
-                    #     # Fail (incompletely) if lower is empty
-                    #     self.solve(m, lower, upper.negation) == []
-                    # )
-                ]   
+            if self.is_pattern_typ(upper.context) and self.is_subtractable_typ(upper.negation):
+                if not bool(self.solve(world, lower, upper.negation)):
+                    return self.solve(world, lower, upper.context)
+                else:
+                    return []
             else:
                 return []
 

### 2025-07-04 src/tapas/slim/language.py

@@ -232,15 +232,11 @@ def infer_typ(code : str, context = {}) -> str:
         else:
             answer = analyzer.concretize_typ(simpres)
             print(f"""
-    =========================================================================================================================
-    INFERENCE: 
-    ~~~~~~~
-    {answer}
-
-    RAW: 
-    ~~~~~~~
-    {simpres}
-    =========================================================================================================================
+=========================================================================================================================
+INFERENCE: 
+~~~~~~~
+{answer}
+=========================================================================================================================
             """)
             return answer
 

### 2025-07-05 src/test_analysis_new.py
add test and notes on backtracking

@@ -55,6 +55,25 @@ def test_intersection_arrow_subtypes_lfp_arrow():
     """
     )
 
+def test_intersection_arrow_single_selection():
+    worlds = solve_subtyping(f"""
+(<true> X -> @) & (<false> X -> @)
+    """, """
+(<true> @) -> @
+    """
+    )
+    assert worlds
+
+    for i, world in enumerate(worlds):
+        print(f"""
+=========================
+world {i}
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+constraints:
+{analyzer.concretize_constraints(world.constraints)}
+=========================
+        """)
+
 
 
 def test_length():
@@ -977,6 +996,8 @@ if __name__ == '__main__':
     # test_max_app()
     # test_subtyping_unrolling()
     # test_length_eta_expansion()
+    # test_intersection_arrow_subtypes_lfp_arrow()
+    test_intersection_arrow_single_selection()
 
 
 #######################################################################

### 2025-07-06 src/tapas/slim/analyzer.py
update intersection elim rewriting rule

@@ -3189,12 +3189,14 @@ SOLVABLE:
             worlds = self.solve(world, lower.left, upper) + self.solve(world, lower.right, upper)
             if worlds:
                 return worlds
-            elif isinstance(upper, Imp) and self.is_finite_paths_typ(lower):
+            # elif isinstance(upper, Imp) and self.is_finite_paths_typ(lower):
+            elif isinstance(upper, Imp):
                 paths = linearize_intersections(lower)
                 param_types = []
                 return_types = []
                 for path in paths:
-                    assert isinstance(path, Imp)
+                    if not isinstance(path, Imp):
+                        return []
                     param_types.append(path.antec)
                     return_types.append(path.consq)
                 return self.solve(world, Imp(make_unio(param_types), make_unio(return_types)), upper)
@@ -3605,6 +3607,7 @@ class ExprRule(Rule):
         # Factored Type
         ################################################
         param_var = self.solver.fresh_type_var()
+        # TODO: simplify find_factors so it only needs a path, and not a fresh var
         param_typs = find_factors_from_pattern(rel_typ, TEntry("head", param_var), param_var)
         assert len(param_typs) == 1
         param_typ = list(param_typs)[0]

### 2025-07-06 src/test_analysis_new.py
update annotation rule with consolidation

@@ -26,6 +26,13 @@ from tapas.slim import exprlib as el, typlib as tl
 from tapas.slim.exprlib import ctx 
 
 
+def test_free_var_annotation():
+    code = f"""
+let x : X = uno;@ in
+{{dos;@ => @}}(x)
+    """
+    assert not infer_typ(code) 
+
 def test_lted():
     assert infer_typ(el.lted) 
 
@@ -1000,7 +1007,8 @@ if __name__ == '__main__':
     # test_intersection_arrow_single_selection()
     # test_typing_C3()
     # test_max_app()
-    test_variable_subtypes_tag_pair()
+    # test_variable_subtypes_tag_pair()
+    test_free_var_annotation()
 
 
 #######################################################################

### 2025-07-06 src/test_analysis_new.py
update code and rules related to factoring

@@ -1012,6 +1012,8 @@ if __name__ == '__main__':
     # test_typing_structures_2()
     # test_max()
 
+    # test_recursive_relational_factorization_learning_in_typing()
+
 
 
 

### 2025-07-08 src/test_analysis_new.py
update

@@ -1010,7 +1010,7 @@ if __name__ == '__main__':
     # test_variable_subtypes_tag_pair()
     # test_free_var_annotation()
     # test_typing_structures_2()
-    # test_max()
+    test_max()
     # test_recursive_relational_factorization_learning_in_typing()
 
 

### 2025-07-08 extrinsic_type_inference_paper/paper.tex
update

\documentclass[acmsmall]{acmart}
% \documentclass[letterpaper]{llncs}
% \usepackage[letterpaper, margin=1.5in]{geometry}

\usepackage{multicol}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{nccmath}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage[scaled]{beramono}
\usepackage[T1]{fontenc}
\usepackage[dvipsnames]{xcolor}
\usepackage{array}
\usepackage{wasysym}

\usepackage{pifont}% http://ctan.org/pkg/pifont

\usepackage{graphicx}
\graphicspath{ {./images/} }

\usepackage{url}

\newcounter{pdc}
\newcounter{sdc}

\makeatletter % allow us to mention @-commands
\def\arcr{\@arraycr}
\makeatother

\lstset{
    % identifierstyle=\color{violet},
    % textcolor=blue,
    % keywordstyle=\color{blue},
    % keywordstyle=\text,
    basicstyle=\ttfamily\small,
    % mathescape=true,
    % showspaces=false,
    % morekeywords={def, fix, in}
}

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{property}{Property}[section]
% \newtheorem{proof}{Proof}[section]

\newtheorem{experiment}{Experiment}[section]

\title{Extrinsic Type Inference (DRAFT 2025-06-08)}
\author{Thomas Logan}

\begin{document}

\newcommand{\pdc}{\stepcounter{pdc}\arabic{pdc}}
\newcommand{\sdc}{\stepcounter{sdc}\arabic{sdc}}

\newcommand{\Par}[1]{\paragraph{\textbf{#1}}}
\newcommand{\J}[1]{\texttt{\fontfamily{pcr}\selectfont #1}}
\newcommand{\lab}[1]{\text{\color{Gray}\ [#1]}}
\newcommand{\entails}{\vdash}
\newcommand{\satisfies}{\vDash}
\newcommand{\given}{\dashv}
\newcommand{\with}{\ \diamond\ }
\newcommand{\notfree}{\ \#\ }
\newcommand{\consis}{\ \star}
\newcommand{\safe}{\ \checkmark}
\newcommand{\relational}{\ \Re}

% \newcommand{\cmark}{\ding{51}}%
% \newcommand{\xmark}{\ding{55}}%
\newcommand{\pass}{\text{\normalsize \color{Green} \ding{51}}}
\newcommand{\assisted}{\text{\normalsize \color{Purple} \ding{45}}}
\newcommand{\fail}{\text{\normalsize \color{red} \ding{55}}}

\newcommand{\unique}{\ \dagger}

\newcommand{\factorsinto}{\Vdash}

\newcommand{\allsafe}{\ \Re\checkmark}

\newcommand{\qua}{\ .\ }

\newcommand{\ignore}{\varnothing}
\newcommand{\dTheta}{\dot{\Theta}}
\newcommand{\closed}{\bullet}
\newcommand{\local}{\blacktriangle}
\newcommand{\open}{\circ}

\newcommand{\up}{\ \land\ }
\newcommand{\down}{\ \lor\ }

% \newcommand{\tl}{\textasciitilde{}}

\newcommand{\multi}[1]{\widebar{\ #1\ }}
\newcommand{\hastype}{:}
\newcommand{\pattype}{\ \lozenge\ }
\newcommand{\liftfun}{:}

\newcommand{\sz}[1]{\text{\small #1}}

\newcommand{\block}[1]{{\footnotesize \[
\begin{array}[t]{l}
#1
\end{array}
\]}}

\newcommand{\cb}[1]{
\begin{array}[t]{l}
\\
#1
\\
\end{array}
}

\newcommand{\ms}[1]{{\footnotesize $(#1)$}}

\newcommand{\code}[1]{\ms{\J{#1}}}

\newcommand{\R}[1]{\color{teal}\ \ [\emph{#1}]}

\newcommand{\subtypes}{<:}
\newcommand{\supertypes}{:>}
\newcommand{\I}{\hspace{4mm}}
\newcommand{\Z}{.\hspace{4mm}}
\newcommand{\Alpha}{\mathrm{A}}
\newcommand{\Tau}{\mathrm{T}}
\newcommand{\B}[1]{\textbf{#1}}
\newcommand{\F}[1]{\text{#1}}
\newcommand{\bigand}{\bigwedge\nolimits}
\newcommand{\bigor}{\bigvee\nolimits}
\newcommand{\C}[1]{\color{teal} \rhd\ \emph{#1}}
\newcommand{\com}[1]{\I \emph{#1}}
\newcommand{\D}[1]{\small \textsc{#1}}
% \newcommand{\fig}[1]{Fig. {\color{red} \fig{#1}}}
\newcommand{\FIG}[1]{Fig. {\color{red} \ref{#1}}}
\newcommand{\TODO}[1]{\noindent \B{\color{red} TODO: #1}}

\newcommand{\is}{\ ::=\ }
\newcommand{\sep}{\ \ |\ \ }
\newcommand{\nonterm}[1]{#1\ }
\newcommand{\contin}{|\ \ \ \ \ \ \ }
\newcommand{\wrt}{\B{wrt }}
% \newcommand{\let}{\B{def }}
% \newcommand{\for}{\B{for }}

\newcommand{\pc}{\hfill \break \noindent \ $\diamond$ \ }

\newcommand{\N}{\ $\cdot$\ \ }

\newcommand{\pitem}{\item[$\diamond$]}

\newcommand{\tl}{\textasciitilde{}}
\newcommand{\typdiff}{\J{\textbackslash}}

\maketitle

\section{Introduction}
\label{sec:introduction}

Type inference in practical programming languages has historically relied on intrinsic types.
An expression has an intrinsic type if it is annotated with a type, such that the type
constrains the control flow of the program.  
Intrinsic types can occur in at least three ways: by annotating
expressions, constructions, or patterns.     
For example, the Standard ML expression \code{fn x => (f x)}
could be annotated with \code{val f' : int -> int = fn x => (f x)},
which artificially prunes the allowed control flow of expressions that use \code{f'}.
As for constructions and sums of patterns, consider a Standard ML program that computes the length of a list,
using \code{nil} and \code{cons} as data constructors representing lists, 
and using \code{zero} and \code{succ} as data constructors representing natural numbers.

\block{
  \J{fun len(nil) = zero}
  \\
  \J{\ \ | len(cons(x,xs)) = succ(len(xs))}
}

\noindent
Standard ML uses the datatype mechanism to specify intrinsic types on both patterns and structures.

\block{
  \J{datatype nat = zero | succ of nat} 
  \\
  \J{datatype 'a list = nil | cons of 'a * 'a list}
}

\noindent
For the constructions with \code{zero} and \code{succ}, the nat specification 
means that the weakest type the constructions could belong to is \code{nat},
which forbids mixing the constructions with those not specificated in the datatype, like \code{zero} with \code{nil}:

\block{
  \J{fn nil => zero}
  \\
  \J{\ | cons(x,xs) => nil}
}

\noindent
As for the patterns with \code{nil} and \code{cons}, the list specification 
means that the strongest assumed for the sum of the patterns
is \code{'a list}. 
Thus, it forbids pattern matching fewer patterns than those in the datatype.

\block{
  \J{fun nlen(nil) = zero}
}

\noindent
Additionally, the list specification 
means that the strongest assumption allowed for \code{cons} is that its
content has type \code{'a * 'a list}. Thus, 
it forbids overspecifying pattern contents,
like a length function that only works on even length lists.

\block{
  \J{fun evlen(nil) = zero}
  \\
  \J{\ \ | evlen(cons(x,cons(y,zs))) = succ(succ(evlen(zs)))}
}

\noindent
With the datatype being both the strongest type of a sum of patterns and also the weakest
type of each construction, there is no additional loss in expressivity by
requiring that the weakest type of a sum of patterns to be the datatype, 
or inferring the strongest type of each construction to be the datatype.

Even with all the intrinsic specifications required for structural expressions, Standard ML,
and languages like it, ease the burden on the programmer by leveraging Hindley-Milner
type inference \cite{}, which can fairly general predicative types of 
non-structural expressions extrinsically, without intrinsic specifications. 
Despite the mitigating type inference capabilities of intrinsically typed systems, 
much more software these days is written
in untyped scripting languages, like Python and Javascript. 
These programs are not written with any control flow restrictions in mind,
and the are unlikely to conform to the control flow restrictions imposed by intrinsic types.
In fact, the lack of control flow restrictions may be one of the reasons that people
choose scripting languages over intrinsically typed languages.
Without artificial restrictions, programers can write code with a high degree of reusability,
with very little effort.
For instance, one might want a comparison function that works on both numbers and strings,
so that the caller can sort lists of numbers or lists of strings without having to
clutter the call site with a particular comparison function.  
Some have tried reconcile intrinsic types with code reusability by adding
advanced construct to languages, classes and type classes.
Indeed, these constructs do enable reusability, but they require additional
complexity and may fragment code in ways that obscures its essence.   
An alternative approach is to simply relax the intrinsic type requirements,
and infer types extrinsically according to the control flow of the program.

There have been some relaxations of intrinsic types to enable sightly more code reusability.
Refinement types (Pfenning) \cite{} removed the intrinsic strongest type of sums of patters,
allowing pattern matching with incomplete pattern coverage. It inferred stronger
types too, by using intersection against the intrinsic weakest type to find safe but stronger types
that could be used in functions with stronger assumptions. 
Unfortunately, refinements aren't sufficient to handle the full range of structural mixing 
and code reusability allowed by untyped languages.
More recently, SuperF \cite{} demonstrated removal the the weakest intrinsic type,
which enabled it infer principal types with subtyping constraints, 
where second-order parametric
types were insufficient. However, SuperF was not concerned with how functions
with mixed structures could be represented literally.
On the other hand, MLstruct \cite{} consisted of a detailed language design that allows
constructing functions that mix structures, as long as information about the constructors
are specified in particular ways. 
MLStruct relaxed intrinsic requirements in a few ways, but still required a heavy annotation
burden and restricted control flow in a significant way.

This work tackles these limitations, in order to achieve a language design and type inference system,
such that there are almost no annotations are required, code reusability is uninhibited, and error-free programs
are guaranteed. The system is extremely extrinsic, requiring no annotations on structures.  
Despite the lack of assistance from intrinsic types, our system appears to be able to admit
nearly all the programs admitted by MLStruct and SuperF.
Moreover, our system's lack of certain intrinsic restrictions, which exists in MLstruct, 
clears the way for automatic reasoning by induction, which allows even greater code reusability
without sacrificing safety.

\section{Overview}
\label{sec:overview}
We designed the type system around a simple applicative language with pattern matching,
since such a language is syntactically close to predicate logic. Likewise,
we designed the language of types to share many similarities with predicate logic. 
The extrinsic type system is able to infer very precise types by analyzing the control
flow of program in the form of generating and solving subtyping constraints. 
In many cases the type system can infer the weakest possible types of parameters,
and the strongest possible (i.e. most general or principal) types of results,
all without inhibiting the reusability the code.
To get a feel for the languages and reasoning capabilities, it's best to start by
looking at a few examples.

\Par{Finite Isomorphism.}
A function can be constructed from multiple paths. 
Consider the function \ms{\J{talky}}, which completes a simple English phrase.

\block{
  \J{def talky = (}
  \\
  \I \J{[hello;@ => world;@]}
  \\
  \I \J{[good;@ => morning;@]} 
  \\
  \I \J{[thank;@ => you;@]} 
  \\
  \J{)}
}

\noindent
Since there is no recursion, the number of paths is finite.
For these simple finite functions, the type system constructs
an isomorphic type without any weakening in precision.

\block{
  \J{(<hello>@ -> <world>@) \&}
  \\
  \J{(<good>@ -> <morning>@) \&} 
  \\
  \J{(<thank>@ -> <you>@)} 
}

\noindent
Each path is represented using the path type,
and the paths are joined together using the intersection type.

\Par{Selection.}
Although the type of function $\J{talky}$ is no weaker than the actual function,
lifting into the type syntax avails it to approximate reasoning at application.
Consider an application of the function to a variable argument.

\block{
  \J{def x : <hello> @ | <thank> @ = ... in}
  \\
  \J{...}
  \\
  \J{talky(x)} 
  \\
  \J{...}
}

\noindent
If the type system affirms the type annotation,
then it goes on to infer the type of the application.

\block{
  \J{<world> @ | <you> @}
}

\noindent
In effect, the type system applies application to all possible arguments
and unions the results derived from only the paths of those arguments.

\Par{Learning.}
The paths of a function are not always syntactically apparent. 
If a variable is applied to arguments, the variable should implicitly behave 
like a function, or else the program is erroneous. 
Consider the higher order function $\J{foo}$ which takes a function as input
and applies it to the number two and the boolean true.

\block{
  \J{def foo = ? f => (f(<succ> <succ> <zero> @), f(<true> @))}
}

\noindent
Type systems for System F have traditionally struggled to find the best
type for such programs. If restricted to unqualified parametric types, there are many choices,
without there being a clear best or principal type, 
which is because it's not possible to find the weakest type for the function parameter
$\J{f}$.
One possibility is that the type system requires that $\J{f}$ is a function that wraps its
input in an option.

\block{
  \J{f : ALL[X] X -> <some> X}
}

\noindent
Another possibility is that the type system requires that $\J{f}$ is a function that wraps its
input in an list.

\block{
  \J{f : ALL[X] X -> (LFP[R] <nil> @ | <cons> X*R)}
}

\noindent
Both are valid, but neither is weaker than the other. The problem is that unqualified parametric
types are insufficient for describing the weakest safe input type.
It was not until SuperF introduced a technique for solving subtyping constraints
that a principal type could be automatically constructed for functions like $\J{foo}$
Our extrinsic type system relies on similar principles of solving subtyping constraints,
and likewise, can construct the weakest safe type for $\J{f}$, and thus the principal type for $\J{foo}$.

\block{
  \J{ALL[X Y] (<succ> <succ> <zero> @ -> X) \& (<true> @ -> Y) -> X*Y}
}

\noindent
Although the subtyping constraints are not present in the type's notation,
the intersection of paths is chosen because it represents the weakest type
consistent with the subtyping constraints on a variable representing the parameter's type. 
Suppose the type system chooses the variable type $\J{F}$ for the input parameter $\J{f}$.
Eventually, due to the applications of the parameter $\J{f}$, the type
system will attempt to find a solution to two constraints on $\J{F}$

\block{
  \J{F} \subtypes \J{(<succ> <succ> <zero> @ -> X)}
  \\
  \J{F} \subtypes \J{(<true> @ -> Y)}
}

\Par{Factorization.}
A function can represent infinite paths by using recursion.
Consider the function $\J{repeat}$, which takes an item and a natural number
as inputs and returns a list of that item repeated the input number of times.

\block{
  \J{def repeat = [x => loop([self => [}
  \\
  \I \J{[zero;@ => nil;@]}
  \\
  \I \J{[succ;n => <cons>(x,self(n))]}
  \\
  \J{])]}
}

\noindent
One possibility for constructing the type of the recursive function would
be to follow the idea of finite isomorphism and construct a type 
that preserves all the paths. If our type system had a greatest fixed point
type constructor, then we could concisely represent such a type with minimal
rewriting.

\block{
  \J{ALL[T] T -> GFP[R](} 
  \\
  \I \J{(<zero> @ -> <nil> @) \&}
  \\
  \I \J{(ALL[N L](R <: N->L) <succ> N -> <cons> T*L)}
  \\
  \J{)} 
}

\noindent
Although the strength of precision of this type would be very beneficial,
reasoning with such a type would be complicated. To keep the type system as
simple as possible for the purpose of extrinsic type inference with a high-degree
of code reusability, we leave such relational reasoning of recursive and co-recursive types out of
the type system. 
In this work, the type system actually avoids constructing a relational type, and instead, it
constructs a type that is significantly weaker than the actual behavior of the function,
by factoring the paths into a recursive space of inputs and a recursive space of outputs.

\block{
  \J{ALL[T] T ->} 
  \\
  \I \J{(LFP[R] <zero> @ | <succ> R)}
  \\
  \I \J{->}
  \\
  \I \J{(LFP[R] <nil> @ | <cons> T*R)}
}

\noindent
This eager weakening approximation enables the type system
to simply reason with recursive scalars and avoids the need for complicated
relational reasoning. In a follow up paper,
we will show how to remedy this limitation by augmenting the type system with
mechanisms for inductive relational reasoning.

\Par{Scalar Inflation.} 
Since the inputs to functions may have recursive types, the type 
system must be able to check for errors at function application by unrolling
recursive types.
Consider applying the function $\J{repeat}$ to some item and the number two, 
which results in the type system checking that two is in fact an instance of the natural numbers.

\block{
  \J{ <succ> <succ> <zero> @} \subtypes \J{(LFP[R] <zero> @ | <succ> R)}
}

\noindent
The type system confirms that two is a valid input by unrolling the recursive type three times.
Unfortunately, because the type system eagerly factored the type of $\J{repeat}$, it
is unable to infer a stronger type for the application than the original list type. 
In the followup paper, inductive relational reasoning will remedy this limitation.

\Par{Scalar Induction.} 
Another scenario that may arise is one in which the input to a function has an recursive type.
Consider the application of the function $\J{repeat}$ to a variable whose type represents the even natural numbers.

\block{
  \J{def x : (LFP[R] <zero> @ | <succ> <succ> R) = ... in}
  \\
  \J{...}
  \\
  \J{repeat(...)(x)} 
  \\
  \J{...}
}

\noindent
The application results in the type system checking that natural numbers are weaker than even numbers.

\block{
  \J{LFP[R] <zero> @ | <succ> <succ> R}
  \\
  \subtypes
  \\
  \J{LFP[R] <zero> @ | <succ> R}
}

\noindent
Since even numbers are represented recursively, the type system performs mathematical induction,
proving under the assumption that the subtyping relation holds for the self reference of the even numbers.

\block{
  \J{E} \subtypes \J{(LFP[R] <zero> @ | <succ> R)}
  \\
  \entails
  \\
  \J{(<zero> @ | <succ> <succ> E)} \subtypes \J{(LFP[R] <zero> @ | <succ> R)}
}

\noindent
In practice, the type system actually substitutes the supertype into
the self reference of the subtype, if the self reference occurs in a positive position of the subtype.

\block{
  \J{(<zero> @ | <succ> <succ> (LFP[R] <zero> @ | <succ> R))} 
  \\
  \subtypes 
  \\
  \J{(LFP[R] <zero> @ | <succ> R)}
}

\noindent
In effect, by using mathematical induction, the type system ensures that 
every type contained within the type of even natural numbers is also
contained within the type of natural numbers.
In followup work, we will extend this inductive reasoning to handle inductive relations.

\Par{Structural Generalization} 
Functions on lists can be generic, in that they may be independent of the item values, as in \code{repeat}.
Additionally, they may factor out the details that depend on item values, as in \code{sort}.

\block{
  \J{def sort : ALL[T] (T*T->} \textbf{Bool} \J{) -> }\textbf{List}(\J{T}) \J{ -> } \textbf{List}(\J{T}) \J{ = }
  \\
  \J{[ cmp => ...] in ...}
}

\noindent
The notation \ms{\textbf{Bool}} and \ms{\textbf{List}(\J{X})} are meta-language terms representing types.
The function $\J{sort}$ can be specialized by passing in different implementations of comparison functions.
Suppose we already have one scalar comparison function and one lexicographic comparison function.

\block{
  \J{def scalarCmp : }\textbf{Nat} \J{ -> } \textbf{Nat} \J{ -> } \textbf{Bool} \J{ = ... in}
  \\
  \J{def lexicoCmp : }\textbf{List}(\textbf{Nat}) \J{ -> } \textbf{List}(\textbf{Nat}) \J{ -> } \textbf{Bool} \J{ = ... in} 
  \\
  \J{...} 
}

\noindent
In Standard ML, the programmer would need a different specialization for every predefined
datatype.

\block{
  \J{def scalarSort = sort(scalarCmp) in}
  \\
  \J{def lexicoSort = sort(lexicoCmp) in} 
  \\
  \J{...} 
}

\noindent
Having many different specializations burdens the client of
the generic function to know many different names for roughly the same behavior and can obscure the meaning
of code at the call site. 
Untyped languages or extrinsically typed languages can bypass this rigidity found in intrinsically typed by specializing in multiple ways.

\block{
  ...
  \\
  \J{def genCmp = [a,b => } 
  \\
  \I \J{a |> (}
  \\
  \I\I \J{[zero;@ => scalarCmp(a,b)]}
  \\
  \I\I \J{[succ;n => scalarCmp(a,b)]}
  \\
  \I\I \J{[nil;@ => lexicoCmp(a,b)]}
  \\
  \I\I \J{[cons;(x,xs) => lexicoCmp(a,b)]}
  \\
  \I \J{)}
  \\
  \J{] in} 
  \\
  \J{def genSort : (TOP}
  \\
  \I \J{\& (}\textbf{List}(\textbf{Nat}) \J{ -> } \textbf{List}(\textbf{Nat})
  \\
  \I \J{\& (}\textbf{List}(\textbf{List}(\textbf{nat})) \J{ -> } \textbf{List}(\textbf{List}(\textbf{Nat}))
  \\
  \J{) = sort(genCmp) in}
  \\
  \J{...} 
}

\noindent
By combining the scalar and lexicographic comparison procedures into a single comparison
procedure, a generalized sort function can be used to work on either scalars or vectors,
thereby removing the burden of choosing the proper comparison function at the call site.

\section{Dynamic System}
\label{sec:dynamic_system}

\noindent
To demonstrate the capabilities of our extrinsic type system, 
we rely on an applicative programming language with pattern matching.
The design attempts to keep the forms of expressions and types relatively 
similar to each other when their semantics are also similar. 
Although the expression language is based on pattern matching, 
the type inference techniques apply just as well to languages
that are based on conditional branching and destructors instead of pattern matching.
In fact, conditional branching and destructors can be encoded from pattern matching,
and used without any loss in type inference precision.

\begin{definition}
  \label{def:expression}
  \small
  \[\begin{array}[t]{rr@{}l}
    \com{expression} &
    \nonterm{e} 
    & 
    \is 
    \begin{array}[t]{@{}l}
      x \sep 
      \J{@} \sep
      r \sep
      f \sep 
      e\J{.}l \sep
      e\J{(}e\J{)} \sep
      \J{loop(}e\J{)} \sep
      \J{def}\ x\ \J{:}\ \tau\ \J{=}\ e\ \J{in}\ e \sep
      s
    \end{array}
    \\
    \com{record} &
    \nonterm{r} & \is \epsilon \sep r\ \J{<}l\J{>}e 
    \\
    \com{function} &
    \nonterm{f} & \is \epsilon \sep f\ \J{[}p\J{=>}e \J{]}
    \\
    \com{sugar} &
    \nonterm{s} 
    & 
    \is 
    \begin{array}[t]{@{}l}
      l \J{;} e \sep
      e \J{,} e \sep
      e\ \J{|>}\ e \sep
      \J{if}\ e\ \J{then}\ e\ \J{else}\ e \sep
      \J{def}\ x\ \J{=}\ e\ \J{in}\ e
    \end{array}
  \end{array}\]
\end{definition}

\noindent
An expression is a variable \ms{x}, unit \code{@},
a record \ms{... \J{<}l\J{>} e ...}, a function \ms{...\J{[}p \J{=>} e \J{]}...}, 
a projection \ms{e\J{.}l}, an application \ms{e\J{(}e\J{)}}, 
a loop \ms{\J{loop(}e\J{)}},
an annotation \ms{\J{def}\ x\ \J{:}\ \tau\ \J{=}\ e\ \J{in}\ e},
or sugar.
These forms are similar to those seen in other applicative structural languages,
with a few choices that are a bit less common. One thing that may be
a bit unconventional is that a function may syntactically consist of 
multiple paths guarded by patterns.

\begin{definition} 
  \label{def:expression_desugaring}
  \emph{expression desugaring}
  \hfill 
  \small
  \boxed{\llbracket s \rrbracket^\square = e}
  \\
  \begin{align*}
      \llbracket l \J{;} e \rrbracket^\square &= \J{<}l \J{>}\ e
      \\
      \llbracket e_l \J{,} e_r \rrbracket^\square &= \J{<left>}\ e_l \ \J{<right>}\ e_r
      \\
      \llbracket e_l\ \J{|>}\ e_r \rrbracket^\square &= e_r\J{(}e_l\J{)}
      \\
      \llbracket \J{if}\ e_c\ \J{then}\ e_t\ \J{else}\ e_f \rrbracket^\square &= 
      \begin{array}[t]{@{} l}
        e_c \J{ |> (}
          \\
          \I \J{[true;@ => } e_t \J{]}
          \\
          \I \J{[false;@ => } e_f \J{]}
          \\
        \J{)}
      \end{array}
      \\
      \llbracket \J{def}\ x\ \J{=}\ e\ \J{in}\ e' \rrbracket^\square &= \J{[}x \J{=>} e'\J{]}\J{(}e\J{)}
  \end{align*}

\end{definition}

\noindent
For convenience and clarity, there is sugar for 
singleton records \ms{l\J{;}e},
tuples \ms{e\J{,}e}, 
unannotated bindings \ms{\J{def}\ x\ \J{=}\ e\ \J{in}\ e}, flow \ms{e\ \J{|>}\ e},
and conditional branching \ms{\J{if}\ e\ \J{then}\ e\ \J{else}\ e}.
For any metalanguage statement or term containing sugar $s$, we assume that $s$ is implicitly replaced with $\llbracket s \rrbracket^\square$.

\begin{definition}
  \label{def:pattern}
  \[\begin{array}[t]{rr@{}l}
    \com{pattern} &
    \nonterm{p} & \is 
      x \sep
      \J{@} \sep
      k
    \\
    \com{record} &
    \nonterm{k} & \is \epsilon \sep k\ \J{<}l\J{>}p
  \end{array}\]
\end{definition}

\noindent
A pattern is either unit or a record consisting of patterns.

\begin{definition}
  \label{def:value}
  \[\begin{array}[t]{rr@{}l}
    \com{value} &
    \nonterm{v} & \is 
      \J{@} \sep
      g \sep
      f
    \\
    \com{record} &
    \nonterm{g} & \is \epsilon \sep g\ \J{<}l\J{>}v
  \end{array}\]
\end{definition}

\noindent
A value may be unit, a function, or a record consisting of only values.

\begin{definition}
  \label{def:progression}
  \emph{progression}
  \hfill 
  \small
  \boxed{e \rightsquigarrow e}
  \\
  \begin{mathpar}
    \inferrule {
      e \rightsquigarrow e'
    } {
      r\ \J{<}l\J{>}e \rightsquigarrow r\ \J{<}l\J{>}e'
    }

    \inferrule {
      r \rightsquigarrow r'
    } {
      r\ \J{<}l\J{>} v \rightsquigarrow r'\ \J{<} l \J{>} v
    }

    \inferrule {
      e \rightsquigarrow e'
    } {
      e\J{.}l \rightsquigarrow e'\J{.}l
    }

    \inferrule {
      \J{<}l\J{>}v \in g
      \\
      \forall\ e .\ \J{<}l\J{>}e \in g \implies e = v
    } {
      g\J{.}l \rightsquigarrow v
    }

    \inferrule {
      e_f \rightsquigarrow e_f'
    } {
      e_f\J{(}e\J{)} \rightsquigarrow e_f'\J{(}e\J{)}
    }

    \inferrule {
      e \rightsquigarrow e'
    } {
      f\J{(}e\J{)} \rightsquigarrow f\J{(}e'\J{)}
    }

    \inferrule {
      f\J{(}v\J{)} \rightsquigarrow e'
      \\
      \textbf{fv}(e) \subseteq \textbf{fv}(p)
    } {
      (f\ \J{[}p\J{=>}e\J{]})\J{(}v\J{)} \rightsquigarrow e' 
    }

    \inferrule {
      \exists e' .\ f\J{(}v\J{)} \rightsquigarrow e'
      \\
      v  \simeq p \given \sigma 
    } {
      (f\ \J{[}p\J{=>}e\J{]})\J{(}v\J{)} \rightsquigarrow e[\sigma]
    }

    \inferrule {
      e \rightsquigarrow e'
    } {
      \J{loop(}e\J{)} \rightsquigarrow \J{loop(}e'\J{)}
    }

    \inferrule {
    } {
      \J{loop([}x\J{=>}e\J{])} \rightsquigarrow e[x \slash \J{loop([}x\J{=>}e\J{])}]
    }
  \end{mathpar}
\end{definition}

\noindent
Progression of an expression is a fairly typical small-step operational semantics. 
Its most atypical aspect is handling the unconventional multi-path syntax of
functions.
Since a function may consist of multiple paths guarded by patterns, 
application subsumes pattern matching.
Recursion is enabled by the \code{loop} keyword, such that
\code{loop} behaves like a fixed point combinator.

\begin{definition}
  \label{def:value_map}
  \[\begin{array}[t]{rr@{}l}
    \com{value map} &
    \nonterm{\sigma} & \is \epsilon \sep \sigma\ x \slash v 
  \end{array}\]
\end{definition}

\noindent
A value map assigns variables to values.
Progression relies on interpretation maps to collect
values that are relevant from pattern matching \ms{v \simeq p \given \sigma},
and then to use them in a subsequent expression via substitution 
\ms{e[\sigma]}

\begin{definition}
  \label{def:type}
  \[\begin{array}[t]{rr@{}l}
    \com{type} &
    \nonterm{\tau} & \is 
    \begin{array}[t]{@{}l}
      \alpha \sep
      \J{@} \sep
      \kappa \sep 
      \phi \sep 
      \psi \sep 
      \J{LFP[}\alpha\J{]}\tau \sep
      \tau \typdiff \eta \sep
      \gamma
    \end{array}
    \\
    \com{implication} &
    \nonterm{\kappa} & \is 
    \begin{array}[t]{@{}l}
      \J{<} l \J{>} \tau \sep 
      \tau\J{->}\tau
    \end{array}
    \\
    \com{expansion} &
    \nonterm{\phi} & \is 
    \begin{array}[t]{@{}l}
      \tau\J{|}\tau \sep 
      \J{EXI[}\Theta\J{]} \Delta \J{:} \tau
    \end{array}
    \\
    \com{refinement} &
    \nonterm{\psi} & \is 
    \begin{array}[t]{@{}l}
      \tau\J{\&}\tau \sep 
      \J{ALL[}\Theta\J{]} \Delta \J{:} \tau
    \end{array}
    \\
    \com{negatable} &
    \nonterm{\eta} & \is 
    \begin{array}[t]{@{}l}
      \J{EXI[}\Theta \J{]} \Omega \J{:} \eta \sep 
      \rho
    \end{array}
    \\
    \com{pattern} &
    \nonterm{\rho} & \is 
    \begin{array}[t]{@{}l}
      \alpha \sep
      \J{@} \sep
      \J{<}l\J{>}\rho \sep 
      \rho\J{\&}\rho \sep
      \rho\J{*}\rho
    \end{array}
    \\
    \com{sugar} &
    \nonterm{\gamma} & \is 
    \begin{array}[t]{@{}l}
      \J{TOP} \sep \J{BOT} \sep 
      \tau \J{*} \tau 
    \end{array}
  \end{array}\]
\end{definition}

\noindent
Types provide the representation for approximate specification.
To accommodate a high degree of structure reusability,
it is important that the forms of types accommodate a wide range of precision.
To manage the complexity of the various forms, we subdivide the type forms
according to some themes.
A type \ms{\tau} may consist of unit \code{@}, variables \ms{\alpha}, 
implications, expansions, refinements,
least fixed points \ms{\J{LFP[}\alpha \J{]} \tau},
differences \ms{\tau \typdiff \eta}, and sugar.

An implication \ms{\kappa} may take the form of  
an entry \ms{\J{<}l\J{>}\tau} or a path \ms{\tau \J{->} \tau},
reflecting the structure of its inhabitants.
An expansion \ms{\phi} is either a union \ms{\tau \J{|} \tau} or 
an existential \ms{\J{EXI[}\Theta \J{]} \Delta \J{:} \tau},
representing weakenings of their constituent types. 
An existential is parameterized by type variables \ms{\Theta}
and positively qualified by subtyping constraints \ms{\Delta}, 
such that the weaker the qualification, the weaker the existential.
A refinement \ms{\psi} is either an intersection \ms{\tau \J{\&} \tau} or 
a universal \ms{\J{ALL[} \Theta \J{]} \Delta \J{:}\tau},
representing strengthenings of their constituent types. 
An universal is parametrized by type variables \ms{\Theta} 
and negatively qualified by subtyping constraints $\Delta$,
such that the weaker the qualification, the stronger the universal.

A least fixed point \ms{\J{LFP[}\alpha \J{]}}
represents an infinite space of structures, 
such as natural numbers or lists. 
A difference \ms{\tau \typdiff \eta} is a space without inhabitation by certain structures.
The subtracted part is restricted to a negatable form,
which is a form that our system can decide the inhabitation
assuming additional semantics restrictions.

\begin{definition} 
  \label{def:type_desugaring}
  \emph{type desugaring}
  \hfill 
  \small
  \boxed{\llbracket \gamma \rrbracket^\square = \tau}
  \\
  \begin{align*}
    \llbracket \J{TOP} \rrbracket^\square &= \J{EXI[} \alpha \J{]} \alpha
    \\
    \llbracket \J{BOT} \rrbracket^\square &= \J{ALL[} \alpha \J{]} \alpha
    \\
    \llbracket \tau_l \J{*} \tau_r \rrbracket^\square &= (\J{<left>}\ \tau_l)  \J{\&} (\J{<right>}\ \tau_r)
  \end{align*}
\end{definition}

\noindent
For convenience and clarity, types include syntactic sugar for top $\J{TOP}$, 
bottom $\J{BOT}$, and pairs $\tau \J{*} \tau$.
For any metalanguage statement or term containing sugar $\gamma$, we assume that $\gamma$ is implicitly replaced with $\llbracket \gamma \rrbracket^\square$.

\begin{definition}
  \label{def:sequence}
  \small
  \[\begin{array}[t]{rr@{}l}
    \com{type var seq} &
    \nonterm{\Theta} & \is \epsilon \sep \Theta\ \alpha 
    \\
    \com{subtyping seq} &
    \nonterm{\Delta} & \is \epsilon \sep \Delta \  \tau \J{<:} \tau
    \\
    \com{typing seq} &
    \nonterm{\Gamma} & \is \epsilon \sep \Gamma\ x \hastype \tau
    \\
    \com{schema seq} &
    \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
    \\
    \com{negatable type seq} &
    \nonterm{\Xi} & \is \epsilon \sep \Xi\ \eta 
    \\
    \com{negatable subtyping seq} &
    \nonterm{\Omega} & \is \epsilon \sep 
      \Omega\ \tau \J{<:} \eta \sep 
      \Omega\ \alpha \J{<:} \tau
    \\
    \com{composite seq} &
    \nonterm{C} & \is \epsilon \sep C\ \iota 
    \\
    \com{item} &
    \nonterm{\iota} & \is 
    \alpha
    \sep 
    \tau \J{<:} \tau
    \sep
    x \hastype \tau
    \sep
    (\Theta, \Delta, \tau)
    \sep
    \eta
    \sep
    \tau \J{<:} \eta

  \end{array}\]
\end{definition}

\noindent
There are various scenarios where objects should be collected into in sequences or sets.
For metalanguage operations or predicates that are generic with respect the form of a collection's items,
it is convenient to have a composite sequence.
A sequence is either empty \ms{\epsilon} or an extension,
such as \ms{C\ \iota} in the case of the composite sequence. 
For clarity, we omit writing \ms{\epsilon} explicitly when describing semantics or instances of the forms.
That is, for any structure, if there is a malformed structure due to a missing piece, 
we assume it is replaced by a well-formed structure, 
where \ms{\epsilon} is inserted wherever necessary to make it well-formed.

\begin{definition} 
  \label{def:sequence_setting}
  \emph{sequence setting}
  \hfill 
  \small
  \boxed{\textbf{set}(C) = S}
  \\
  \[
  \begin{array}[t]{r@{}c@{}l}
      \textbf{set}(\epsilon)
      &\ =\ & 
      \emptyset

      \\

      \textbf{set}(C\ \iota)
      &\ =\ & 
      \textbf{set}(C) \cup [ \iota ] 
  \end{array}
  \]
\end{definition}

\noindent
In order to leverage set arithmetic and its theory,
we implicitly convert sequences into sets. 
We assume that any sequence \ms{C} may be implicitly replaced with \ms{\textbf{set}(C)}.

\begin{definition} 
  \label{def:set_sequencing}
  \emph{set sequencing}
  \hfill 
  \small
  \boxed{\textbf{seq}(S) = C}
  \\
  \[
  \begin{array}[t]{r@{}c@{}l}
      \textbf{seq}(\emptyset)
      &\ =\ & 
      \epsilon

      \\

      \textbf{seq}(S \cup [\iota])
      &\ =\ & 
      \textbf{seq}(S \backslash [\iota])\ \iota 
  \end{array}
  \]
\end{definition}

\noindent
Likewise, it also beneficial to convert sets back into sequences, to leverage
definitions that decompose sequences according to their structures. 
We assume that any set \ms{S} may be implicitly replaced with \ms{\textbf{seq}(S)}.

% \begin{definition}
%   \label{def:stability}
%   Stability 
%   \hfill
%   \boxed{\VDash e}
%   \\
%   \begin{mathpar}
%     \inferrule {
%       e = v
%     } {
%       \VDash e
%     }

%     \inferrule { 
%       e \rightsquigarrow e' 
%       \\
%       \VDash e'
%     } {
%       \VDash e
%     } 
%   \end{mathpar}
% \end{definition}

% \noindent
% A expression is stable if and only if it is a value or it can take a step to a stable expression.
% In other words, if an expression is not a value then its behavior is defined by progression.

% \begin{property}
%   \label{def:model_typing_soundness}
%   Model Typing Soundness 
%   \\
%   \begin{mathpar}
%     \inferrule { 
%       \delta \satisfies e \hastype \tau
%     } {
%       \VDash e 
%     } 
%   \end{mathpar}
% \end{property}

% \noindent
% Model typing $\delta \satisfies e \hastype \tau$ attempts to imbue types
% with meaning.  At the very least, types should inform us that an expression is stable.

\begin{definition}
  \label{def:type_map}
  \[\begin{array}[t]{rr@{}l}
    \com{type map} &
    \nonterm{\delta} & \is \epsilon \sep \delta\ \alpha \slash \tau
  \end{array}\]
\end{definition}

\noindent
A type map assigns variables to types.
Since a type may be constructed from variables, it 
is necessary to specify an interpretation of the variables
when judging the inhabitation of the type.
Thus, typing relies on type maps to provide interpretations
for type variables.

\begin{definition}
  \label{def:model_typing}
  \emph{model typing}
  \hfill
  \small
  \boxed{\delta \satisfies e \hastype \tau}
  \\
  \begin{mathpar}
    \inferrule { 
      e \rightsquigarrow e'
      \\
      \delta \satisfies e' \hastype \tau 
    } {
      \delta \satisfies e \hastype \tau 
    }

    \inferrule { 
      \J{<}l \J{>} v \in g
      \\
      \delta \satisfies v \hastype \tau 
      \\
      \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
    } {
      \delta \satisfies g \hastype \J{<} l \J{>} \tau
    }

    \inferrule { 
      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
    }

    \inferrule { 
      \forall \sigma \qua
      \delta \satisfies p[\sigma] \hastype \tau_l
      \implies
      \delta  \satisfies e[\sigma] \hastype \tau_r
      \\\\
      \forall \tau_l', \tau_r' \qua
      \delta \satisfies f \hastype \tau_l' \J{->} \tau_r'
      \implies
      \delta \satisfies \tau_l \subtypes \tau_l'
      \implies
      \delta \satisfies \tau_r' \subtypes \tau_r
    } {
      \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_l
      \\
      \neg\ (\delta \satisfies e \hastype \tau_r)
    } {
      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_l
      \\
      \delta \satisfies e \hastype \tau_r
    } {
      \delta \satisfies e \hastype \tau_l \J{\&} \tau_r
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_l
    } {
      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_r
    } {
      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
    }

    \inferrule { 
      \forall \delta' \qua
      \textbf{dom}(\delta') \subseteq \Theta \implies
      \delta \oplus \delta' \satisfies \Delta \implies 
      \delta \oplus \delta' \satisfies e \hastype \tau
      \\\\
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta \up 
      \delta \oplus \delta' \satisfies \Delta
    } {
      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
    }

    \inferrule { 
      \textbf{dom}(\delta') \subseteq \Theta
      \\
      \delta \oplus \delta' \satisfies \Delta 
      \\
      \delta \oplus \delta' \satisfies e \hastype \tau
    } {
      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
    }

    \inferrule { 
      \alpha \slash \tau \in \delta
      \\
      \delta \satisfies e \hastype \tau
    } {
      \delta \satisfies e \hastype \alpha 
    }

    \inferrule { 
    } {
      \delta \satisfies \J{@} \hastype \J{@}
    }

    \inferrule { 
      \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
    } {
      \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
    }

  \end{mathpar}
\end{definition}

\noindent
Model typing describes what expressions inhabit what types,
according to the forms of types, the forms of values, and how expressions
progress into values. 
It imbues types with semantics, which we use
to guide and judge the design of our type inference system. 
In general, a typing holds if its expression can make progress and 
the typing holds for the reduced form.

For inhabitation of a entry \ms{\delta \satisfies g \hastype \J{<}l\J{>}\tau},
the typing holds if the the expression is a record value, and
there is an entry in the record whose content inhabits     
the content of the entry type. Additionally, there must be just one
value per entry in the record. 
For inhabitation of a path 
\ms{\delta \satisfies f\ \J{[}p \J{=>} e \J{]} \hastype \tau_l\J{->}\tau_r},
the typing holds if the the expression is a function,
and there is one path in the function where 
the typing of the path body holds whenever the typing of the pattern holds.   
Additionally, since the paths of a function are tried in order, such that 
only the result of the first matched path is returned,
it is required that either the antecedent of the path type
is not weaker than the antecedent of preceding paths 
or the consequent of the path type is weaker than the consequent
of preceding paths.

For inhabitation of a difference 
\ms{\delta \satisfies e \hastype \tau_l \typdiff \tau_r},
the typing holds if the expression inhabits the left type
but not the right right.
For inhabitation of a intersection 
\ms{\delta \satisfies e \hastype \tau_l \J{\&} \tau_r},
the typing holds if the expression inhabits both the
left type and the right type. 
For inhabitation of a union 
\ms{\delta \satisfies e \hastype \tau_l \J{|} \tau_r},
the typing holds if the expression inhabits either the
left type or the right type.

For inhabitation of a universal 
\ms{\delta \satisfies e \hastype \J{ALL[} \Theta \J{]}\Delta\J{:}\tau},
the typing holds if the expression inhabits the constituent type,
under all interpretations of bound variables that satisfy the qualifiers. 
To avoid accepting an erroneous expression inhabiting a vacuous
type, it is required that the qualifiers are satisfiable.  
For inhabitation of an existential 
\ms{\delta \satisfies e \hastype \J{ALL[} \Theta \J{]}\Delta\J{:}\tau},
the typing holds if qualifiers and the inhabitation of the constituent type
are satisfiable by an interpretation of the bound variables.

For inhabitation of a variable \ms{\delta \satisfies e \hastype \alpha},
the typing holds if there's an interpretation for which the typing holds. 
For inhabitation of unit \ms{\delta \satisfies \J{@} \hastype \J{@}},
the typing holds if the expression is unit.
For inhabitation of a least fixed point \ms{\delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\alpha},
the typing holds if the expression also inhabits the inflated type.

\begin{definition}
  \label{def:model_subtyping}
  \emph{model subtyping}
  \hfill
  \small
  \boxed{\delta \satisfies \tau <: \tau}
  \\
  \begin{mathpar}
    \inferrule { 
      \forall e .\ 
      \delta \satisfies e \hastype \tau_l \implies 
      \delta \satisfies e \hastype \tau_r 
    } {
      \delta \satisfies \tau_l <: \tau_r
    } 
  \end{mathpar}
\end{definition}

\noindent
Model subtyping describes how the spaces of two types compare to each other.

\begin{definition}
  \label{def:model_subtyping_sequence}
  \emph{model subtyping sequence} 
  \hfill
  \small
  \boxed{\delta \satisfies \Delta}
  \\
  \begin{mathpar}
    \inferrule { 
    } {
      \delta \satisfies \epsilon 
    }

    \inferrule { 
      \delta \satisfies \Delta 
      \\
      \delta \satisfies \tau_l \subtypes \tau_r
    } {
      \delta \satisfies \Delta \  \tau_l \J{<:} \tau_r
    } 
  \end{mathpar}
\end{definition}

\noindent
Model subtyping sequence describes multiple subtyping constraints.

\begin{definition}
  \label{def:model_typing_sequence}
  \emph{model typing sequence} 
  \hfill
  \small
  \boxed{\delta, \sigma \satisfies \Gamma}
  \\
  \begin{mathpar}
    \inferrule {
      \forall x \qua
      x \hastype \tau \in \Gamma
      \implies
      \delta \satisfies x[\sigma] \hastype \tau
    } {
      \delta, \sigma \satisfies \Gamma
    }
  \end{mathpar}
\end{definition}

\noindent
Model typing sequence describes multiples typings of variables.

\section{Static System}
\label{sec:static_system}

The static system consists of ways to characterize abstract expressions 
and abstract types without relying on progression to concretize and reduce expressions. 
The two main components are proof typing and proof subtyping.
Proof typing is a strengthening of model typing, and 
proof subtyping is a strengthening of model subtyping.
When viewed as an algorithm,
they form the core of the type inference system. 
Proof typing generates subtyping constraints approximating the control flow of an expression, 
and then leverages proof subtyping to solve for the variables in the the subtyping constraints,
where the variables represent possible types of expressions.

\begin{definition} 
  \label{def:proof_typing}
  \emph{proof typing}
  \hfill 
  \small
  \boxed{\Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
    } {
      \Theta, \Delta, \Gamma \entails \J{@} \hastype \J{@} \given \Theta, \Delta
    }

    \inferrule {
      x:\tau \in \Gamma
    } {
      \Theta, \Delta, \Gamma \entails x \hastype \tau \given \Theta, \Delta 
    }

    \inferrule {
    } {
      \Theta, \Delta, \Gamma \entails \epsilon \hastype \J{TOP} \given \Theta, \Delta
    }

    \inferrule {
      \Theta, \Delta, \Gamma \entails r \hastype \tau \given \Theta', \Delta'  
      \\\\
      \Theta', \Delta', \Gamma \entails e \hastype \tau' \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta, \Gamma \entails r\ \J{<}l \J{>} e \hastype (\tau\ \J{\&}\ \J{<} l \J{>} \tau') 
      \given \Theta'', \Delta'' 
    }

    \inferrule {
      \exists \Xi \qua \Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi 
      \\\\
      \textbf{pack}^+(\Theta \cup \textbf{ftv}(\Delta) \cup \textbf{ftv}(\Gamma) \entails \Pi) = \tau 
    } {
      \Theta, \Delta, \Gamma \entails f \hastype \tau \given \Theta, \Delta
    }

    \inferrule {
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta' 
      \\\\
      \Theta', \Delta' \entails 
      \tau \subtypes  \J{<} l \J{>} \alpha 
      \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta, \Gamma \entails e\J{.}l \hastype \alpha \given \Theta'', \Delta''
    }

    \inferrule {
      \Theta, \Delta, \Gamma \entails e_f \hastype \tau_f \given \Theta', \Delta'
      \\
      \Theta', \Delta', \Gamma \entails e_a \hastype \tau_a \given \Theta'', \Delta''
      \\\\
      \Theta'', \Delta'', \tau_f \subtypes \tau_a\J{->}\alpha \given \Theta''', \Delta'''
    } {
      \Theta, \Delta, \Gamma \entails e_f\J{(}e_a \J{)} \hastype \alpha \given \Theta''', \Delta'''
    }

    \inferrule {
      \textbf{ftv}(\tau_a) \subseteq \emptyset
      \\
      \Theta, \Delta, \Gamma \entails e \hastype \Pi
      \\
      \textbf{pack}^+(\Theta \cup \textbf{ftv}(\Delta) \cup \textbf{ftv}(\Gamma) \entails \Pi) = \tau 
      \\\\
      \exists \Theta^\dagger, \Delta^\dagger \qua \Theta', \Delta' \entails \tau \subtypes \tau_a \given \Theta^\dagger, \Delta^\dagger
      \\
      \Theta', \Delta', \Gamma, x:\tau_a \entails e' \hastype \tau' \given \Theta'', \Delta''
    } {
      \Theta, \Delta, \Gamma \entails \J{def } x \J{:} \tau_a \J{ = } e \J{ in } e' \hastype \tau' \given \Theta'', \Delta''
    }

    \inferrule {
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau \subtypes \alpha_\nu \rightarrow \Pi_\nu
      \\\\
      \alpha_\nu \downarrow \Pi_\nu
      \fallingdotseq 
      \alpha_\mu \uparrow \Pi_\mu
      \\
      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha_\mu \entails \Pi_\mu) = \tau_\mu
      \\\\
      \factorsinto \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{left} \mapsto \tau_l
      \\
      \factorsinto \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{right} \mapsto \tau_r
    } {
      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
      \hastype 
      \tau_l \J{->} \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

\noindent
For a given context and expression, 
proof typing attempts to infer a type and a subtyping sequence,
such that if the context holds,
then there is a solution to the generated subtyping sequence, under which 
the expression inhabits the inferred type.
The context consists of skolems \ms{\Theta}, subtyping constraints \ms{\Delta}, and typings \ms{\Gamma}. 
The contextual subtyping constraints and typings constrain the search space of valid typings, 
such that the generated subtyping constraints are consistent with them. 
In fact, the generated subtyping constraints contain
all the contextual subtyping constraints, as it keeps the definition a bit less cluttered.
Skolems are variables that further constrain the search space. 
The generated subtyping constraints must be satisfiable under every interpretation of
skolems for which the contextual constraints are satisfiable.  
Proof typing may also generate additional skolems, which can be used to constrain
subsequent sub-problems.
If a variable is not in the set of skolems, then we consider it to be plastic,
in that it can be constrained in any way that is consistent with the other subtyping constraints
and the typing.

\noindent
For unit \code{@}, it infers the unit type \J{@}. 
For a variable, \ms{x}, it looks for a corresponding typing in the context, 
and infers the type it finds.
For an empty sequence \ms{\epsilon}, 
it infers top \code{TOP},
thereby providing a basis for inferring types of records. 
For a record \ms{r\ \J{<}l\J{>}e},  
it recursively constructs the type for each entry, and
refines their types against each other by compounding intersection.

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of schemas \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of negatable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each schema into a type,
such that the skolems and plastics are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

For projection \ms{e\J{.}l}, 
it delegates to proof subtyping 
\ms{... \entails \tau \subtypes \J{<}l\J{>}\alpha \given ...}
to check that there is an entry with the label \ms{l}
in the source expression's type \ms{\tau}.
With subtyping, it learns a lower bound on the type of the entry's content \ms{\alpha}.
For application \ms{e_f\J{(}e_a\J{)}}, 
it delates to proof subtyping 
\ms{... \entails \tau_f \subtypes \tau_a \J{->} \alpha \given ...}
to check that that the function \ms{e_f} can actually
map the argument \ms{e_a} to a result. With subtyping, 
it learns a lower bound on the type of the result \ms{\alpha}.
For an annotation \ms{\J{def}\ x \J{:} \tau_a\ \J{=}\ e\ \J{in}\ e'}, 
it first delegates to a few helpers. 
First it leverages typing consolidation \ms{\Theta, \Delta, \Gamma \entails e \hastype \Pi}
to infer as many possible schemas for the source expression \ms{e}.
Then it packs all the schemas together and ensures that the packed type
subtypes the type annotation.
It assigns the annotated type to the defined variable in the contextual typing
as it proceeds to infer the type of the continuation \ms{e'}, thereby
artificially restricting the permissible control flow.

For a loop \ms{\J{loop(}e\J{)}}, 
it eagerly factors the paths of the recursive function 
to construct a single path type 
with an recursive type as its antecedent and an recursive type as its consequent. 
First, it infers the type of the fixed point combinators argument \ms{e},
and it leverages loop subtyping consolidation 
\ms{\Theta', \Delta' \entails \tau \subtypes \alpha_\nu \rightarrow \Pi^\nu}
to ensure the argument has a  
form compatible with looping and 
to collect as many type schemas of its paths as possible.  
It leverage schema sequence duality 
\ms{
  \alpha_\nu \downarrow \Pi_\nu
  \fallingdotseq 
  \alpha_\mu \uparrow \Pi_\mu
}, which rewrites the implicitly co-recursive path schemas into
implicitly recursive pair schemas. 
Then, it leverages negative packing \ms{\textbf{pack}^-(... \entails \Pi_\mu) = \tau_\mu},
and constructs an explicitly recursive type by wrapping it in a least fixed point
\ms{\J{LFP[}\alpha_\mu\J{]}\tau_\mu}.
It then leverages factorization \ms{
  \Vdash \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{<left>}\ \tau_l
} to factor out the left part of the recursive pair, and does the same for the right part. 
Finally from the left and right parts, it forms a single path type
with a least fixed point for both the antecedent and the consequent.

\begin{definition} 
  \label{def:proof_subtyping}
  \emph{proof subtyping}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'}
\end{definition}

\noindent
For a given context and two abstract types,
proof subtyping attempts to solve the subtyping between the two types,
by generating a subtyping sequence, such that if the context holds,
then there is a solution to the generated subtyping sequence, under which   
the initial subtyping also holds.

The context consists of skolems \ms{\Theta} and subtyping constraints \ms{\Delta}. 
The contextual subtyping constraints constrain the search space of valid subtyping constraints, 
such that the generated subtyping constraints are consistent with them. 
In fact, the generated subtyping constraints contain all the contextual subtyping constraints.
As with proof typing, skolems constrain the search space,
such that the generated subtyping constraints must be satisfiable
for every interpretation of skolems for which the contextual subtyping constraints are satisfiable.

In order to find a solution, the solver considers the forms 
of both types in the subtyping relation.
There are elimination rules, which reason according to the form of the subtype,
and there are introduction rules, which reason according to the form of the supertype.
Some rules handle both elimination and introduction of type forms.
Due to the numerous forms of types where each form is considered for each of the two types,
proof subtyping relies on many rules.  
For clarity, we subdivide into the rules into smaller groups according to some themes.

\begin{definition} 
  \label{def:proof_subtyping_reflection}
  \emph{proof subtyping (reflection)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
    } {
      \Theta, \Delta \entails \tau \subtypes \tau \given \Theta, \Delta 
    }
  \end{mathpar}
\end{definition}

\noindent
Proof subtyping is reflexive. A type
is considered both weaker and stronger than itself. 
We implicitly assume alpha conversion for variables
bound in existentials and universals.
Our implementation of proof typing converts to a nameless representation,
based on de Bruijn indexes \cite{} before checking equality.

\begin{definition} 
  \label{def:proof_subtyping_implication_preservation}
  \emph{proof subtyping (implication preservation)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \kappa_l \subtypes \kappa_r \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r
      \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{<}l \J{>} \tau_l \subtypes \J{<} l \J{>} \tau_r
      \given \Theta', \Delta'
    }

    \inferrule {
      \Theta, \Delta \entails \tau_{x} \subtypes \tau_{p} \given \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau_{q} \subtypes \tau_{y} \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails 
      \tau_{p} \J{->} \tau_{q} 
      \subtypes 
      \tau_{x} \J{->} \tau_{y}
      \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}

\noindent
Both entry types and path types are analogous to propositional implication. 
However, unlike the typical separate rules for introduction and elimination in propositional implication,
the subtyping solver combines the semantics for introduction
and elimination into rules that preserve the implication structures.

For preservation of entries 
\ms{\J{<}l \J{>} \tau_l \subtypes \J{<} l \J{>} \tau_r},
it checks that the entries' labels match and that 
subtyping holds for their contents.
For preservation of paths 
\ms{\tau_p \J{->} \tau_q \subtypes \tau_x \J{->} \tau_y},
it checks that the antecedent of the subtype
is weaker than the antecedent of the supertype, 
and the consequent of the subtype 
is stronger than the consequent of the supertype.

\begin{definition} 
  \label{def:proof_subtyping_expansion_elimination}
  \emph{proof subtyping (expansion elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \phi \subtypes \tau \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails
      \tau_a \subtypes \tau
      \given \Theta', \Delta'
      \\
      \Theta', \Delta' \entails
      \tau_b \subtypes \tau
      \given \Theta'', \Delta''
    } {
      \Theta, \Delta \entails
      \tau_a\J{|}\tau_b \subtypes \tau
      \given \Theta'', \Delta'' 
    }

    \inferrule {
      \Theta, \Delta \Vvdash \Omega
      \\
      \Theta, \Delta \entails \Omega \given \Theta', \Delta'
      \\
      \Theta' \cup \Theta_l, \Delta' \entails
      \tau_l \subtypes \tau_r \given \Theta'', \Delta''
    } {
      \Theta, \Delta \entails
      \J{EXI[}\Theta_l\J{]}\Omega \J{:} \tau_l \subtypes \tau_r
      \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}

\noindent
For union elimination \ms{
\tau_a\J{|}\tau_b \subtypes \tau
},
it solves elimination for both of union's constituents.
For existential elimination \ms{ 
\J{EXI[}\Theta_l \J{]} \Omega \J{:}\tau_l \subtypes \tau_r
}, 
it ensures that the existential's qualifiers
are negatable \ms{\Theta,\Delta \Vvdash \Omega}, 
and then it finds a solution for the qualifiers.
It then solves the elimination of the existential's constituent type,
constrained by the skolemized bound variables,
and the solution to the qualifiers. 
The negatable check is necessary to ensure
that a solution the the qualifiers implies
a solution to the subtyping constraints generated by solving the qualifiers.
Thus, if the solver can eliminate the existential, 
then a solution to the qualifiers implies a solution to the constituent's elimination.

\begin{definition} 
  \label{def:proof_subtyping_refinement_introduction}
  \emph{proof subtyping (refinement introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \psi \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails
      \tau \subtypes \tau_{a} \given \Theta', \Delta' 
      \\
      \Theta', \Delta' \entails
      \tau \subtypes \tau_{b} \given \Theta'', \Delta''
    } {
      \Theta, \Delta \entails
      \tau \subtypes \tau_{a}\J{\&}\tau_{b} \given \Theta'', \Delta''
    }

    \inferrule {
      \Theta, \Delta \Vvdash \Omega
      \\
      \Theta, \Delta \given \Omega \given \Theta', \Delta'
      \\
      \Theta' \cup \Theta_r, \Delta' \entails \tau_l \subtypes \tau_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails \tau_l \subtypes \J{ALL[}\Theta_r\J{]}\Omega \J{:} \tau_r \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

\noindent
Refinement introduction is the dual of expansion elimination, so its rules
rely on analogous techniques.

\begin{definition} 
  \label{def:proof_subtyping_plastic_elimination}
  \emph{proof subtyping (plastic elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'} 
  \\
  \begin{mathpar}
    \inferrule {
      \alpha \notin \Theta
      \\
      (\forall \tau' \qua \tau' \J{<:} \alpha \in \Delta \implies \tau' \J{<:} \tau \in \Delta^\dagger)
      \\
      \Theta, \Delta \entails \Delta^\dagger \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'\ \alpha\J{<:}\tau
    }
    % \inferrule {
    %   \alpha \notin \Theta
    %   \\
    %   \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{+} \backslash \Delta \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'\ \alpha\J{<:}\tau
    % }
  \end{mathpar}
\end{definition}

\noindent
For plastic elimination \ms{\alpha \subtypes \tau} where \ms{\alpha \notin \Theta}, 
it ensures that the supertype is consistent with all lower bounds of the plastic,
solutions against those lower bounds. Then it learns that the supertype is an upper bound 
of the plastic.

\begin{definition} 
  \label{def:proof_subtyping_plastic_introduction}
  \emph{proof subtyping (plastic introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \alpha \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \alpha \notin \Theta
      \\
      (\forall \tau' \qua \alpha \J{<:} \tau' \in \Delta \implies \tau \J{<:} \tau' \in \Delta^\dagger) 
      \\
      \Theta, \Delta \entails \Delta^\dagger \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha \given \Theta', \Delta'\ \tau\J{<:}\alpha
    }
    % \inferrule {
    %   \alpha \notin \Theta
    %   \\
    %   \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{-} \backslash \Delta \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails 
    %   \tau \subtypes \alpha \given \Theta', \Delta'\ \tau\J{<:}\alpha
    % }
  \end{mathpar}
\end{definition}

\noindent
Plastic introduction is the dual of plastic elimination, so its rule
relies on analogous techniques.

\begin{definition} 
  \label{def:proof_subtyping_skolem_introduction}
  \emph{proof subtyping (skolem introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \alpha \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \alpha \in \Theta 
      \\
      \exists \alpha' \qua \alpha' \J{<:} \alpha \in \Delta \up \alpha' \notin \Theta 
      % \\\\
      % \exists \Theta', \Delta' \qua \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{-} \backslash \Delta \given \Theta', \Delta'
      \\\\
      (\forall \tau' \qua \alpha \J{<:} \tau' \in \Delta \implies \tau \J{<:} \tau' \in \Delta^\dagger)
      \\
      \Theta, \Delta \entails 
      \Delta^\dagger 
      \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha 
      \given \Theta', \Delta'\ \tau \J{<:} \alpha 
    }

    \inferrule {
      \alpha \in \Theta 
      \\
      \tau' \J{<:} \alpha \in \Delta
      \\\\
      (\forall \alpha' \qua \tau' = \alpha' \implies \alpha' \in \Theta) 
      \\
      \Theta, \Delta \entails \tau \subtypes \tau' \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha 
      \given \Theta', \Delta' 
    }
  \end{mathpar}
\end{definition}

\noindent
For skolem introduction \ms{\tau \subtypes \alpha} where \ms{\alpha \in \Theta},
there are two possibilities.
In the first scenario, the skolem has a plastic as a lower bound.
If effect, this enables learning lower bounds on the skolem,
so the solver uses the same techniques as it did for plastic elimination.
I the second scenario, there is a lower bound that's not a plastic,
so the solver ensures that the skolem introduction is weaker
than the constraints by checking that the subtype subtypes the lower bound that
was found. In practice, the solver avoids checking the subtype against the lower bound, 
if the lower bound is a plastic, as that would cause the solver to diverge due to
a mutual dependence between plastic elimination and skolem elimination.

\begin{definition} 
  \label{def:proof_subtyping_skolem_elimination}
  \emph{proof subtyping (skolem elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'} 
  \\
  \begin{mathpar}
    \inferrule {
      \alpha \in \Theta 
      \\
      \exists \alpha' \qua \alpha \J{<:} \alpha' \in \Delta \up \alpha' \notin \Theta 
      % \\\\
      % \exists \Theta', \Delta' \qua \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{+} \backslash \Delta \given \Theta', \Delta'
      \\\\
      (\forall \tau' \qua \tau' \J{<:} \alpha \in \Delta \implies \tau' \J{<:} \tau \in \Delta^\dagger)
      \\
      \Theta, \Delta \entails 
      \Delta^\dagger 
      \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \alpha \subtypes \tau
      \given \Theta', \Delta'\ \alpha \J{<:} \tau
    }

    \inferrule {
      \alpha \in \Theta 
      \\
      \alpha \J{<:} \tau' \in \Delta
      \\\\
      (\forall \alpha' \qua \tau' = \alpha' \implies \alpha' \in \Theta) 
      \\
      \Theta, \Delta \entails \tau' \subtypes \tau \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \alpha \subtypes \tau
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

\noindent
Skolem elimination is the dual of skolem introduction, so its rules
rely on analogous techniques.

\begin{definition} 
  \label{def:proof_subtyping_implication_rewriting}
  \emph{proof subtyping (implication rewriting)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \tau_l \subtypes
      (\tau_a\J{->}\tau_r)\J{\&}(\tau_b\J{->}\tau_r)
      \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes 
      (\tau_a\J{|}\tau_b)\J{->}\tau_r
      \given \Theta', \Delta' 
    }

    \inferrule {
      \Theta, \Delta \entails 
      \tau_l
      \subtypes
      (\tau_r\J{->}\tau_a)
      \J{\&}
      (\tau_r\J{->}\tau_b)
      \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes 
      \tau_r\J{->}(\tau_{a}\J{\&}\tau_{b})
      \given \Theta', \Delta'
    }

    \inferrule {
      \Theta, \Delta \entails 
      \tau
      \subtypes
      (\J{<} l \J{>} \tau_a)
      \J{\&}
      (\J{<} l \J{>} \tau_b)
      \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau \subtypes 
      \J{<} l \J{>} (\tau_{a}\J{\&}\tau_{b})
      \given \Theta', \Delta'
    }

  \end{mathpar}
\end{definition}

\noindent
Just as entries and paths are analogous to propositional implication,
unions and intersections are analogous to disjunction and conjunction, respectively.
Just as disjunction elimination or propositional logic relies on two cases of implication,
the solver recognizes that union in the antecedent of a path,
can be solved by rewriting into two cases of paths.  
Additionally, since must preserve implications, there is no direct no single
rule that corresponds modus ponens. Thus, the solver 
recognizes that intersection in the consequent of a path or (the content of an entry),
may need to be rewritten as two paths (or two entries).

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_r' 
      \\
      \Theta, \Delta \entails \tau_r' \subtypes \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

    \inferrule {
      \alpha \notin \textbf{ftv}(\tau_l)
      \\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r 
      \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r 
      \given \Theta', \Delta' 
    }

    \inferrule {
      \tau_l[\alpha \slash \tau_r]^+ = \tau^\dagger
      \\
      \Theta, \Delta \entails 
      \tau^\dagger \subtypes \tau_r \given 
      \Theta', \Delta'
      \\
      \alpha \not\in \textbf{ftv}(\tau^\dagger)
      \\
      \alpha \not\in \textbf{ftv}(\tau_r)
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r 
      \given \Theta', \Delta' 
    }
  \end{mathpar}
\end{definition}

\noindent
For least fixed point elimination \ms{
\J{LFP[} \alpha \J{]} \tau_l \subtypes \tau_r
}, there are a few scenarios
that may work. 
In the first scenario, the solver recognizes that the supertype
is an entry, so it factors out the corresponding column from
the least fixed point.
In the second scenario, the solver recognizes that 
the constituent type doesn't contain the self reference,
so it's sufficient to simply solve the elimination of the constituent.  
In the third case, the solver performs mathematical induction.
It substitutes the super type \ms{\tau_r}
into all positive occurrences of the self reference \ms{\alpha} within
the least fixed point's constituent type \ms{\tau_l}.
It then solves the elimination of this substitution into the super type.
In effect, solving the least fixed point elimination under the positive substitution is 
akin to deriving the least fixed point elimination under the assumption of the induction hypothesis,
\ms{\alpha \subtypes \tau_r}.

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \Vvdash \eta
      \\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \tau \subtypes \eta \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \typdiff \eta
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

\noindent
For difference introduction \ms{\tau \subtypes \tau_r \typdiff \eta}, 
it checks that subtyping holds for the positive part 
and fails for the subtracted part.
Additionally, it requires that the subtracted part is negatable
\ms{\Theta, \Delta \Vvdash \eta}, 
which ensures that its introduction is complete,
thus failure of its introduction is sound.

\begin{definition} 
  \label{def:proof_subtyping_lfp_intro}
  \emph{proof subtyping (least fixed point introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \circlearrowleft \J{LFP[} \alpha \J{]} \tau_r
      \\
      \Theta, \Delta \entails
      \tau_l \subtypes \tau_r[\alpha \slash \J{LFP[} \alpha \J{]} \tau_r]
      \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails
      \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'
    }

    \inferrule {
      \alpha \Vdash \tau_r \supertypes \tau_r'
      \\
      \Theta, \Delta \entails
      \tau_l \subtypes \tau_r' \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails
      \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'
    }

  \end{mathpar}
\end{definition}

\noindent
For least fixed point introduction \ms{
\tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r
}, the solver has a couple options.
One option is to check for a syntactic guarantee that the least fixed point is inflatable without diverging 
\ms{
\tau_l \circlearrowleft \J{LFP[} \alpha \J{]} \tau_r
} and then solve the inflated introduction.
Alternatively, it can strengthen the least fixed point
with union drop \ms{\alpha \Vdash \tau_r \supertypes \tau_r'},
which drops any parts of the union containing the self reference.

\begin{definition} 
  \label{def:proof_subtyping_diff_elimination}
  \emph{proof subtyping (difference elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \tau_l \subtypes \eta \J{|} \tau_r \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

\noindent
For difference elimination \ms{ 
\rho \typdiff \eta \subtypes \tau
},
it simply rewrites the problem in terms of union.

\begin{definition} 
  \label{def:proof_subtyping_expansion_introduction}
  \emph{proof subtyping (expansion introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \phi \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau \subtypes \tau_{l} \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails \tau \subtypes \tau_{l}\J{|}\tau_{r} \given \Theta', \Delta' 
    }

    \inferrule {
      \Theta, \Delta \entails \tau \subtypes \tau_{r} \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails \tau \subtypes \tau_{l}\J{|}\tau_{r} \given \Theta', \Delta' 
    }

    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
      \\\\
      \Theta', \Delta' \entails \Delta_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails 
      \tau_l
      \subtypes 
      \J{EXI[}\Theta_r\J{]}\Delta_r \J{:} \tau_r \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}
\noindent
For union introduction \ms{
\tau \subtypes \tau_{l}\J{|}\tau_{r}
}, 
it can choose from two possible options:
solving the introduction of of left part, 
or solving the introduction of the right part.
For existential introduction \ms{ 
\tau_l \subtypes \J{EXI[}\Theta_r \J{]} \Delta_r \J{:}\tau_r
},
it constructs witnesses by solving the introduction
of the existential's constituent type \ms{\entails \tau_l \subtypes \tau_r \given}, 
from which the witnesses can be inferred from the generated subtyping constraints.
It then attempts to solve the qualifiers by leveraging the witnesses.

\begin{definition} 
  \label{def:proof_subtyping_refinement_elimination}
  \emph{proof subtyping (refinement elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \psi \subtypes \tau \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_{l} \subtypes \tau \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau \given \Theta', \Delta' 
    }

    \inferrule {
      \Theta, \Delta \entails \tau_{l} \subtypes \tau \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau \given \Theta', \Delta'
    }

    \inferrule {
      \Vdash \tau_{l}\J{\&}\tau_{r} \subtypes \tau_x \rightarrow \tau_y 
      \\\\
      \Theta, \Delta \entails \tau_x \J{->} \tau_y  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    }

    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \Delta_l \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails 
      \J{ALL[}\Theta_l\J{]} \Delta_l \J{:} \tau_l
      \subtypes 
      \tau_r
      \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}

\noindent
Refinement elimination is the dual of expansion introduction, so its rules
rely on analogous techniques.
However, there is one additional rule
for eliminating an intersection and introducing an implication \ms{
\tau_l \& \tau_r \subtypes \tau_a \J{->} \tau_b
}.
As we mentioned earlier when discussing packing together the inferred types of 
multiple paths of functions, sometimes the argument type is weaker than each
path type antecedent, yet could be stronger than union of all the path type antecedents.    
As such, the solver may leverage function type factorization \ms{
\Vdash \tau_l \J{\&} \tau_r \subtypes \tau_x \rightarrow \tau_y 
} to weaken a function type
into a single path type where the antecedent and the consequent may consist
of compounded unions.

\begin{definition} 
  \label{def:proof_subtyping_sequence}
  \emph{proof subtyping sequence} 
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \Delta_q \given \Theta', \Delta'}
  \\
  \begin{mathpar}
    \inferrule {
    } {
      \Theta, \Delta \entails \epsilon \given \Theta', \Delta'
    }

    \inferrule {
      \Theta, \Delta \entails \tau_l \J{<:} \tau_r \given \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau_l \J{<:} \tau_r \given \Theta'', \Delta''
    } {
      \Theta, \Delta \entails \Delta_q\ \tau_l \J{<:} \tau_r \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

\noindent
Proof subtyping sequence very simply solves a sequence of subtyping goals
and compounds their results, such that the solution from one constrains the next.

\section{Theory}
\label{sec:theory}

\begin{theorem} 
  \label{thm:proof_subtyping_soundness}
  \emph{proof subtyping soundness} 
  \small
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
    } {
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta' \backslash \Theta
      \up
      (
      \forall \delta \qua 
      \delta \oplus \delta' \satisfies \Delta' \implies
      \delta \oplus \delta' \satisfies \tau_l \subtypes \tau_r
      )
    }
  \end{mathpar}
\end{theorem}

\noindent
Proof subtyping soundness states that if proof subtyping generates subtyping constraints \ms{\Delta'} 
for a subtyping goal \ms{\tau_l \subtypes \tau_r},
then any solution for the generated subtyping constraints
is a solution for the subtyping goal, with the caveat that
it only needs to hold for some interpretation of the generated skolems \ms{\Theta'\backslash\Theta}. 
The proof is by induction, leveraging numerous lemmas to derive the semantics
defined by model typing in each case. 
Completeness of restricted forms is necessary whenever the corresponding model semantics 
is used in a negative position, such as in negation or in the antecedent of an implication.     
Additionally, when rewriting an object, it is necessary to characterize how the model
semantics of its input form relates to the model semantics of its output form.
Lastly, when pruning subtyping constraints generated by the solver, it is necessary
that the remaining subtyping are as strong or stronger that the removed subtyping constraints
with respect to the variables used in the goal.
The property of proof subtyping soundness is a bit weak on its own. 
It does not guarantee that there is a solution for the subtyping goal,
it could be met vacuously by having proof subtyping 
generate an unsatisfiable subtyping constraints.
To make the stronger guarantee, we can combine it with another theorem.

\begin{theorem}
  \label{thm:proof_subtyping_substance}
  \emph{proof subtyping substance}
  \small
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
    } {
      \forall \delta \qua \textbf{dom}(\delta) \subseteq \Theta \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta
      \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta'
    }
  \end{mathpar}
\end{theorem}

\noindent
Proof subtyping substance states that generate constraints are satisfiable
assuming the contextual constraints are satisfiable, with respect
to all interpretations of the contextual skolems.   
The proof is by induction, 
where in most cases the property follows directly from the the induction hypothesis.
The only cases that are actually interesting are the variable rules
where new constraints are generated.
Because the rules check for consistency by searching for constraints on single variables,
it is necessary that the constraints on a single variable are as strong or stronger
than all the constraints with respect to that variable.
It then boils down to checking the bound in the goal against all the bounds on the other side
of the variable, as those are the only opportunities for contradiction.

\begin{theorem}
  \label{thm:proof_typing_soundness}
  \emph{proof typing soundness} 
  \small
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta'
    } {
      \exists \delta' \qua \textbf{dom}(\delta') \subseteq \Theta' \backslash \Theta \up 
      (
      \forall \delta \qua
      \delta \oplus \delta' \satisfies \Delta' \implies
      \delta, \sigma \satisfies \Gamma \implies
      \delta \oplus \delta' \satisfies e[\sigma] \hastype \tau
      )
    }
  \end{mathpar}

\end{theorem}

\noindent
Proof typing soundness states that if proof typing generates subtyping constraints \ms{\Delta'} 
for a typing goal \ms{e \subtypes \tau},
then any solution for the contextual typing constraints 
and the generated subtyping constraints
is a solution for the typing goal, with the caveat that
it only needs to hold for some interpretation of the generated skolems \ms{\Theta'\backslash\Theta}. 
We prove soundness by induction, leveraging subtyping soundness,
and additional lemmas the characterize the relationships between rewritten forms of objects. 
Proof typing soundness as the same limitations as proof subtyping soundness. 
It's not strong enough to guarantee our goal of type inference,
but we can easily compose it with another property to do so.

\begin{theorem}
  \label{thm:proof_typing_substance}
  \emph{proof typing substance} 
  \small
  \\
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta'
    } {
      \forall \delta \qua \textbf{dom}(\delta) \subseteq \Theta \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta 
      \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta'
    }
  \end{mathpar}
\end{theorem}

\noindent
Proof typing substance states that the generated subtyping constraints are satisfiable
assuming the contextual constraints are satisfiable, with respect
to all interpretations of the contextual skolems.   
The proof is by induction, and it follows easily from the induction hypothesis
and proof subtyping soundness.

% \begin{theorem}
%   \label{prop:proof_typing_annotation_soundness}
%   \emph{proof typing annotation soudness}
%   \\
%   \begin{mathpar}
%     \inferrule {
%       \Theta, \Delta, \Gamma \entails \J{def } x \J{:} \tau_a \J{ = } e \J{ in } e' \hastype \tau' \given \Theta', \Delta'
%     } {
%       \forall \delta, \sigma \qua 
%       \delta, \sigma \satisfies \Gamma \implies
%       \delta \satisfies e \hastype \tau_a
%     }
%   \end{mathpar}
% \end{theorem}

% \noindent
% In addition to inferring type according to the control flow of a program,
% we are also interested in checking that programs meet specifications 
% annotated by the programmer. Proof typing soundness 
% does not directly reference these annotations, 
% so we need an additional theorem to ensure
% the annotations are respected.

\section{Experiments}
\label{sec:experiments}

\noindent
We performed experiments to see how type inference performs on both functions and constructors. 
These two categories have historically been tackled independently of each other, where parametric types
tend to rely on notions of instantiation, and structural types rely on notions of equality or subtyping.
Our system unifies these historically distinct problems by describing both kinds of types in terms of subtyping.
However, the distinction is useful for comparing our design to previous work.

We implemented an extrinsic type inference system that corresponds to the predicate definitions
presented thus far. 
To observe the expressiveness of our system on functions, 
we evaluated our implementation on the same suite of programs 
that SuperF was evaluated on, as presented in the paper on SuperF \cite{}.
Based on these experiments, our system all programs admitted by SuperF, as well as some programs 
that SuperF is unable to admit.
Thus, our system appears to be more expressive on functions than the previous most expressive
type inference system that we know of.
To observe the expressiveness of our system on constructors, 
we constructed simple programs that represent the kind of flexible and reusable code
often found programs written in untyped scripting languages. 
We ran these examples in our implementation, and created analogous version to run
in the online implementation of MLScript \cite{},
which represents the theoretical system MLStruct \cite{}. 
All of these examples are admitted by our system, while MLStruct
can either requires intrinsic specifications to admit the program, 
or simply can't admit the program.

\begin{definition} Experimental Expressions and Types  
  \label{def:experimental_expressions_and_types}
  \scriptsize
  \[
  \begin{array}[t]{ll}
    \begin{array}[t]{r@{}c@{}l}
      \textbf{Bool} 
      &\ \ =\ \ & 
      \J{(<true>@ | <false>@)} 
      \\
      \textbf{Nat} 
      &\ \ =\ \ & 
      \J{(LFP[R] <zero>@ | <succ>R)} 
      \\
      \textbf{Even} 
      &\ \ =\ \ & 
      \J{(LFP[R] <zero>@ | <succ><succ>R)} 
      \\
      \textbf{List}(\alpha) 
      &\ \ =\ \ & 
      \J{(LFP[R] <nil>@ | <cons>((}\alpha\J{)*R))} 
      \\
      \textbf{ChurchNum} 
      &\ \ =\ \ & 
      \J{(ALL[T] (T -> T) -> T -> T)}
    \end{array}
    &
    \begin{array}[t]{r@{}c@{}l}
      \textbf{ST}(\alpha^\dagger, \alpha) 
      &\ \ =\ \ & 
      \J{(<tag>(}\alpha^\dagger \J{) <result>(} \alpha \J{))}
      \\
      \textbf{z} 
      &\ \ =\ \ & 
      \J{[f => [x => x]]} 
      \\
      \textbf{s} 
      &\ \ =\ \ & 
      \J{[n => [f => [x => f(n(f(x)))]]]} 
      \\
      \textbf{n3} 
      &\ \ =\ \ & 
      \textbf{s}\J{(}\textbf{s}\J{(}\textbf{s}\J{(}\textbf{z}\J{)}\J{)}\J{)} 
    \end{array}
  \end{array}
  \]
\end{definition}

\noindent
For clarity, we define a few metalinguistic terms to stand in for 
types and expressions used as inputs to our system.

\begin{definition} Experimental Typing Context  
  \label{def:experimental_typing_context}
  \tiny
  \[
  \begin{array}[t]{ll}
    \begin{array}[t]{r@{}c@{}l}
      \textbf{ctx} & = &  \hfill
      \\
      \J{head} &\J{ : }& \J{ALL[T] }\textbf{List}(\J{T})\J{ -> T}
      \\
      \J{tail} &\J{ : }& \J{ALL[T] }\textbf{List}(\J{T})\J{ -> }\textbf{List}(\J{T})
      \\
      \J{nil} &\J{ : }& \J{ALL[T] }\textbf{List}(\J{T})
      \\
      \J{single} &\J{ : }& \J{ALL[T] T -> }\textbf{List}(\J{T})
      \\
      \J{append} &\J{ : }& \J{ALL[T] }\textbf{List}(\J{T}) \J{ -> } \textbf{List}(\J{T}) \J{ -> } \textbf{List}(\J{T})
      \\
      \J{length} &\J{ : }& \J{ALL[T] }\textbf{List}(\J{T}) \J{ -> } \textbf{Nat}
      \\
      \J{const} &\J{ : }& \J{ALL[T U] T -> U -> T}
      \\
      \J{id} &\J{ : }& \J{ALL[T] T -> T}
      \\
      \J{ids} &\J{ : }& \textbf{List}(\J{ALL[T] T -> T}\J{)}
      \\
      \J{inc} &\J{ : }& \textbf{Nat}\J{ -> }\textbf{Nat}
      \\
      \J{choose} &\J{ : }& \J{ALL[T] T -> T -> T}
      \\
      \J{poly} &\J{ : }& \J{(ALL[T] T -> T) -> } \textbf{Nat}\J{*}\textbf{Bool}
      \\
      \J{fst} &\J{ : }& \J{ALL[T U] T*U -> T}
      \\
      \J{sort} &\J{ : }& \J{ALL[T] (T * T -> } \textbf{Bool} \J{) -> }\textbf{List}(\J{T})\J{ -> }\textbf{List}(\J{T})
      \\
      \J{scalarCmp} &\J{ : }& \textbf{Nat} \J{\*} \textbf{Nat} \J{ -> } \textbf{Bool} 
      \\
      \J{lexicoCmp} &\J{ : }& \textbf{List}(\textbf{Nat}) \J{\*} \textbf{List}(\textbf{Nat}) \J{ -> } \textbf{Bool} 
    \end{array}
    &
    \begin{array}[t]{r@{}c@{}l}
      \\
      \J{auto} &\J{ : }& \J{(ALL[T] T -> T) -> (ALL[T] T -> T)}
      \\
      \J{auto'} &\J{ : }& \J{ALL[U] (ALL[T] T -> T) -> U -> U}
      \\
      \J{map} &\J{ : }& \J{ALL[T U] (T -> U) -> } \textbf{List}(\J{T}) \J{ -> } \textbf{List}(\J{U})
      \\
      \J{app} &\J{ : }& \J{ALL[T U] (T -> U) -> T -> U}
      \\
      \J{revapp} &\J{ : }& \J{ALL[T U] T -> (T -> U) -> U}
      \\
      \J{runST} &\J{ : }& \J{ALL[T] (ALL[U] } \textbf{ST}(\J{U}, \J{T})\J{) -> T}
      \\
      \J{argST} &\J{ : }& \J{ALL[U] } \textbf{ST}(\J{U}, \textbf{Nat})
      \\
      \J{zero} &\J{ : }& \textbf{ChurchNum}
      \\
      \J{succ} &\J{ : }& \textbf{ChurchNum} \J{ -> } \textbf{ChurchNum}
      \\
      \J{foo} &\J{ : }& \J{ALL[T] (T -> T) -> } \textbf{List}(\J{T}) \J{ -> T}
      \\
      \J{g} &\J{ : }& \J{ALL[T] }\textbf{List}(\J{T}) \J{ -> } \textbf{List}(\J{T}) \J{ -> T}
      \\
      \J{k} &\J{ : }& \J{ALL[T] T -> } \textbf{List}(\J{T}) \J{ -> T}
      \\
      \J{h} &\J{ : }& \textbf{Nat} \J{-> ALL[T] T -> T}
      \\
      \J{l} &\J{ : }& \textbf{List}(\J{ALL[T] } \textbf{Nat} \J{ -> T -> T}\J{)}
      \\
      \J{r} &\J{ : }& \J{(ALL[T] T -> ALL[U] B -> B) -> }\textbf{Nat}
    \end{array}
  \end{array}
  \]
\end{definition}

\noindent
Additionally, for the purpose of experiment, it is useful to assume that   
the typing of some identifiers have already been learned or verified.

\begin{experiment}
  \label{experi:part_1}
  Part 1
  \\
  \scriptsize
  % \setlength{\arrayrulewidth}{0.5mm}
  % \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1}
  \begin{center}
  \centering
  % \setlength{\tabcolsep}{2pt}
  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
    \multicolumn{5}{l}{
      \sz{
      Does $ impl(\textbf{ctx}\entails e)
      $ admit the expression? 
      }
      \sz{
      \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
      }
    } \\
    \hline
    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
    \hline
    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
    \hline\hline
    \pdc. &
    \J{[x => [y => y]]} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{choose(id)} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{choose(nil)(id)} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{[x => x(x)]} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{id(auto)} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{id(auto')} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{choose(id)(auto)} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{choose(id)(auto')} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{foo(choose(ids))(ids)} 
    & \fail & \pass & \pass \\
    \hline
    \pdc. &
    \J{poly(id)} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{poly([x => x])} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{id(poly)([x => x])} 
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{[f => f(succ;zero;@), f(true;@)]}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{[xs => poly(head(xs))]}
    & \pass & \pass & \pass \\
    \hline
  \end{tabular}
  \end{center}
\end{experiment}

\noindent
In the experiment on functions, we 
are interested to observe if our implementation (\textbf{Extrinsic})
can admit various programs on functions. 
For our implementation, we reject a program if the implementation
finds a contradiction, diverges, or infers the top type. 
In our system, we infer a top type when the top level 
expression represents a function and all of its branches 
have errors. Since the top level function is not applied,
the program technically cannot produce an error, but it's 
perfectly useless and would produce an error if applied to any value. 
We compare our system to SuperF and Super without Distributivity (\textbf{SF no D}).

% $\approx \exists \tau, \Delta, \Theta \qua \epsilon, \epsilon, \textbf{ctx} \entails e \hastype \tau \given \Theta, \Delta$

\begin{experiment}
  \label{experi:part_2}
  Part 2
  \\
  \begin{center}
  \scriptsize
  % \setlength{\arrayrulewidth}{0.5mm}
  \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1}
  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
    \multicolumn{5}{l}{
      \sz{
      Does $ impl(\textbf{ctx}\entails e)
      $ admit the expression? 
      }
      \sz{
      \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
      }
    } \\
    \hline
    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
    \hline
    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
    \hline\hline
    \pdc. &
    \J{length(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{tail(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{head(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{single(id)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{cons(id)(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{cons([x => x])(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{append(single(inc))(single(id))}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{g(single(id))(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{map(poly)(single(id))}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{map(head)(single(ids))}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{app(poly)(id)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{revapp(id)(poly)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{runST(argST)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{app(runST)(argST)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{revapp(argST)(runST)}
    & \pass & \pass & \pass \\
    \hline
  \end{tabular}
  \end{center}
\end{experiment}

\begin{experiment}
  \label{experi:part_3}
  Part 3
  \\
  \begin{center}
  \scriptsize
  % \setlength{\arrayrulewidth}{0.5mm}
  \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1}
  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
    \multicolumn{5}{l}{
      \sz{
      Does $ impl(\textbf{ctx}\entails e)
      $ admit the expression? 
      }
      \sz{
      \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
      }
    } \\
    \hline
    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
    \hline
    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
    \hline\hline
    \pdc. &
    \J{k(h)(l)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{k([x => h(x)])(l)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{r([x => [y => y]])}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{auto(id)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{cons(head(ids))(ids)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{head(ids)(succ;succ;succ;zero;@)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{choose(head(ids))}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{def f = revapp(id) in f(poly)}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{choose(id)([x => auto'(x)])}
    & \pass & \pass & \pass \\
    \hline
  \end{tabular}
  \end{center}
\end{experiment}

\begin{experiment}
  \label{experi:part_4}
  Part 4
  \\
  \begin{center}
  \scriptsize
  % \setlength{\arrayrulewidth}{0.5mm}
  \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1}
  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
    \multicolumn{5}{l}{
      \sz{
      Does $ impl(\textbf{ctx}\entails e)
      $ admit the expression? 
      }
      \sz{
      \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
      }
    } \\
    \hline
    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
    \hline
    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
    \hline\hline
    \pdc. &
    \J{def ignore : } \textbf{ChurchNum} \J{ = z in @}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{s}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{def ignore : } \textbf{ChurchNum} \J{ -> } \textbf{ChurchNum} \J{ = s in @}
    & \pass & \pass & \fail \\
    \hline
    \pdc. &
    \J{def ignore : } \textbf{ChurchNum} \J{ = n3 in @}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{def ignore : } \J{@ -> }\textbf{ChurchNum} \J{ = [@ => n3(n3)] in @}
    & \pass & \pass & \fail \\
    \hline
    \pdc. &
    \J{fst(fst(fst(n3([x => x,(zero;@)])(succ;zero;@))))}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{(s(s(z)))(s(s(z)))}
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \cb{
      \J{def to\_church = loop([self => } 
      \\
      \I \J{[zero;@ => z]}
      \\
      \I \J{[succ;n => s(self(n))]}
      \\
      \J{]) in ...}
    }
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \J{def ignore : } \textbf{Nat} \J{ -> } \textbf{ChurchInt} \J{ = to\_church in @}
    & \pass & \pass & \fail \\
    \hline
    \pdc. &
    \J{loop([self => [x => if true;@ then x else self(self)(x)]]} 
    & \fail & \fail & \fail \\
    \hline
    \pdc. &
    \J{[x => x(x)]([x => x(x)])}
    & \pass & \fail & \fail \\
    \hline
    \pdc. &
    \J{auto(auto'(id))}
    & \pass & \pass & \fail \\
    \hline
    \pdc. &
    \cb{
      \J{[y => } 
      \\
      \I \J{def tmp = y in}
      \\
      \I \J{y(const)}
      \\
      \J{][x => x(x)]}
    }
    & \fail & \pass & \pass \\
    \hline
    \pdc. &
    \cb{
      \J{[k => (k([x => x])),(k([x => single(x)])) ](}
      \\
      \I \J{[f => (f(succ;zero;@)),(f(true;@))]}
      \\
      \J{)}
    }
    & \pass & \pass & \pass \\
    \hline
    \pdc. &
    \cb{
      \J{[f => } 
      \\
      \I \J{def a = [@ => f(id)] in }
      \\
      \I \J{def y : } \textbf{Nat} \J{ -> (ALL[T] T -> T) = a(@) in }
      \\
      \I \J{y}
      \\
      \J{](const(const(id)))}
    }
    & \fail & \fail & \pass \\
    \hline
  \end{tabular}
  \end{center}
\end{experiment}

\begin{experiment}
  \label{experi:part_5}
  Part 5 
  \\
  \begin{center}
  \scriptsize
  % \setlength{\arrayrulewidth}{0.5mm}
  \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1}
  \begin{tabular}{|l m{35em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
    \multicolumn{4}{l}{
      \sz{
      Does $ impl(\textbf{ctx}\entails e)
      $ admit the expression? 
      }
      \sz{
      \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
      }
    } \\
    \hline
    \multicolumn{2}{|c|}{} & \multicolumn{2}{c|}{$impl$} \\ 
    \hline
    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{MLStruct} \\ 
    \hline\hline
    \sdc. &
    \cb{
      \J{def stdCmp = [ a,b => } 
      \\
      \I \J{a |> (}
      \\
      \I\I \J{[zero;@ => scalarCmp(a,b)]}
      \\
      \I\I \J{[succ;n => scalarCmp(a,b)]}
      \\
      \I\I \J{[nil;@ => lexicoCmp(a,b)]}
      \\
      \I\I \J{[cons;(x,xs) => lexicoCmp(a,b)]}
      \\
      \I \J{)}
      \\
      \J{] in ...}
    }
    & \pass & \assisted \\
    \hline
    \sdc. &
    \cb{
      \J{def stdSort : (TOP } 
      \\
      \I \J{ \& (} \textbf{List}(\textbf{Nat}) \J{ -> } \textbf{List}(\textbf{Nat}) \J{)}
      \\
      \I \J{ \& (} \textbf{List}(\textbf{List}(\textbf{Nat})) \J{ -> } \textbf{List}(\textbf{List}(\textbf{Nat})) \J{)}
      \\
      \J{) = sort(stdCmp) in ...} 
    }
    & \pass & \assisted \\
    \hline
    \sdc. &
    \cb{
      \J{def double : } \textbf{Nat} \J{ -> } \textbf{Even} \J{ = loop([self => } 
      \\
      \I \J{[zero;@ => zero;@]}
      \\
      \I \J{[succ;n => succ;succ;(self(n))]}
      \\
      \J{]) in ...}
    }
    & \pass & \fail \\
    \hline
    \sdc. &
    \cb{
      \J{def halve : } \textbf{Even} \J{ -> } \textbf{Nat} \J{ = loop([self => } 
      \\
      \I \J{[zero;@ => zero;@]}
      \\
      \I \J{[succ;succ;n => succ;(self(n))]}
      \\
      \J{]) in ...}
    }
    & \pass & \fail \\
    \hline
    \sdc. &
    \cb{
      \J{[zero;@ => [zero;@ => @](@)]}
      \\
      \J{[nil;@ => @]}
    }
    & \pass & \fail \\
    \hline
  \end{tabular}
  \end{center}
\end{experiment}

\noindent
The experiment on programs with constructors demonstrate
that our system is roughly as expressive as untyped languages
without sacrificing safety guarantees.
Our system can easily mix structures together, reducing
the need for programmers to call different interfaces for different data.  
If there is an error in one branch of a function, the
system can still admit the program by ensuring 
that the erroneous branch is never used, and there
is at least one branch that could be used.

\section{Related Work}
\label{sec:related_work}

Over the past decades many type inference have been designed and analyzed 
with varying degrees of expressivity and varying degrees of intrinsic requirements
or extrinsic relaxations, with respect to both functions (e.g. functions or paths dependent on variables)
and constructors (e.g. data constructors or records).  
The first technique to infer parametric types for parametric
data without any assistance from annotations was Hindley-Milner type inference \cite{}. Despite its age, 
it remains one of the foundational 
theoretical underpinnings of practical programming languages used today, such as Standard ML \cite{}. 
For instance, in Standard ML, the system infers the identifier \code{foo} in \code{fun foo f = (f 123)}
to have type \code{forall 'a . (int -> 'a) -> 'a}
Despite its practical utility, a significant drawback of Hindley-Milner type inference is that
certain kinds of flexible programs are simply not allowed.
For instance, in Standard ML, the system rejects the program \code{fun fooPair f = (f 123, f true)}

In order to type the program \code{fooPair}, a type system would either need to determine that
\code{f} is parametric, or that the types of \code{123} and \code{true} are compatible.
Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, which
was introduced into ML-style languages using boxed polymorphism \cite{} and developed further with ML$^F$ \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
or an infinite number of
other parametric types could be used, but none is weaker than all the rest.
Therefore, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which we believe constitutes intrinsic typing of
of functions. Since there is no parameter type that is weaker than all the rest,
relying on intrinsic second order parametric type significantly restricts the reusability
of the program.

In order to avoid this loss of expressivity and the annotation burden of intrinsic typing, 
SuperF \cite{} introduced 
a technique that relies on extrinsic subtyping polymorphism, 
which bypasses the need for intrinsic parametric polymorphism.
In the previous example, SuperF can infer the principal type for \code{fooPair} by inferring the 
weakest type of its parameter \code{f}, which it achieved by leveraging subtyping to find 
a type that is compatible with both \code{123} and \code{true}.
The SuperF type system infers subtyping constraints that represent the data flow represented by
function application and pattern matching. It relies 
intersection and union type combinators to represent the strongest or weakest interpretations
of type variables that meet certain subtyping constraints.   
With SuperF's type inference mechanism, it is possible to infer the principal type for the 
previous problematic example \code{fooPair: (forall 'a 'b . (int -> 'a) /\textbackslash\ (bool -> 'b) -> 'a * 'b)}.

SuperF has a restricted language of types. It is concerned with the type inference
of functions, and it is agnostic towards type inference of constructors. 
It restricts the type and subtyping syntax according to polarity; that is, a
types position within an implication type or subtyping relation. Depending on polar position,
universal types (i.e. universals) may or may not be qualified by subtyping constraints.
The authors of SuperF highlight some technical features, including extrusion and distributivity.
Their extrusion technique ensures that types are not overly specialized
when inferring types of nested functions. Their distributivity technique allows
rewriting universal quantifiers to increase the chance that skolemization occurs before instantiation. 
They point out that the distributivity rule is an improvement for most cases, but there are certain
problems for which it more actually more restrictive.

The high level ideas in SuperF of relying on subtyping constraints were very influential in our work.
Like SuperF, our system relies on syntax-directed subtyping rules to check constraints and to learn 
new subtyping constraints by traversing previously learned constraints;
Additionally, our system also relies on instantiation and skolemization in its subtyping rules. 
Type polarity is also important in our system.
However, our system does not restrict the syntax of types based on polarity. Instead, we rely on
polarity to constrain proof typing and proof subtyping when interpreting or checking interpretations
of type variables.    
Our system does not need extrusion because Packing \ms{\textbf{pack(...)}} preserves the nested structure of 
functions at the type level, such that the constraint between inner and outer bound type variables
qualifies the inner parametric type, which ensures that instantiation of the inner parametric type 
continues to constrain the outer bound type variables.  
Our system does not need distributivity of type variable bindings, because it requires
that skolemization occurs before learning constraints on type variables. Therefore, it is fine to allow
instantiation to occur before skolemization in our system, 
since it does not cause learning constraints before skolemization.

\TODO{look into what SuperF folks mean by no backtracking in light of rhs union}

\TODO{backtracking is necesary to infer strongest possible functions (all the cases)}

\TODO{therefore, they must mean they avoid redundant backtracking, that mearely learns weaker constraints}

In addition to functions, practical programming languages need to have some form of constructors.
Although Hindley-Milner type inference offered an extrinsic type inference solution for predicative functions
very early on, type inference has always relied on some degree of intrinsic types for constructors.
intrinsically typed constructors is exemplified by the datatype mechanism of Standard ML.
For instance, the list datatype \code{'a list = nil | cons 'a * 'a list} specifies that 
data constructed with \code{nil} or \code{cons} have the type \code{'a list}. Additionally,
it requires that the input to the constructor \code{cons}  has the types \code{'a * 'a list}
Although the datatype mechanism has been used with great success across many languages (including Lean, Haskell, Ocaml, and Isabelle),
it enforces a very rigid form of programming. For instance, there is no way to guarantee
that the output of a function is a nonempty list and then use that output in a function that expects an optionally empty list.

To get around this limitation, refinement types \cite{} relaxed the meaning of the datatype mechanism, to 
allow constructions to represent types that are stronger than their intrinsic type.
For instance, \code{nil} could have a type as strong as the single type \code{?nil} or as weak as \code{'a list}.
Likewise, a list of two elements \code{cons(x, cons(y, nil))} could have a type as strong 
\code{?cons 'a * (?cons 'a * nil)}
or as weak as \code{'a list}. 
A refinement type system can choose how strong to make the type based on how the data is used, 
which depends on how data flows, for which subtyping provides a natural formulation. 
It can find these stronger types by starting with intrinsic weakest type and using intersection with 
all the ways it sees the data being used.
However, the refinement relaxation is not enough to allow common flexible idioms used in untyped languages. 
For instance, it would allow pattern matching that chooses between a list and a number.

To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
must be abolished so that there is no intrinsic weakest type other than a top type. 
Typescript \cite{} does not have an intrinsic weakest type for constructors,
and it can infer output types of functions that return different structural forms in different branches.
However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
which can refine the types of variables in branches based on the condition guarding each branch.
For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
the weakest type of parameters.

To extend type inference of scripting languages further, MLStruct \cite{} and others like MLSub \cite{}, 
leveraged the flow information to infer both input types and output types of functions.
MLStruct, along with SuperF, represents some of the foundational ideas of the MLScript language.
Like SuperF, it learns the data flow of programs in terms of subtyping constraints.
Other than our system, it appears that MLStruct has the best type inference of constructors
with the least amount of intrinsic type requirements. 
However, MLStruct still has a significant intrinsic type requirement, which reduces flexibility compared
to untyped languages. It requires that the inputs to data constructors have an intrinsic type.  
In essence, this requirement prescribes a lower bound on data constructions,
which reduces the flexibility of programs compared to untyped languages. 
As our tests (Experiment \ref{experi:typing_structures})
illustrate, it is not possible to type a function that expects only even numbers that is interoperable with natural numbers.
Moreover, the additional annotation burden alone would prevent the analysis of untyped code without
significant effort in augmenting the code with annotations.
In contrast to MLStruct, our system relies completely on learning data flow in the form of subtyping constraints
without any annotation requirements on constructors.

The common ingredient in all of these systems that perform some amount of extrinsic type inference is
learning data flow. Data flow analysis comes in many forms. As the goal for these systems is verification,
subtyping is a natural choice. There has been some work on extracting rules for checking subtyping relations
from the semantic interpretation of types, known as semantic subtyping \cite{}, or set-theoretic types \cite{}.
Other approaches, like algebraic subtyping \cite{}, and Cretin \cite{}, and superF, 
rely on designing syntax-directed subtyping rules and then proving their soundness with respect to some semantic interpretation. 
Additionally, there are similarities between type inference with subtyping and control-flow analysis \cite{}, or polyvariant flow analysis \cite{}.
In control-flow analysis, expressions can be associated with a label, and each label can map to a set of abstract values,
determined by learning subset constraints from the control-flow of the program. 
Instead of labels, type inference with subtyping associate types  
with expressions and learns subtyping constraints.

\TODO{discuss recursive types}

\TODO{discuss regular trees}

\TODO{look into the issue of caching related to recursion}

\section{Conclusion}
\label{sec:conclusion}
We have presented a novel extrinsic type inference system
that can guarantee programs are bug-free while maintaining
code-reusability nearly on par with untyped scripting languages,
and exceeding that of an previous type inference system we know of.  
We have incorporated previous studied ideas of subtyping constraints,
intersection types, union types, parametric types, and recursive types into
a cohesive design that enables more principal types to be inferred 
and more subtyping relations to be verified than in previous systems. 
In particular, our system is able to learn recursive types of function inputs
and outputs purely from the construction of recursive functions without 
any intrinsic recursive type specifications.
Additionally, our system can perform inductive reasoning to
verify than one recursive type subtypes another.
In the future, we will extend the recursive type reasoning capabilities
to handle recursive relational types, enabling the infinite paths
of recursive functions to be represented and reasoned with at the type level. 
With inductive reasoning over relational types, we will be
moving closer to a full fledged theorem proving system,
in which subtyping enables a form of proof automation,
by reusing the proof of one proposition as a proof for another.

\newpage

\begin{thebibliography}{9}

% \bibitem{syngar}
% X. Wang, I, Dillig, R. Singh. Program synthesis using abstraction refinement. In POPL, 2017.

% \bibitem{love}
% M. Heizmann, J. Hoenicke, A. Podelski. Software model checking for people who love automata. In CAV, 2013.

% \bibitem{synquid}
% N. Polikarpova, I. Kuraj, A. Solar-Lezama. Program synthesis from polymorphic refinement types. In PLDI, 2016.

\end{thebibliography}

\end{document}

### 2025-07-08 extrinsic_type_inference_paper/paper.tex
update proof subtying substance proof

@@
 \noindent
-Proof subtyping substance states that generate constraints are satisfiable
+Proof subtyping substance states that the generated constraints are satisfiable
 assuming the contextual constraints are satisfiable, with respect
 to all interpretations of the contextual skolems.   
-The proof is by induction, 
-where in most cases the property follows directly from the the induction hypothesis.
+The proof is by following a slightly stronger statement,
+which says that the generated constraints are satisfiable, and 
+that the constraints on a single variable are as strong or stronger
+than all the constraints with respect to that variable.
+The proof of the stronger statement is by induction.
 The only cases that are actually interesting are the variable rules
 where new constraints are generated.
-Because the rules check for consistency by searching for constraints on single variables,
-it is necessary that the constraints on a single variable are as strong or stronger
-than all the constraints with respect to that variable.
-It then boils down to checking the bound in the goal against all the bounds on the other side
-of the variable, as those are the only opportunities for contradiction.
+In every case that a new constraint is added, it is a constraint on
+a single variable, and in each case the solver compares the new bound on the variable
+to the previous bounds on the other side of the variable. Since bounds on
+opposite sides are the only opportunities for contradictions, the requirement
+preserves the strengthened substance property from the induction hypothesis.
 

### 2025-07-08 extrinsic_type_inference_paper/paper.tex
update

@@
 to all interpretations of the contextual skolems.   
-The proof is by following a slightly stronger statement,
-which says that the generated constraints are satisfiable, and 
+This property follows directly from a slightly stronger theorem,
+which states that the generated constraints are satisfiable and 
 that the constraints on a single variable are as strong or stronger

### 2025-07-08 src/test_analysis_new.py
update experiment formatting

@@ -1043,7 +1043,7 @@ if __name__ == '__main__':
     # test_recursive_relational_factorization_learning_in_typing()
     # test_subtyping_debug()
     # test_max()
-    # test_typing_A9()
+    test_typing_A9()
     # test_subtyping_debug()
 
 

### 2025-07-08 extrinsic_type_inference_paper/paper.tex
update tables; remove lines
update experiment formatting
update some formatting
update formatting
update table formatting
update

@@
 
+\usepackage[table]{xcolor}
+
 \newcounter{pdc}
@@
 \begin{document}
+
+\newcommand{\stoprule}{
+  \addlinespace[2pt]
+  \toprule
+  \addlinespace[5pt]
+}
+
+\newcommand{\sbottomrule}{
+  \addlinespace[2pt]
+  \bottomrule
+  \addlinespace[2pt]
+}
+
+\newcommand{\smidrule}[1]{
+  \addlinespace[2pt]
+  \midrule
+  \addlinespace[2pt]
+}
+
+\newcommand{\scmidrule}[1]{
+  \addlinespace[2pt]
+  \cmidrule{#1}
+  \addlinespace[2pt]
+}
+
+\newcommand{\sline}{
+  \arrayrulecolor{gray!50}
+  \specialrule{0.4pt}{2pt}{2pt}
+  \arrayrulecolor{black}
+}
 
@@
 % \newcommand{\xmark}{\ding{55}}%
-\newcommand{\pass}{\text{\normalsize \color{Green} \ding{51}}}
-\newcommand{\assisted}{\text{\normalsize \color{Purple} \ding{45}}}
-\newcommand{\fail}{\text{\normalsize \color{red} \ding{55}}}
+\newcommand{\pass}{\text{\scriptsize \color{Green} \ding{51}}}
+\newcommand{\assisted}{\text{\scriptsize \color{Purple} \ding{45}}}
+\newcommand{\fail}{\text{\scriptsize \color{red} \ding{55}}}
 
@@
 
-\newcommand{\block}[1]{{\footnotesize \[
-\begin{array}[t]{l}
+\newcommand{\nopad}{
+  \setlength{\abovedisplayskip}{4pt}
+  \setlength{\belowdisplayskip}{4pt}
+}
+
+\newcommand{\examp}[1]{{\nopad \footnotesize 
+\vspace{-1em}
+\[
+\begin{array}[t]{@{}l}
 #1
 \end{array}
-\]}}
+\]
+\vspace{-1em}
+}}
 
 \newcommand{\cb}[1]{
-\begin{array}[t]{l}
-\\
-#1
-\\
-\end{array}
+$
+\begin{array}[t]{@{}l}#1\end{array}
+$
 }
@@
 
-\block{
+\examp{
   \J{fun len(nil) = zero}
@@
 
-\block{
+\examp{
   \J{datatype nat = zero | succ of nat} 
@@
 
-\block{
+\examp{
   \J{fn nil => zero}
@@
 
-\block{
+\examp{
   \J{fun nlen(nil) = zero}
@@
 
-\block{
+\examp{
   \J{fun evlen(nil) = zero}
@@
 
-\block{
+\examp{
   \J{def talky = (}
@@
 
-\block{
+\examp{
   \J{(<hello>@ -> <world>@) \&}
@@
 
-\block{
+\examp{
   \J{def x : <hello> @ | <thank> @ = ... in}
@@
 
-\block{
+\examp{
   \J{<world> @ | <you> @}
@@
 
-\block{
+\examp{
   \J{def foo = ? f => (f(<succ> <succ> <zero> @), f(<true> @))}
@@
 
-\block{
+\examp{
   \J{f : ALL[X] X -> <some> X}
@@
 
-\block{
+\examp{
   \J{f : ALL[X] X -> (LFP[R] <nil> @ | <cons> X*R)}
@@
 
-\block{
+\examp{
   \J{ALL[X Y] (<succ> <succ> <zero> @ -> X) \& (<true> @ -> Y) -> X*Y}
@@
 
-\block{
+\examp{
   \J{F} \subtypes \J{(<succ> <succ> <zero> @ -> X)}
@@
 
-\block{
+\examp{
   \J{def repeat = [x => loop([self => [}
@@
 
-\block{
+\examp{
   \J{ALL[T] T -> GFP[R](} 
@@
 
-\block{
+\examp{
   \J{ALL[T] T ->} 
@@
 
-\block{
+\examp{
   \J{ <succ> <succ> <zero> @} \subtypes \J{(LFP[R] <zero> @ | <succ> R)}
@@
 
-\block{
+\examp{
   \J{def x : (LFP[R] <zero> @ | <succ> <succ> R) = ... in}
@@
 
-\block{
+\examp{
   \J{LFP[R] <zero> @ | <succ> <succ> R}
@@
 
-\block{
+\examp{
   \J{E} \subtypes \J{(LFP[R] <zero> @ | <succ> R)}
@@
 
-\block{
+\examp{
   \J{(<zero> @ | <succ> <succ> (LFP[R] <zero> @ | <succ> R))} 
@@
 
-\block{
+\examp{
   \J{def sort : ALL[T] (T*T->} \textbf{Bool} \J{) -> }\textbf{List}(\J{T}) \J{ -> } \textbf{List}(\J{T}) \J{ = }
@@
 
-\block{
+\examp{
   \J{def scalarCmp : }\textbf{Nat} \J{ -> } \textbf{Nat} \J{ -> } \textbf{Bool} \J{ = ... in}
@@
 
-\block{
+\examp{
   \J{def scalarSort = sort(scalarCmp) in}
@@
 
-\block{
+\examp{
   ...
@@
   \label{def:expression}
+  \nopad 
   \small
@@
   \boxed{\llbracket s \rrbracket^\square = e}
-  \\
+  \nopad
   \begin{align*}
@@
   \label{def:pattern}
+  \small
+  \nopad
   \[\begin{array}[t]{rr@{}l}
@@
   \label{def:value}
+  \small
+  \nopad
   \[\begin{array}[t]{rr@{}l}
@@
   \boxed{e \rightsquigarrow e}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \label{def:value_map}
+  \small
+  \nopad
   \[\begin{array}[t]{rr@{}l}
@@
   \label{def:type}
+  \small
+  \nopad
   \[\begin{array}[t]{rr@{}l}
@@
   \boxed{\llbracket \gamma \rrbracket^\square = \tau}
-  \\
+  \nopad
   \begin{align*}
@@
   \small
+  \nopad
   \[\begin{array}[t]{rr@{}l}
@@
   \boxed{\textbf{set}(C) = S}
-  \\
+  \nopad
   \[
@@
   \boxed{\textbf{seq}(S) = C}
+  \nopad
   \\
@@
 
-% \begin{definition}
-%   \label{def:stability}
-%   Stability 
-%   \hfill
-%   \boxed{\VDash e}
-%   \\
-%   \begin{mathpar}
-%     \inferrule {
-%       e = v
-%     } {
-%       \VDash e
-%     }
-
-%     \inferrule { 
-%       e \rightsquigarrow e' 
-%       \\
-%       \VDash e'
-%     } {
-%       \VDash e
-%     } 
-%   \end{mathpar}
-% \end{definition}
-
-% \noindent
-% A expression is stable if and only if it is a value or it can take a step to a stable expression.
-% In other words, if an expression is not a value then its behavior is defined by progression.
-
-% \begin{property}
-%   \label{def:model_typing_soundness}
-%   Model Typing Soundness 
-%   \\
-%   \begin{mathpar}
-%     \inferrule { 
-%       \delta \satisfies e \hastype \tau
-%     } {
-%       \VDash e 
-%     } 
-%   \end{mathpar}
-% \end{property}
-
-% \noindent
-% Model typing $\delta \satisfies e \hastype \tau$ attempts to imbue types
-% with meaning.  At the very least, types should inform us that an expression is stable.
-
 \begin{definition}
   \label{def:type_map}
+  \small
+  \nopad
   \[\begin{array}[t]{rr@{}l}
@@
   \boxed{\delta \satisfies e \hastype \tau}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\delta \satisfies \tau <: \tau}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\delta \satisfies \Delta}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\delta, \sigma \satisfies \Gamma}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \kappa_l \subtypes \kappa_r \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \phi \subtypes \tau \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau \subtypes \psi \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'} 
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau \subtypes \alpha \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \tau \subtypes \phi \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \psi \subtypes \tau \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \boxed{\Theta, \Delta \entails \Delta_q \given \Theta', \Delta'}
-  \\
+  \nopad
   \begin{mathpar}
@@
   \small
-  \\
+  \nopad
   \begin{mathpar}
@@
   \end{mathpar}
-
 \end{theorem}
@@
 
-\begin{definition} Experimental Expressions and Types  
+\begin{definition} 
   \label{def:experimental_expressions_and_types}
+  \emph{experimental expressions and types} 
   \scriptsize
+  \nopad
   \[
@@
 
-\begin{definition} Experimental Typing Context  
+\begin{definition} 
   \label{def:experimental_typing_context}
+  \emph{experimental typing context}
   \tiny
+  \nopad
   \[
@@
 \begin{experiment}
-  \label{experi:part_1}
-  Part 1
-  \\
+  \label{exp:part_1}
+  Something
+  \hfill
   \scriptsize
-  % \setlength{\arrayrulewidth}{0.5mm}
-  % \setlength{\tabcolsep}{4pt}
-  \renewcommand{\arraystretch}{1}
+  \nopad
   \begin{center}
-  \centering
-  % \setlength{\tabcolsep}{2pt}
-  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
+  \begin{tabular}{l m{30em} >{\centering}m{5em} >{\centering}m{5em} >{\centering\arraybackslash}m{5em} } 
     \multicolumn{5}{l}{
-      \sz{
-      Does $ impl(\textbf{ctx}\entails e)
-      $ admit the expression? 
-      }
-      \sz{
+      \small
+      Does the implementation admit an expression under typing constraints \textbf{ctx}?
+    } \\
+
+    \multicolumn{4}{l}{
+      \small
       \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
-      }
     } \\
-    \hline
-    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
-    \hline
-    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
-    \hline\hline
+
+    \stoprule
+
+    \multicolumn{2}{c}{} & \multicolumn{3}{c}{\small implementation} \\
+
+    \scmidrule{3-5}
+
+    \multicolumn{2}{c}{\small expression} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\
+
+    \smidrule
     \pdc. &
@@
     & \pass & \pass & \pass \\
-    \hline
+
+    \sline
+
     \pdc. &
@@
     & \pass & \pass & \pass \\
-    \hline
+
+    \sline
+
     \pdc. &
     \J{foo(choose(ids))(ids)} 
-    & \fail & \pass & \pass \\
-    \hline
+    & \pass & \pass & \pass \\
+
+    \sline
+
     \pdc. &
@@
     & \pass & \pass & \pass \\
-    \hline
+
+    \sbottomrule
   \end{tabular}
@@
   \label{experi:part_2}
-  Part 2
-  \\
+  \hfill
+  \scriptsize
+  \nopad
   \begin{center}
-  \scriptsize
-  % \setlength{\arrayrulewidth}{0.5mm}
-  \setlength{\tabcolsep}{4pt}
-  \renewcommand{\arraystretch}{1}
-  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
+  \begin{tabular}{l m{30em} >{\centering}m{5em} >{\centering}m{5em} >{\centering\arraybackslash}m{5em} } 
     \multicolumn{5}{l}{
-      \sz{
-      Does $ impl(\textbf{ctx}\entails e)
-      $ admit the expression? 
-      }
-      \sz{
+      \small
+      Does the implementation admit an expression under typing constraints \textbf{ctx}?
+    } \\
+
+    \multicolumn{4}{l}{
+      \small
       \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
-      }
     } \\
-    \hline
-    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
-    \hline
-    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
-    \hline\hline
+
+    \stoprule
+
+    \multicolumn{2}{c}{} & \multicolumn{3}{c}{\small implementation} \\
+
+    \scmidrule{3-5}
+
+    \multicolumn{2}{c}{\small expression} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\
+
+    \smidrule
+
     \pdc. &
@@
   \label{experi:part_3}
-  Part 3
-  \\
+  \hfill
+  \scriptsize
+  \nopad
   \begin{center}
-  \scriptsize
-  % \setlength{\arrayrulewidth}{0.5mm}
-  \setlength{\tabcolsep}{4pt}
-  \renewcommand{\arraystretch}{1}
-  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
+  \begin{tabular}{l m{30em} >{\centering}m{5em} >{\centering}m{5em} >{\centering\arraybackslash}m{5em} } 
     \multicolumn{5}{l}{
-      \sz{
-      Does $ impl(\textbf{ctx}\entails e)
-      $ admit the expression? 
-      }
-      \sz{
+      \small
+      Does the implementation admit an expression under typing constraints \textbf{ctx}?
+    } \\
+
+    \multicolumn{4}{l}{
+      \small
       \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
-      }
     } \\
-    \hline
-    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
-    \hline
-    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
-    \hline\hline
+
+    \stoprule
+
+    \multicolumn{2}{c}{} & \multicolumn{3}{c}{\small implementation} \\
+
+    \scmidrule{3-5}
+
+    \multicolumn{2}{c}{\small expression} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\
+
+    \smidrule
+
     \pdc. &
@@
   \label{experi:part_4}
-  Part 4
-  \\
+  \hfill
+  \nopad
+  \scriptsize
   \begin{center}
-  \scriptsize
-  % \setlength{\arrayrulewidth}{0.5mm}
-  \setlength{\tabcolsep}{4pt}
-  \renewcommand{\arraystretch}{1}
-  \begin{tabular}{|l m{30em} | >{\centering}m{5em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
+  \begin{tabular}{l m{30em} >{\centering}m{5em} >{\centering}m{5em} >{\centering\arraybackslash}m{5em} } 
     \multicolumn{5}{l}{
-      \sz{
-      Does $ impl(\textbf{ctx}\entails e)
-      $ admit the expression? 
-      }
-      \sz{
+      \small
+      Does the implementation admit an expression under typing constraints \textbf{ctx}?
+    } \\
+
+    \multicolumn{4}{l}{
+      \small
       \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
-      }
     } \\
-    \hline
-    \multicolumn{2}{|c|}{} & \multicolumn{3}{c|}{$impl$} \\ 
-    \hline
-    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\ 
-    \hline\hline
+
+    \stoprule
+
+    \multicolumn{2}{c}{} & \multicolumn{3}{c}{\small implementation} \\
+
+    \scmidrule{3-5}
+
+    \multicolumn{2}{c}{\small expression} & \textbf{Extrinsic} & \textbf{SuperF} & \textbf{SF no D} \\
+
+    \smidrule
+
     \pdc. &
@@
     & \pass & \pass & \fail \\
-    \hline
+
+    \sline
+
     \pdc. &
@@
     & \fail & \fail & \fail \\
-    \hline
+
+    \sline
+
     \pdc. &
@@
     & \pass & \fail & \fail \\
-    \hline
+
+    \sline
+
     \pdc. &
@@
     & \fail & \pass & \pass \\
-    \hline
+
+    \sline
+
     \pdc. &
@@
     & \fail & \fail & \pass \\
-    \hline
+
+    \sbottomrule
+
   \end{tabular}
@@
   \label{experi:part_5}
-  Part 5 
-  \\
+  \hfill
+  \nopad
+  \scriptsize
   \begin{center}
-  \scriptsize
-  % \setlength{\arrayrulewidth}{0.5mm}
-  \setlength{\tabcolsep}{4pt}
-  \renewcommand{\arraystretch}{1}
-  \begin{tabular}{|l m{35em} | >{\centering}m{5em} | >{\centering\arraybackslash}m{5em} |} 
+  \begin{tabular}{p{1em} l >{\centering}p{5em} >{\centering\arraybackslash}p{5em}} 
     \multicolumn{4}{l}{
-      \sz{
-      Does $ impl(\textbf{ctx}\entails e)
-      $ admit the expression? 
-      }
-      \sz{
+      \small
+      Does the implementation admit an expression under typing constraints \textbf{ctx}?
+    } \\
+
+    \multicolumn{4}{l}{
+      \small
       \pass\ yes \ \ $\cdot$\ \ \fail\ no \ \ $\cdot$\ \ \assisted\ conditionally w/ annotations 
-      }
     } \\
-    \hline
-    \multicolumn{2}{|c|}{} & \multicolumn{2}{c|}{$impl$} \\ 
-    \hline
-    \multicolumn{2}{|c|}{$e$} & \textbf{Extrinsic} & \textbf{MLStruct} \\ 
-    \hline\hline
+
+    \stoprule
+
+    \multicolumn{2}{c}{} & \multicolumn{2}{c}{\small implementation} \\
+
+    \scmidrule{3-4}
+
+    \multicolumn{2}{c}{\small expression} & \textbf{Extrinsic} & \textbf{MLStruct} \\
+
+    \smidrule
+
     \sdc. &
@@
     & \pass & \assisted \\
-    \hline
+
+    \sline
+
     \sdc. &
@@
     & \pass & \fail \\
-    \hline
+
+    \sline
+
     \sdc. &
@@
     & \pass & \fail \\
-    \hline
+
+    \sbottomrule
+
   \end{tabular}
@@
 \end{experiment}
+
+\TODO{add more interesting examples and explain why they're interesting}
+
+\TODO{add example of of inductive reasoning to prove that an even number is not three}
+
+\TODO{add example of path sensitivity for if then else conditional}
 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
add generator examples

@@
 
-\TODO{discuss recursive types}
-
-\TODO{discuss regular trees}
+\TODO{discuss implicit and flow based approaches}
+
+\TODO{discuss universal, existential, iso-recursive types}
+
+\TODO{note how our structural requirements on recursive types prevent typing generators}
+
+\TODO{mention regular tree reasoning}
+
+\TODO{discuss how every types of expresions are implictly determine}
+
+\TODO{discuss how types are iso-recursive, but are implicitl}
 

### 2025-07-09 src/tapas/slim/analyzer.py

@@ -1619,6 +1619,15 @@ class Solver:
         checker._checking = True 
         return checker
 
+    def is_useless(self, positive : bool, t : Typ):
+        ult = Top() if positive else Bot()
+        if t == ult: 
+            return True
+        elif isinstance(t, Imp): 
+            return self.is_useless(not positive, t.antec) or self.is_useless(positive, t.consq)
+
+
+
     def simplify_typ(self, typ : Typ) -> Typ:
         if False:
             assert False
@@ -1671,11 +1680,19 @@ class Solver:
         elif isinstance(typ, Imp): 
             return Imp(self.simplify_typ(typ.antec), self.simplify_typ(typ.consq))
         elif isinstance(typ, Exi):
-            return Exi(typ.ids, self.simplify_constraints(typ.constraints), self.simplify_typ(typ.body))
+            if not typ.ids and not typ.constraints:
+                return self.simplify_typ(typ.body)
+            elif not typ.constraints and isinstance(typ.body, TVar) and typ.body.id in typ.ids:
+                return Top()
+            else:
+                return Exi(typ.ids, self.simplify_constraints(typ.constraints), self.simplify_typ(typ.body))
         elif isinstance(typ, All):
             return All(typ.ids, self.simplify_constraints(typ.constraints), self.simplify_typ(typ.body))
         elif isinstance(typ, LeastFP):
-            return LeastFP(typ.id, self.simplify_typ(typ.body))
+            if typ.id not in extract_free_vars_from_typ(s(), typ.body):
+                return self.simplify_typ(typ.body)
+            else:
+                return LeastFP(typ.id, self.simplify_typ(typ.body))
         else:
             return typ
         
@@ -3570,6 +3587,22 @@ result:
 
             influential_ids = foreignids.union([self_typ.id]).union(local_closedids).union(extract_free_vars_from_typ(s(), rel_pattern))
             influential_constraints = filter_constraints_by_all_variables(local_constraints, influential_ids)
+
+            print(f"""
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+COMBINE FIX
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+goal:
+{concretize_typ(body_typ)} <: {concretize_typ(Imp(self_typ, Imp(in_typ, out_typ)))}
+
+IH_typ.id
+{IH_typ.id}
+
[…]

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
update notes and notation

--- 2025-07-08 extrinsic_type_inference_paper/paper.tex
@@
       \\
-      \Theta', \Delta' \entails \tau \subtypes \alpha_\nu \rightarrow \Pi_\nu
+      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
       \\\\
-      \alpha_\nu \downarrow \Pi_\nu
+      \alpha \downarrow \Pi
       \fallingdotseq 
-      \alpha_\mu \uparrow \Pi_\mu
-      \\
-      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha_\mu \entails \Pi_\mu) = \tau_\mu
+      \alpha' \uparrow \Pi'
+      \\
+      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha' \entails \Pi') = \tau'
       \\\\
-      \factorsinto \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{left} \mapsto \tau_l
-      \\
-      \factorsinto \J{LFP[} \alpha_\mu \J{]}\tau_\mu \subtypes \J{right} \mapsto \tau_r
+      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{left} \mapsto \tau_l
+      \\
+      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{right} \mapsto \tau_r
     } {
@@
 
-\TODO{discuss recursive types}
-
-\TODO{discuss regular trees}
-
-\TODO{look into the issue of caching related to recursion}
+\TODO{discuss implicit and flow based approaches}
+
+\TODO{enumerate the various work with various type forms}
+
+\TODO{note how our structural requirements on recursive types prevent typing generators}
+
+\TODO{note the distinction between our least fixed point types and the more general 
+recursive types}
 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
organize and add notes

@@
 without sacrificing safety.
+
+\TODO{summarize contributions}
+\TODO{full extrinsic, no annotations required to very programs on par with untyped pgograms in terms of code reusability}
+\TODO{Normalizing all approximate control flow information into subtyping, allowing types to be infered in both directions}
+\TODO{inductive reasoning on scalars without any intrinsic specifications}
 
@@
 
+\TODO{add example of a counter stream}
+
 \noindent
@@
 
-Over the past decades many type inference have been designed and analyzed 
-with varying degrees of expressivity and varying degrees of intrinsic requirements
-or extrinsic relaxations, with respect to both functions (e.g. functions or paths dependent on variables)
-and constructors (e.g. data constructors or records).  
-The first technique to infer parametric types for parametric
-data without any assistance from annotations was Hindley-Milner type inference \cite{}. Despite its age, 
-it remains one of the foundational 
+Our system approximates values with many forms of types
+and reasons over their control with with subtyping. Our work
+builds on the techniques from MLStruct \cite{} and SuperF \cite{}, which
+like our system must reason over the combination of numerous forms of types.
+Leading up to these systems, there have been numerous techniques developed
+to reason over subsets of these type forms, with varying degrees 
+of intrinsic typing.
+
+% Over the past decades many type inference have been designed and analyzed 
+% with varying degrees of expressivity and varying degrees of intrinsic requirements
+% or extrinsic relaxations, with respect to both functions (e.g. functions or paths dependent on variables)
+% and constructors (e.g. data constructors or records).
+
+The first technique to infer parametric types without any assistance 
+from annotations was Hindley-Milner type inference (HM)\cite{}. 
+Despite its age, it remains one of the foundational 
 theoretical underpinnings of practical programming languages used today, such as Standard ML \cite{}. 
@@
 to have type \code{forall 'a . (int -> 'a) -> 'a}
-Despite its practical utility, a significant drawback of Hindley-Milner type inference is that
-certain kinds of flexible programs are simply not allowed.
-For instance, in Standard ML, the system rejects the program \code{fun fooPair f = (f 123, f true)}
+Despite its practical utility, a significant drawback of HM is that
+certain kinds of general-purpose expressions are simply not allowed.
+For instance, in Standard ML, the system rejects the program \code{fun fooPair f = (f 123, f true)}.
 
@@
 \code{f} is parametric, or that the types of \code{123} and \code{true} are compatible.
-Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, which
-was introduced into ML-style languages using boxed polymorphism \cite{} and developed further with ML$^F$ \cite{}. 
+Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, 
+which was introduced by System F \cite{}, 
+introduced into ML-style languages with boxed polymorphism \cite{} 
+and developed further with ML$^F$ \cite{} and various other systems \cite{}. 
 Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
@@
 other parametric types could be used, but none is weaker than all the rest.
-Therefore, these systems with impredicative parametric polymorphism rely on annotations of parameters to
-determine second order parametric types, which we believe constitutes intrinsic typing of
-of functions. Since there is no parameter type that is weaker than all the rest,
-relying on intrinsic second order parametric type significantly restricts the reusability
-of the program.
+Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to
+determine second order parametric types, which constitutes intrinsic typing of
+of functions. 
+Since there is no parameter type that is weaker than all the rest,
+merely relying on intrinsic second order parametric types 
+significantly restricts the reusability of expressions.
+Additionally, many of these techniques require explicit instantiation to use parametrically typed
+expressions, which can be seen as an intrinsic restriction that 
+delays pruning the control-flow until a particular usage.
+
+\TODO{mention impredicative types with subtyping, bidirectional typing, local type inferenc (Pierce)}
+
+\TODO{mention how inferring parameter types and instations is processed before and independent from their subtyping check}
+
+In addition to functions, practical programming languages need to have some form of constructors.
+Although Hindley-Milner type inference offered an extrinsic type inference solution for predicative functions
+very early on, type inference has always relied on some degree of intrinsic types for constructors.
+intrinsically typed constructors is exemplified by the datatype mechanism of Standard ML.
+For instance, the list datatype \code{'a list = nil | cons 'a * 'a list} specifies that 
+data constructed with \code{nil} or \code{cons} have the type \code{'a list}. Additionally,
+it requires that the input to the constructor \code{cons}  has the types \code{'a * 'a list}
+Although the datatype mechanism has been used with great success across many languages (including Lean, Haskell, Ocaml, and Isabelle),
+it enforces a very rigid form of programming. For instance, there is no way to guarantee
+that the output of a function is a nonempty list and then use that output in a function that expects an optionally empty list.
+
+To get around this limitation, refinement types \cite{} relaxed the meaning of the datatype mechanism, to 
+allow constructions to represent types that are stronger than their intrinsic type.
+For instance, \code{nil} could have a type as strong as the single type \code{?nil} or as weak as \code{'a list}.
+Likewise, a list of two elements \code{cons(x, cons(y, nil))} could have a type as strong 
+\code{?cons 'a * (?cons 'a * nil)}
+or as weak as \code{'a list}. 
+A refinement type system can choose how strong to make the type based on how the data is used, 
+which depends on how data flows, for which subtyping provides a natural formulation. 
+It can find these stronger types by starting with intrinsic weakest type and using intersection with 
+all the ways it sees the data being used.
+However, the refinement relaxation is not enough to allow common flexible idioms used in untyped languages. 
+For instance, it would allow pattern matching that chooses between a list and a number.
+
+To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
+must be abolished so that there is no intrinsic weakest type other than a top type. 
+Typescript \cite{} does not have an intrinsic weakest type for constructors,
+and it can infer output types of functions that return different structural forms in different branches.
+However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
+Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
+which can refine the types of variables in branches based on the condition guarding each branch.
+For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
+the weakest type of parameters.
+
+\TODO{mention algebraic subtyping, Cretin, and flow analysis}
+
+\TODO{note how solving subtyping constraints subsume the notion of bidirection typig}
+
+The common ingredient in all of these systems that perform some amount of extrinsic type inference is
+learning data flow. Data flow analysis comes in many forms. As the goal for these systems is verification,
+subtyping is a natural choice. There has been some work on extracting rules for checking subtyping relations
+from the semantic interpretation of types, known as semantic subtyping \cite{}, or set-theoretic types \cite{}.
+Other approaches, like algebraic subtyping \cite{}, and Cretin \cite{}, and superF, 
+rely on designing syntax-directed subtyping rules and then proving their soundness with respect to some semantic interpretation. 
+Additionally, there are similarities between type inference with subtyping and control-flow analysis \cite{}, or polyvariant flow analysis \cite{}.
+In control-flow analysis, expressions can be associated with a label, and each label can map to a set of abstract values,
+determined by learning subset constraints from the control-flow of the program. 
+Instead of labels, type inference with subtyping associate types  
+with expressions and learns subtyping constraints.
 
@@
 
-In addition to functions, practical programming languages need to have some form of constructors.
-Although Hindley-Milner type inference offered an extrinsic type inference solution for predicative functions
-very early on, type inference has always relied on some degree of intrinsic types for constructors.
-intrinsically typed constructors is exemplified by the datatype mechanism of Standard ML.
-For instance, the list datatype \code{'a list = nil | cons 'a * 'a list} specifies that 
-data constructed with \code{nil} or \code{cons} have the type \code{'a list}. Additionally,
-it requires that the input to the constructor \code{cons}  has the types \code{'a * 'a list}
-Although the datatype mechanism has been used with great success across many languages (including Lean, Haskell, Ocaml, and Isabelle),
-it enforces a very rigid form of programming. For instance, there is no way to guarantee
-that the output of a function is a nonempty list and then use that output in a function that expects an optionally empty list.
-
-To get around this limitation, refinement types \cite{} relaxed the meaning of the datatype mechanism, to 
-allow constructions to represent types that are stronger than their intrinsic type.
-For instance, \code{nil} could have a type as strong as the single type \code{?nil} or as weak as \code{'a list}.
-Likewise, a list of two elements \code{cons(x, cons(y, nil))} could have a type as strong 
-\code{?cons 'a * (?cons 'a * nil)}
-or as weak as \code{'a list}. 
-A refinement type system can choose how strong to make the type based on how the data is used, 
-which depends on how data flows, for which subtyping provides a natural formulation. 
-It can find these stronger types by starting with intrinsic weakest type and using intersection with 
-all the ways it sees the data being used.
-However, the refinement relaxation is not enough to allow common flexible idioms used in untyped languages. 
-For instance, it would allow pattern matching that chooses between a list and a number.
-
-To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
-must be abolished so that there is no intrinsic weakest type other than a top type. 
-Typescript \cite{} does not have an intrinsic weakest type for constructors,
-and it can infer output types of functions that return different structural forms in different branches.
-However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
-Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
-which can refine the types of variables in branches based on the condition guarding each branch.
-For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
-the weakest type of parameters.
-
 To extend type inference of scripting languages further, MLStruct \cite{} and others like MLSub \cite{}, 
@@
 without any annotation requirements on constructors.
-
-The common ingredient in all of these systems that perform some amount of extrinsic type inference is
-learning data flow. Data flow analysis comes in many forms. As the goal for these systems is verification,
-subtyping is a natural choice. There has been some work on extracting rules for checking subtyping relations
-from the semantic interpretation of types, known as semantic subtyping \cite{}, or set-theoretic types \cite{}.
-Other approaches, like algebraic subtyping \cite{}, and Cretin \cite{}, and superF, 
-rely on designing syntax-directed subtyping rules and then proving their soundness with respect to some semantic interpretation. 
-Additionally, there are similarities between type inference with subtyping and control-flow analysis \cite{}, or polyvariant flow analysis \cite{}.
-In control-flow analysis, expressions can be associated with a label, and each label can map to a set of abstract values,
-determined by learning subset constraints from the control-flow of the program. 
-Instead of labels, type inference with subtyping associate types  
-with expressions and learns subtyping constraints.
 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
add preamble to theory

@@
     \inferrule {
+      \TODO{modify with case for non-recursive parameter type}
+      \\\\
       \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
@@
 
+Our design depends on numerous helpers to rewrite and check the syntax
+of types. The details of these helpers play a key role in 
+determining the level of expressivity allowed by the type inference system. 
+However, these details are beyond the scope of the theory in work.
+In this theory, we are agnostic to their specific details,
+but we require that they meet certain requirements. 
+From these abstract requirements, we can guarantee
+that the design of our proof semantics is correct with respect
+to our model semantics.
+
+\TODO{add requirements on packing, polar sub and tidying, and negatble checks}
+
+\TODO{for polar sub, dye type by subbing in a fresh skolem}
+
 \begin{theorem} 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
add comment on bidirectional type inference

@@
 
-\TODO{add requirements on packing, polar sub and tidying, and negatble checks}
-
-\TODO{for polar sub, dye type by subbing in a fresh skolem}
+\TODO{add requirements on factoring, packing, polar sub and tidying, and negatble checks}
+
+\begin{theorem}
+  \label{thm:proof_subtyping_substance}
+  \emph{proof subtyping substance}
+  \small
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
+    } {
+      \forall \delta \qua \textbf{dom}(\delta) \subseteq \Theta \implies
+      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta
+      \implies
+      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta'
+    }
+  \end{mathpar}
+\end{theorem}
+
+\noindent
+Proof subtyping substance states that the generated constraints are satisfiable
+assuming the contextual constraints are satisfiable, with respect
+to all interpretations of the contextual skolems.   
+This property follows directly from a slightly stronger theorem,
+which states that the generated constraints are satisfiable and 
+that the constraints on a single variable are as strong or stronger
+than all the constraints with respect to that variable.
+The proof of the stronger statement is by induction.
+The only cases that are actually interesting are the variable rules
+where new constraints are generated.
+In every case that a new constraint is added, it is a constraint on
+a single variable, and in each case the solver compares the new bound on the variable
+to the previous bounds on the other side of the variable. Since bounds on
+opposite sides are the only opportunities for contradictions, the requirement
+preserves the strengthened substance property from the induction hypothesis.
 
@@
 \begin{theorem}
-  \label{thm:proof_subtyping_substance}
-  \emph{proof subtyping substance}
+  \label{thm:proof_typing_substance}
+  \emph{proof typing substance} 
   \small
@@
     \inferrule {
-      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
+      \Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta'
     } {
@@
 \noindent
-Proof subtyping substance states that the generated constraints are satisfiable
+Proof typing substance states that the generated subtyping constraints are satisfiable
 assuming the contextual constraints are satisfiable, with respect
 to all interpretations of the contextual skolems.   
-This property follows directly from a slightly stronger theorem,
-which states that the generated constraints are satisfiable and 
-that the constraints on a single variable are as strong or stronger
-than all the constraints with respect to that variable.
-The proof of the stronger statement is by induction.
-The only cases that are actually interesting are the variable rules
-where new constraints are generated.
-In every case that a new constraint is added, it is a constraint on
-a single variable, and in each case the solver compares the new bound on the variable
-to the previous bounds on the other side of the variable. Since bounds on
-opposite sides are the only opportunities for contradictions, the requirement
-preserves the strengthened substance property from the induction hypothesis.
+The proof is by induction, and it follows easily from the induction hypothesis
+and proof subtyping soundness.
 
@@
 but we can easily compose it with another property to do so.
-
-\begin{theorem}
-  \label{thm:proof_typing_substance}
-  \emph{proof typing substance} 
-  \small
-  \nopad
-  \begin{mathpar}
-    \inferrule {
-      \Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta'
-    } {
-      \forall \delta \qua \textbf{dom}(\delta) \subseteq \Theta \implies
-      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta 
-      \implies
-      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta'
-    }
-  \end{mathpar}
-\end{theorem}
-
-\noindent
-Proof typing substance states that the generated subtyping constraints are satisfiable
-assuming the contextual constraints are satisfiable, with respect
-to all interpretations of the contextual skolems.   
-The proof is by induction, and it follows easily from the induction hypothesis
-and proof subtyping soundness.
 
@@
 
-\TODO{mention impredicative types with subtyping, bidirectional typing, local type inferenc (Pierce)}
-
-\TODO{mention how inferring parameter types and instations is processed before and independent from their subtyping check}
+To reduce the annotation burden and increase expressivity, researchers
+developed ways to infer type instantiations and compare type with varying
+amounts of possible instantiations, by combining impredicative types
+with subtyping and type inference techniques. Bidirectional type inference \cite{}  
+is one such technique where bounds on type variables can be inferred in 
+both directions of the control flow.
+It propagates inferred and annotated knowledge directly from
+the structures of the program and relies on subtyping merely to check compatibility between
+types. Additionally, whether a non-parametric type is an instantiation
+of a parametric type is treated as distinct from subtyping.
 
@@
 by reusing the proof of one proposition as a proof for another. 
+We may also explore different strategies for rewriting types 
+to get as much expressivity as possible.
 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
udpate related work

@@
 Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, 
-which was introduced by System F \cite{}, 
-introduced into ML-style languages with boxed polymorphism \cite{} 
-and developed further with ML$^F$ \cite{} and various other systems \cite{}. 
+which was introduced by System F \cite{}.
+
+To reduce the annotation burden and increase expressivity, researchers
+developed ways to infer type instantiations and compare types with varying
+multiple possible instantiations, by combining impredicative types
+with type inference techniques and ways to compare approximations of varying precision .
+
+One of possibility is bidirectional type inference \cite{}  
+is one such technique where bounds on type variables can be inferred in 
+both directions of the control flow.
+It propagates inferred and annotated knowledge directly from
+the structures of the program and relies on subtyping merely to check compatibility between
+types. Additionally, whether a non-parametric type is an instantiation
+of a parametric type is treated as distinct from subtyping.
+
+In order to maintain HM's implicit first-order types in the context
+of higher-order types, other techniques were explored.  
+These techniques include boxed polymorphism \cite{}, ML$^F$ \cite{}, and various other systems \cite{}. 
 Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
@@
 
-To reduce the annotation burden and increase expressivity, researchers
-developed ways to infer type instantiations and compare type with varying
-amounts of possible instantiations, by combining impredicative types
-with subtyping and type inference techniques. Bidirectional type inference \cite{}  
-is one such technique where bounds on type variables can be inferred in 
-both directions of the control flow.
-It propagates inferred and annotated knowledge directly from
-the structures of the program and relies on subtyping merely to check compatibility between
-types. Additionally, whether a non-parametric type is an instantiation
-of a parametric type is treated as distinct from subtyping.
-
-In addition to functions, practical programming languages need to have some form of constructors.
-Although Hindley-Milner type inference offered an extrinsic type inference solution for predicative functions
-very early on, type inference has always relied on some degree of intrinsic types for constructors.
-intrinsically typed constructors is exemplified by the datatype mechanism of Standard ML.
-For instance, the list datatype \code{'a list = nil | cons 'a * 'a list} specifies that 
-data constructed with \code{nil} or \code{cons} have the type \code{'a list}. Additionally,
-it requires that the input to the constructor \code{cons}  has the types \code{'a * 'a list}
-Although the datatype mechanism has been used with great success across many languages (including Lean, Haskell, Ocaml, and Isabelle),
-it enforces a very rigid form of programming. For instance, there is no way to guarantee
-that the output of a function is a nonempty list and then use that output in a function that expects an optionally empty list.
-
-To get around this limitation, refinement types \cite{} relaxed the meaning of the datatype mechanism, to 
-allow constructions to represent types that are stronger than their intrinsic type.
-For instance, \code{nil} could have a type as strong as the single type \code{?nil} or as weak as \code{'a list}.
-Likewise, a list of two elements \code{cons(x, cons(y, nil))} could have a type as strong 
-\code{?cons 'a * (?cons 'a * nil)}
-or as weak as \code{'a list}. 
-A refinement type system can choose how strong to make the type based on how the data is used, 
-which depends on how data flows, for which subtyping provides a natural formulation. 
-It can find these stronger types by starting with intrinsic weakest type and using intersection with 
-all the ways it sees the data being used.
-However, the refinement relaxation is not enough to allow common flexible idioms used in untyped languages. 
-For instance, it would allow pattern matching that chooses between a list and a number.
-
-To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
-must be abolished so that there is no intrinsic weakest type other than a top type. 
-Typescript \cite{} does not have an intrinsic weakest type for constructors,
-and it can infer output types of functions that return different structural forms in different branches.
-However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
-Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
-which can refine the types of variables in branches based on the condition guarding each branch.
-For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
-the weakest type of parameters.
-
-\TODO{mention algebraic subtyping, Cretin, and flow analysis}
-
-\TODO{note how solving subtyping constraints subsume the notion of bidirection typig}
-
-The common ingredient in all of these systems that perform some amount of extrinsic type inference is
-learning data flow. Data flow analysis comes in many forms. As the goal for these systems is verification,
-subtyping is a natural choice. There has been some work on extracting rules for checking subtyping relations
-from the semantic interpretation of types, known as semantic subtyping \cite{}, or set-theoretic types \cite{}.
-Other approaches, like algebraic subtyping \cite{}, and Cretin \cite{}, and superF, 
-rely on designing syntax-directed subtyping rules and then proving their soundness with respect to some semantic interpretation. 
-Additionally, there are similarities between type inference with subtyping and control-flow analysis \cite{}, or polyvariant flow analysis \cite{}.
-In control-flow analysis, expressions can be associated with a label, and each label can map to a set of abstract values,
-determined by learning subset constraints from the control-flow of the program. 
-Instead of labels, type inference with subtyping associate types  
-with expressions and learns subtyping constraints.
-
-In order to avoid this loss of expressivity and the annotation burden of intrinsic typing, 
+All the type inference techniques depend reasoning over the observed control flow  
+of programs, but the reasoning to particular situations and objects and use different
+relations for solving vs checking.  The subtyping relation was generalized in the theory of implicit coercions (Cretin) \cite{},
+such that even instantiation is treated as subtyping (or implicit coercion).  
+Based on these ideas, SuperF \cite{} introduced type inference techniques for generating actually solving
+these generalized subtyping constraints
+
 SuperF \cite{} introduced 
@@
 
+\TODO{recursive types only matter in light of structures}
+
+In addition to functions, practical programming languages need to have some form of constructors.
+Although Hindley-Milner type inference offered an extrinsic type inference solution for predicative functions
+very early on, type inference has always relied on some degree of intrinsic types for constructors.
+intrinsically typed constructors is exemplified by the datatype mechanism of Standard ML.
+For instance, the list datatype \code{'a list = nil | cons 'a * 'a list} specifies that 
+data constructed with \code{nil} or \code{cons} have the type \code{'a list}. Additionally,
+it requires that the input to the constructor \code{cons}  has the types \code{'a * 'a list}
+Although the datatype mechanism has been used with great success across many languages (including Lean, Haskell, Ocaml, and Isabelle),
+it enforces a very rigid form of programming. For instance, there is no way to guarantee
+that the output of a function is a nonempty list and then use that output in a function that expects an optionally empty list.
+
+To get around this limitation, refinement types \cite{} relaxed the meaning of the datatype mechanism, to 
+allow constructions to represent types that are stronger than their intrinsic type.
+For instance, \code{nil} could have a type as strong as the single type \code{?nil} or as weak as \code{'a list}.
+Likewise, a list of two elements \code{cons(x, cons(y, nil))} could have a type as strong 
+\code{?cons 'a * (?cons 'a * nil)}
+or as weak as \code{'a list}. 
+A refinement type system can choose how strong to make the type based on how the data is used, 
+which depends on how data flows, for which subtyping provides a natural formulation. 
+It can find these stronger types by starting with intrinsic weakest type and using intersection with 
+all the ways it sees the data being used.
+However, the refinement relaxation is not enough to allow common flexible idioms used in untyped languages. 
+For instance, it would allow pattern matching that chooses between a list and a number.
+
+To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
+must be abolished so that there is no intrinsic weakest type other than a top type.
+
+In MLSub and algebraic subtyping \cite{}, forgoes the datatype mechanism.
+Instead, it uses type inference to generates subtyping constraints and solve them, akin to unification in HM.
+
+\TODO{understand the capabilities and limitations of MLSub better}
+
+Popular scripting languages have also been augmented with some type inference capabilities.
+Typescript \cite{} does not have an intrinsic weakest type for constructions,
+and it can infer output types of functions that return different structural forms in different branches.
+However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
+Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
+which can refine the types of variables in branches based on the condition guarding each branch.
+For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
+the weakest type of parameters.
+
 To extend type inference of scripting languages further, MLStruct \cite{} and others like MLSub \cite{}, 
@@
 
-\TODO{discuss implicit and flow based approaches}
-
-\TODO{enumerate the various work with various type forms}
-
-\TODO{note how our structural requirements on recursive types prevent typing generators}
-
-\TODO{note the distinction between our least fixed point types and the more general 
-recursive types}
+\TODO{note work on recursive types and regular trees}
+
+The common ingredient in all of these systems that perform some amount of extrinsic type inference is
+learning data flow. Data flow analysis comes in many forms. As the goal for these systems is verification,
+subtyping is a natural choice. There has been some work on extracting rules for checking subtyping relations
+from the semantic interpretation of types, known as semantic subtyping \cite{}, or set-theoretic types \cite{}.
+Other approaches, like algebraic subtyping \cite{}, and Cretin \cite{}, and superF, 
+rely on designing syntax-directed subtyping rules and then proving their soundness with respect to some semantic interpretation. 
+Additionally, there are similarities between type inference with subtyping and control-flow analysis \cite{}, or polyvariant flow analysis \cite{}.
+In control-flow analysis, expressions can be associated with a label, and each label can map to a set of abstract values,
+determined by learning subset constraints from the control-flow of the program. 
+Instead of labels, type inference with subtyping associate types  
+with expressions and learns subtyping constraints.
 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
add comment on tree grammars and recursive types

@@
 must be abolished so that there is no intrinsic weakest type other than a top type. 
-
 In MLSub and algebraic subtyping \cite{}, forgoes the datatype mechanism.
 Instead, it uses type inference to generates subtyping constraints and solve them, akin to unification in HM.
+
+Structures, when combined with loops, enable the construction of nested data (i.e. trees) 
+of arbitrary size. To ensure recursive functions on trees are error-free,
+systems have used regular tree grammars and recursive types to characterize spaces of trees.
+There has been plenty of research on reasoning over regular tree grammars/automata and recursive types \cite{}.
+Most relevant to our system is previous work on subtyping of recursive types \cite{}.
 

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
update
add positive substitution requirement

@@
-\documentclass[acmsmall]{acmart}
+\documentclass[table,dvipsnames,acmsmall]{acmart}
 % \documentclass[letterpaper]{llncs}
@@
 \usepackage[T1]{fontenc}
-\usepackage[dvipsnames]{xcolor}
+\usepackage[table,dvipsnames]{xcolor}
 \usepackage{array}
@@
 \usepackage{url}
-
-\usepackage[table]{xcolor}
 
@@
 \newtheorem{theorem}{Theorem}[section]
+\newtheorem{requirement}{Requirement}[section]
 \newtheorem{definition}{Definition}[section]
@@
 
-\title{Extrinsic Type Inference (DRAFT 2025-06-08)}
+\title{Extrinsic Type Inference (DRAFT 2025-06-09)}
 \author{Thomas Logan}
@@
 \newcommand{\examp}[1]{{\nopad \footnotesize 
-\vspace{-1em}
 \[
@@
 \]
-\vspace{-1em}
 }}
@@
 
-\TODO{add requirements on factoring, packing, polar sub and tidying, and negatble checks}
+\begin{requirement}
+  \label{req:positive_substitution_soundness}
+  \emph{positive substitution soundness}
+  \small
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \alpha \notin \textbf{ftv}(\tau[\alpha \slash \tau_r]^+)
+    } {
+      \delta \satisfies \tau_l \subtypes \tau_r
+      \implies
+      \delta\ \alpha \slash \tau_l \satisfies \tau \subtypes \tau[\alpha \slash \tau_r]^+
+    }
+  \end{mathpar}
+\end{requirement}
+
+\TODO{say what it means simply}
+This property is necessary so that our proof typing rules can use positive substitution
+as a proxy reasoning under the induction hypothesis.
+
+\TODO{add requirements on factoring, packing, and tidying, and negatble checks}
 
@@
 Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
-For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
-or an infinite number of
-other parametric types could be used, but none is weaker than all the rest.
+There are an infinite number of
+parametric types that could be used, but none is weaker than all the rest.
+
+\examp{
+  \J{forall 'a . 'a -> 'a}
+  \\
+  \J{forall 'a . 'a -> 'a list}
+  \\
+  \J{forall 'a . 'a -> 'a option}
+}
+
 Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to

### 2025-07-10 extrinsic_type_inference_paper/paper.tex
update

@@
 
-\TODO{say what it means simply}
+\noindent
+The requirement of positive substitution soundness says that
+if a variable \ms{\alpha} only occurs in positive positions of a type \ms{\tau},
+then the result of the substituting in another type \ms{\tau_r} is weaker than 
+any interpretation of the original type \ms{\tau} 
+if the replaced variable \ms{\alpha} is interpreted with a type \ms{\tau_l} 
+that is weaker than the subbed in type \ms{\tau_r}.
 This property is necessary so that our proof typing rules can use positive substitution
-as a proxy reasoning under the induction hypothesis.
+as a proxy reasoning under the induction hypothesis without violating soundness.
 
@@
 preserves the strengthened substance property from the induction hypothesis.
+
+\TODO{note how substance allows positive packing to place generated constraints into qualified types without violating soundness }
 

### 2025-07-10 extrinsic_type_inference_paper/paper.tex
update schema sequence duality correspondence

@@
 
-\TODO{add requirements on factoring, packing, and tidying, and negatble checks}
+\begin{requirement}
+  \label{req:schema_sequence_duality_correspondence}
+  \emph{schema sequence duality correspondence}
+  \tiny
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \alpha_\nu \downarrow \Pi_\nu \fallingdotseq \alpha_\mu \uparrow \Pi_\mu
+      \\
+      \delta \satisfies \Delta
+    } {
+      (
+      \forall \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
+      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\nu \implies
+      \exists \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \up
+      \\\\
+      (
+      \forall \delta' \qua
+      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \implies
+      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \implies
+      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \implies
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
+      )
+      )
+      \\\\
+      \iff
+      \\\\
+      (\forall e_a, e_r \qua
+      (
+      \exists \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
+      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\mu \up
+      (\forall \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \implies
+      \\\\
+      \exists \delta' \qua
+      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \up
+      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \up
+      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \up
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
+      )
+      )
+      \implies
+      e_r \cong e_f\J{(}e_a\J{)}
+      )
+    }
+  \end{mathpar}
+\end{requirement}
+
+\noindent
+The requirement of schema sequence duality has an ungodly amount of symbolic clutter,
+but the essence of its meaning is actually quite simple and natural.
+It says that a function inhabits all schemas in some 
+some collection if and only if for any input/ouput pair, if the pair inhabits some
+schema from the dual collection, then the function applied to the input
+gives the output. 
+The requirement is essential to ensure that rewriting inferred schemas
+to construct least fixed point types does not violate soundness.
+
+\noindent
+
+\TODO{add requirements on factoring}
+
+\TODO{add requirements on packing}
+
+\TODO{add requirements on tidying}
+
+\TODO{add requirements on negatble checks}
 

### 2025-07-10 extrinsic_type_inference_paper/paper.tex
add packing requirements
add completeness lemmas

@@
 
-\noindent
+\begin{requirement}
+  \label{req:positive packing correspondence}
+  \emph{positive packing correspondence}
+  \footnotesize
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \textbf{pack}^+(\Theta \entails \Pi) = \tau 
+      \\
+      \delta \satisfies \Delta
+      \\
+      \Theta \subseteq \textbf{dom}(\delta)
+    } {
+      (\forall  \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua 
+      (\Theta^\dagger, \Delta^\dagger, \tau^\dagger) \in \Pi \implies
+      (
+      \exists \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \backslash \Theta \up 
+      \\\\
+      (
+      \forall \delta' \qua 
+      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \emptyset \implies
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \implies
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e \hastype \tau^\dagger
+      )
+      ) 
+      \\\\
+      \iff
+      \\\\
+      \delta \satisfies e \hastype \tau
+    }
+  \end{mathpar}
+\end{requirement}
+
+\begin{requirement}
+  \label{req:negative packing correspondence}
+  \emph{negative packing correspondence}
+  \footnotesize
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \textbf{pack}^-(\Theta \entails \Pi) = \tau 
+      \\
+      \delta \satisfies \Delta
+      \\
+      \Theta \subseteq \textbf{dom}(\delta)
+    } {
+      (\exists  \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua 
+      (\Theta^\dagger, \Delta^\dagger, \tau^\dagger) \in \Pi \up
+      (
+      \forall \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \backslash \Theta \implies
+      \\\\
+      (
+      \exists \delta' \qua 
+      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \emptyset \up
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \up
+      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e \hastype \tau^\dagger
+      )
+      ) 
+      \\\\
+      \iff
+      \\\\
+      \delta \satisfies e \hastype \tau
+    }
+  \end{mathpar}
+\end{requirement}
+
+\noindent
+
+\begin{requirement}
+  \label{req:proof_subtyping_sequence_completeness_under_negatability}
+  \emph{proof subtyping sequence completeness under negatability} 
+  \small
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \Theta, \Delta \Vvdash \Omega
+      \\
+      \Theta, \Delta \entails \Omega \given \Theta', \Delta'
+      \\
+      \delta' \satisfies \Delta'
+    } {
+      \forall \delta^\dagger \qua
+      \delta' \oplus \delta^\dagger \satisfies \Omega \implies
+      \delta' \oplus \delta^\dagger \satisfies \Delta'
+    }
+  \end{mathpar}
+\end{requirement}
+
+\begin{requirement}
+  \label{req:proof subtyping completeness under negatability}
+  \emph{proof subtyping completeness under negatability}  
+  \small
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \Theta, \Delta \Vvdash \eta
+      \\
+      \Theta, \Delta \entails \tau \subtypes \eta \given \Theta', \Delta'
+      \\
+      \delta' \satisfies \Delta'
+    } {
+      \forall \delta^\dagger \qua
+      \delta' \oplus \delta^\dagger \satisfies \tau \subtypes \eta \implies
+      \delta' \oplus \delta^\dagger \satisfies \Delta'
+    }
+  \end{mathpar}
+\end{requirement}
+
+\begin{requirement}
+  \label{req:proof subtyping existential completeness under negatability}
+  \emph{proof subtyping existential completeness under negatability}  
+  \small
+  \begin{mathpar}
+    \inferrule {
+      \Theta, \Delta \Vvdash \eta
+      \\
+      \delta \satisfies \tau \subtypes \eta
+      \\
+      \delta \satisfies \Delta
+    } {
+      \exists \Theta', \Delta' \qua \Theta, \Delta \entails \tau \subtypes \eta \given \Theta', \Delta'
+    }
+  \end{mathpar}
+\end{requirement}
+
+\TODO{add requirements on tidying}
 
 \TODO{add requirements on factoring}
-
-\TODO{add requirements on packing}
-
-\TODO{add requirements on tidying}
-
-\TODO{add requirements on negatble checks}
 

### 2025-07-10 extrinsic_type_inference_paper/paper.tex
udpate

@@
 \begin{requirement}
-  \label{req:proof subtyping completeness under negatability}
-  \emph{proof subtyping completeness under negatability}  
+  \label{req:proof_subtyping_solution_completeness_under_negatability}
+  \emph{proof subtyping solution completeness under negatability}  
   \small
@@
 \begin{requirement}
-  \label{req:proof subtyping existential completeness under negatability}
+  \label{req:proof_subtyping_existential_completeness_under_negatability}
   \emph{proof subtyping existential completeness under negatability}  

### 2025-07-10 src/tapas/slim/analyzer.py
update diff case

@@ -2590,6 +2590,9 @@ SOLVABLE:
         else:
             return False
 
+    def is_guarded_typ(self, t : Typ) -> bool:
+        return isinstance(t, TUnit) or isinstance(t, TEntry)
+
     def is_negatable_typ(self, world : World, t : Typ) -> bool:
         return (
             (
@@ -2609,7 +2612,10 @@ SOLVABLE:
             # NOTE: compatible check only necessary for relational reasoning 
             # self.is_compatible(st.lower, st.upper)
             # if isinstance(st.upper, LeastFP) else
-            self.is_negatable_typ(world, st.upper) or (
+
+
+            # TODO: rethink the constraints
+            (self.is_negatable_typ(world, st.lower) and self.is_negatable_typ(world, st.upper)) or (
                 isinstance(st.lower, TVar) and
                 st.lower not in world.closedids and
                 all(
@@ -3075,21 +3081,18 @@ SOLVABLE:
             return self.solve(world, lower, universal_typ)
 
 
-
-
-
         #######################################
         #### Diff Introduction #############
         #######################################
 
-        elif isinstance(upper, Diff): 
-            if self.is_negatable_typ(world, upper.negation):
-                if not bool(self.solve(world, lower, upper.negation)):
-                    return self.solve(world, lower, upper.context)
-                else:
-                    return []
+        elif isinstance(upper, Diff) and self.is_guarded_typ(lower) and self.is_negatable_typ(world, upper.negation):
+            if not bool(self.solve(world, lower, upper.negation)):
+                return self.solve(world, lower, upper.context)
             else:
                 return []
+
+        elif isinstance(upper, Diff) and bool(self.solve(world, upper.negation, lower)) and not bool(self.solve(world, lower, upper.negation)) and self.is_negatable_typ(world, upper.negation):
+            return []
 
 
         #######################################
@@ -3118,6 +3121,7 @@ SOLVABLE:
                 else:
                     return self.solve(world, lower_body, upper)
                 # end-if
+
 
 
         #######################################
@@ -3240,8 +3244,6 @@ SOLVABLE:
                 return self.solve(world, lower.context, Unio(upper, lower.negation))
             else:
                 return []
-
-
 
 
         #######################################

### 2025-07-10 src/test_analysis_new.py

@@ -149,20 +149,74 @@ LFP [R] (<zero> @) | (<succ> R)
     )
     assert bool(worlds)
 
-def test_induction_even_is_nat_not_three():
+def test_induction_nat_is_even_false():
     worlds = solve_subtyping(f"""
-LFP [R] (<zero> @) | (<succ> <succ> R)
+LFP [R] (<zero> @) | (<succ> R)
     """, f"""
-(LFP [R] (<zero> @) | (<succ> R)) \\ (<succ> <succ> <succ> <zero> @)
+LFP [R] (<zero> @) | (<succ> <succ> R)
     """
     )
-    assert bool(worlds)
+    assert not bool(worlds)
+
+# def test_induction_even_is_nat_not_three():
+#     worlds = solve_subtyping(f"""
+# LFP [R] (<zero> @) | (<succ> <succ> R)
+#     """, f"""
+# (LFP [R] (<zero> @) | (<succ> R)) \\ (<succ> <succ> <succ> <zero> @)
+#     """
+#     )
+#     assert bool(worlds)
 
 def test_two_not_three():
     worlds = solve_subtyping(f"""
 (<succ> <succ> <zero> @)
     """, f"""
 TOP \\ (<succ> <succ> <succ> <zero> @)
+    """
+    )
+    assert bool(worlds)
+
+# def test_induction_even_is_not_three():
+#     worlds = solve_subtyping(f"""
+# LFP [R] (<zero> @) | (<succ> <succ> R)
+#     """, f"""
+# TOP \\ (<succ> <succ> <succ> <zero> @)
+#     """
+#     )
+#     assert bool(worlds)
+
+def test_union_is_not_zero_false():
+    worlds = solve_subtyping(f"""
+(<zero> @) | (<succ> <zero> @)
+    """, f"""
+TOP \\ (<zero> @)
+    """
+    )
+    assert not bool(worlds)
+
+def test_induction_even_is_not_zero_false():
+    worlds = solve_subtyping(f"""
+LFP [R] (<zero> @) | (<succ> <succ> R)
+    """, f"""
+TOP \\ (<zero> @)
+    """
+    )
+    assert not bool(worlds)
+
+def test_induction_nat_is_not_one_false():
+    worlds = solve_subtyping(f"""
+LFP [R] (<zero> @) | (<succ> R)
+    """, f"""
+TOP \\ (<succ> <zero> @)
+    """
+    )
+    assert not bool(worlds)
+
[…]

### 2025-07-11 extrinsic_type_inference_paper/paper.tex
update
add factorization soundness
add fresh check
update diff case
udpate overview examples

@@
 \examp{
-  \J{def foo = ? f => (f(<succ> <succ> <zero> @), f(<true> @))}
+  \J{def foo = [f => (f(succ;succ;zero;@), f(true;@))]}
 }
@@
 
-\Par{Structural Generalization} 
+\Par{Path Sensitivity} 
 Functions on lists can be generic, in that they may be independent of the item values, as in \code{repeat}.
@@
 of code at the call site. 
-Untyped languages or extrinsically typed languages can bypass this rigidity found in intrinsically typed by specializing in multiple ways.
+Untyped languages or extrinsically typed languages can bypass this rigidity found in 
+intrinsically typed languages by mixing the specialized procedures into a single generalized procedures.
 
 \examp{
-  ...
-  \\
   \J{def genCmp = [a,b => } 
@@
   \\
+  \J{...} 
+}
+
+\noindent
+Extrinsic type inference allows this mixing by inferring types locally for each path,
+such that the type inferred from one path is not required for another.
+As such, a generalized comparison procedure may be constructed and the
+programmer can use the same sort function no matter if its inputs are scalars or vectors.
+
+\examp{
   \J{def genSort : (TOP}
@@
   \\
-  \J{) = sort(genCmp) in}
-  \\
-  \J{...} 
-}
-
-\noindent
-By combining the scalar and lexicographic comparison procedures into a single comparison
-procedure, a generalized sort function can be used to work on either scalars or vectors,
-thereby removing the burden of choosing the proper comparison function at the call site.
+  \J{) = sort(genCmp)}
+}
+
+\noindent
+This particular form of type inference, known as occurrence typing \cite{},
+has already been studied and used to great effect in type inference for Python and Javascript.
+Additionally, the path sensitivity of type inference obviates the need 
+for destruction in pattern matching.  
+For example, the \code{repeat} function can be rewritten.
+
+\examp{
+  \J{def repeat = [x => loop([self => [n =>}
+  \\
+  \I\J{n |> [}
+  \\
+  \I\I \J{[zero;\_ => nil;(n.zero)]}
+  \\
+  \I\I \J{[succ;\_ => cons;(x,self(n.succ))]}
+  \\
+  \I \J{]}
+  \\
+  \J{])]}
+}
+
+\noindent
+Thus, programs in Python and Javascript, which only having conditional branching,
+and no pattern matching, can rely on path sensitivity to simulate pattern matching.
+
+\Par{Pruning}
+
+\TODO{...}
+
+\Par{Streaming}
+
+\TODO{...}
 
@@
       \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
+      \\
+      \alpha \notin \textbf{ftv}(\Delta')
       \\\\
@@
       \Theta, \Delta \entails 
-      \tau \subtypes \eta \given \Theta', \Delta'
+      \tau_l \subtypes \eta \given \Theta', \Delta'
     } {
@@
   \emph{schema sequence duality correspondence}
-  \tiny
+  \footnotesize
   \nopad
@@
       \delta \satisfies \Delta
+      \\
+      \alpha_\nu \notin \textbf{ftv}(\Delta)
+      \\
+      \alpha_\mu \notin \textbf{ftv}(\Delta)
     } {
@@
 
-\TODO{add requirements on tidying}
-
-\TODO{add requirements on factoring}
+\begin{requirement}
+  \label{req:factorization_soundness}
+  \emph{factorization soundness}  
+  \small
+  \begin{mathpar}
+    \inferrule {
+      \Vdash \J{LFP[}\alpha\J{]}\tau \subtypes l \mapsto \tau'
+    } {
+      \forall e' \qua
+      e' \hastype \tau' 
+      \iff 
+      (\exists e \qua e \J{.} l \cong e' \up \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau)
+    }
+  \end{mathpar}
+\end{requirement}
 
@@
       \J{]) in ...}
+    }
+    & \pass & \fail \\
+
+    \sline
+
+    \sdc. &
+    \cb{
+      \J{[x => halve(double(x))]}
     }

### 2025-07-11 src/tapas/slim/analyzer.py

@@ -3564,19 +3564,21 @@ result:
             for inner_world in self.solver.solve(outer_world, body_typ, Imp(self_typ, Imp(in_typ, out_typ)))
             for local_constraints in [inner_world.constraints.difference(world.constraints)]
             for left_typ in [self.solver.interpret_id_weakest(local_constraints, in_typ.id)] 
-            for pre_right_typ in [self.solver.interpret_id_strongest(local_constraints, out_typ.id)]
-            for (right_typ, right_constraints) in [
-                # TODO: handle recursion nested in function
-                [
-                    (right_body, right_constraints)
-                    for renaming in [self.solver.make_renaming(pre_right_typ.ids)]
-                    for right_constraints in [sub_constraints(renaming, pre_right_typ.constraints)]
-                    for right_body in [sub_typ(renaming, pre_right_typ.body)]
-                ][-1]
-                if isinstance(pre_right_typ, All) else
-                (pre_right_typ, [])
-            ]
-            for w in self.solver.solve_multi(inner_world, right_constraints)
+            for right_typ in [self.solver.interpret_id_strongest(local_constraints, out_typ.id)]
+            # for pre_right_typ in [self.solver.interpret_id_strongest(local_constraints, out_typ.id)]
+            # for (right_typ, right_constraints) in [
+            #     # TODO: handle recursion nested in function
+            #     [
+            #         (right_body, right_constraints)
+            #         for renaming in [self.solver.make_renaming(pre_right_typ.ids)]
+            #         for right_constraints in [sub_constraints(renaming, pre_right_typ.constraints)]
+            #         for right_body in [sub_typ(renaming, pre_right_typ.body)]
+            #     ][-1]
+            #     if isinstance(pre_right_typ, All) else
+            #     (pre_right_typ, [])
+            # ]
+            # for w in self.solver.solve_multi(inner_world, right_constraints)
+            for w in [inner_world]
         ]
         is_single = (len(inner_schemas) == 1) 
         # foreignids = foreignids.union(
@@ -3680,6 +3682,10 @@ rel_typ:
         assert param_upper_bound 
         print(f"RAW: {param_upper_bound.body}")
         if is_single and is_unstructured(self.solver.simplify_typ(param_upper_bound.body)):
+            param_typ = param_upper_bound 
+            return_typ = find_factor_from_label(rel_typ, "tail")
+            assert param_typ and return_typ
+            return Result(pid, outer_world, Imp(param_typ, return_typ))
             ################################################
             # Stream Type 
             ##################################################

### 2025-07-11 src/test_analysis_expressivity.py

@@ -1086,7 +1086,7 @@ if __name__ == '__main__':
     ##########################
     # test_max()
     # test_max_app()
-    # test_succ_stream()
+    test_succ_stream()
     # test_single_case_loop()
     # test_typing_G8A()
     # test_something_subtypes_non_decreasing_lfp()

### 2025-07-11 extrinsic_type_inference_paper/paper.tex
add notes on recursive types

@@
 \maketitle
+
+\TODO{
+I think the main novelty of my type inference design beyond previous systems is that it can infer recursive types extrinsically and then reason inductively with those types.
+Do you think that the title should reflect that?
+}
+\TODO{
+e.g. “Extrinsic Reconstruction of Recursive Types”
+}
+\TODO{
+e.g. “Inductive Subtyping over Reconstructed Recursive Types”
+}
+
+\TODO{don't need to call out extrinsic, because recursive types are the final hurdle in type inference}
+
+\TODO{if recursive types can be reconstructed than so can everything else}
 

### 2025-07-11 extrinsic_type_inference_paper/paper.tex
add some notes on induction

@@
 \TODO{if recursive types can be reconstructed than so can everything else}
+
+\TODO{inductive reasoning, in the general sense, consists of learning sufficient conditions P that enable deriving Q }
+
+\TODO{reasoning by cases consists of deriving P implies Q 
+by ensuresing P' implies Q for every P' that implie P
+}
+
+\TODO{mathematical induction is a particular form of reasoning by cases
+where P is recursively defines, so P' resembles it, and for any P' that resembles P,
+we can assume that P' -> Q', where Q' resembles Q.
+}
 

### 2025-07-13 extrinsic_type_inference_paper/paper.tex
update construction of LFP for streams

@@
 
+\TODO{compare inductive subtyping to tree interpolation}
+
+\TODO{is this something we should/can prove?}
+
+\TODO{note how CHC uses predefined constraints, while this system learns constraints from applications}
+
+\TODO{other systems (using CHC) can construct recursive horn clauses from programs, but not local constraints, right?}
+
+\TODO{that is the language of recursive types is used as both the constraint language and the target language}
+
 \TODO{understand the capabilities and limitations of MLSub better}

### 2025-07-13 src/tapas/slim/analyzer.py

@@ -3580,18 +3580,24 @@ result:
             for w in self.solver.solve_multi(inner_world, right_constraints)
             # for w in [inner_world]
         ]
-        is_single = (len(inner_schemas) == 1) 
-        # foreignids = foreignids.union(
-        #     pset(
-        #         id
-        #         for (inner_world, left_typ, right_typ) in inner_schemas
-        #         for id in extract_free_vars_from_typ(s(), left_typ).intersection(
-        #             extract_free_vars_from_typ(s(), right_typ)
-        #         )
-        #     )
-        #     if is_single else
-        #     pset()
-        # )
+
+        left_unstructured_vars = pset(
+            left_typ.id
+            for (inner_world, left_typ, right_typ) in inner_schemas
+            if (len(inner_schemas) == 1)
+            if isinstance(left_typ, TVar) 
+        )
+        foreignids = foreignids.union(left_unstructured_vars)
+
+        # TODO: 
+        # - extrude param type variables, and replace positive occurences in result 
+        # - replace pattern with extrusions.  
+        # - add constraint that the weakened inferred param type subtypes the extrusion
+
+
+        # LFP[W] EXI[T] (<succ> T <: W):T <: <succ> T' 
+        # LFP[W] EXI[T] T | <succ> W  <: <succ> T' 
+        #  EXI[T] T | <succ> <succ> T'  <: <succ> T' 
 
         induc_body = Bot()
         for i, (inner_world, left_typ, right_typ) in enumerate(reversed(inner_schemas)):
@@ -3672,38 +3678,37 @@ rel_typ:
         # return Result(pid, outer_world, All(tuple([param_var.id]), tuple([param_constraint]), Imp(param_typ, return_typ)))
         ################################################
 
-        def is_unstructured(t : Typ) -> bool:
-            return isinstance(t, TVar) or (
-                isinstance(t, Exi) and
-                is_unstructured(t.body)
-            )
+        def weaken_to_decreasining_lfp(t : LeastFP) -> LeastFP:
+            # NOTE: weaken from an increasing LFP to a decreasing LFP
+            content = self.solver.simplify_typ(t.body)  
+            if isinstance(content, Exi) and isinstance(content.body, TVar):
+                choices = [
+                    self.solver.sub_polar_typ(True, st.lower, content.body.id, TVar(t.id))
+                    for st in content.constraints
+                    if isinstance(st.upper, TVar) and st.upper.id == t.id
+                ] + [content.body]
[…]

### 2025-07-13 src/test_analysis_expressivity.py

@@ -36,7 +36,10 @@ def test_succ_stream():
     ######################
     # expected: T ->  LFP[R] @ -> <pair> ((LFP[W] EXI[U](<succ> U <: W): U) * R)    
     # TODO: what is the rationale for why it's save to flip around the constraint into an LFP?
-    # does (LFP[W] EXI[U](<succ> U <: W): U) <: (LFP[W] T | <succ> W) ?
+    # does (LFP[W] EXI[](<succ> T <: W): T) <: (LFP[W] T | <succ> W) ?
+    # does (LFP[W] T | <succ> W) <: (LFP[W] EXI[](<succ> T <: W): T) ?
+    ######
+    # does (T | <succ> (LFP[W] EXI[](<succ> T <: W): T)) <: (LFP[W] EXI[](<succ> T <: W): T) ?
     ######################
     # expected: T ->  LFP[R] @ -> <pair> ((LFP[W] T | <succ> W) * R)    
     ######################
@@ -1092,6 +1095,29 @@ def test_typing_structures_6():
     assert infer_typ(code)
 
 
+def test_subtyping_increasing_recursion():
+    # INCOMPLETENESS
+    # the subtyping never halts with a contradiction, so should be accepted
+    # since this is actually rejected, but should be accepted,
+    # this form must NOT be allowed in negations of subtyping.
+    worlds = solve_subtyping(f"""
+@ 
+    """, f"""
+LFP [R] EXI [T] (<succ> T <: R) : T 
+    """
+    )
+    assert bool(worlds)
+
+def test_subtyping_decreasing_recursion():
+    worlds = solve_subtyping(f"""
+<succ> <nil> @ 
+    """, f"""
+LFP [R] <nil> @ | <succ> R 
+    """
+    )
+    assert bool(worlds)
+
+
 ###############################################################
 ###############################################################
 
@@ -1100,11 +1126,12 @@ if __name__ == '__main__':
     ##########################
     # test_max()
     # test_max_app()
-    test_succ_stream()
+    # test_succ_stream()
     # test_single_case_loop()
     # test_typing_G8A()
     # test_something_subtypes_non_decreasing_lfp()
     # test_something_subtypes_increasing_lfp()
+    test_subtyping_increasing_recursion()
 
 
 #######################################################################

### 2025-07-13 src/test_analysis_rejections.py

@@ -35,15 +35,6 @@ LFP [R] (<zero> @) | (<succ> <succ> R)
     assert not bool(worlds)
 
 
-def test_diff():
-    worlds = solve_subtyping(f"""
-LFP [R] R -> @ 
-    """, f"""
-TOP \\ LFP [R] EXI [T] (T <: R) : R -> @ 
-    """
-    )
-    assert not bool(worlds)
-
 def test_union_is_not_zero():
     worlds = solve_subtyping(f"""
 (<zero> @) | (<succ> <zero> @)

### 2025-07-14 extrinsic_type_inference_paper/paper.tex
update diff case and negatable and subtractable checks

@@
     \\
-    \com{negatable} &
+    \com{subtractable} &
     \nonterm{\eta} & \is 
     \begin{array}[t]{@{}l}
-      \J{EXI[}\Theta \J{]} \Omega \J{:} \eta \sep 
+      \J{EXI[}\Theta \J{]} \eta \sep 
       \rho
@@
 A difference \ms{\tau \typdiff \eta} is a space without inhabitation by certain structures.
-The subtracted part is restricted to a negatable form,
+The subtracted part is restricted to a subtractable form,
 which is a form that our system can decide the inhabitation
@@
     \\
-    \com{negatable type seq} &
+    \com{subtractable type seq} &
     \nonterm{\Xi} & \is \epsilon \sep \Xi\ \eta 
@@
 path's pattern does not subtype the type of subsequent paths' patterns. 
-The sequence of negatable types \ms{\Xi} can be ignored, as it is only
+The sequence of subtractable types \ms{\Xi} can be ignored, as it is only
 exposed for the sake of being used recursively.
@@
     \inferrule {
+      \TODO{Update proof, deriving that lower does not intersect with subtraction}
+      \\\\
       \Theta, \Delta \Vvdash \eta
-      \\
-      \Theta, \Delta \entails 
-      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
       \\
@@
       \tau_l \subtypes \eta \given \Theta', \Delta'
+      \\\\
+      \Theta, \Delta \Vvdash \eta \subtypes  \tau_l
+      \\
+      \nexists \Theta', \Delta' \qua  
+      \Theta, \Delta \entails 
+      \eta \subtypes \tau_l \given \Theta', \Delta'
+      \\\\
+      \Theta, \Delta \entails 
+      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
     } {
@@
 and fails for the subtracted part.
-Additionally, it requires that the subtracted part is negatable
+Additionally, it requires that the subtracted part is subtractable 
 \ms{\Theta, \Delta \Vvdash \eta}, 

### 2025-07-14 src/tapas/slim/analyzer.py

@@ -224,7 +224,7 @@ class Inter:
 @dataclass(frozen=True, eq=True)
 class Diff:
     context : Typ 
-    negation : Typ # NOTE:, restrict to a tag/field pattern that is easy decide anti-unification
+    subtraction : Typ # NOTE:, restrict to a tag/field pattern that is easy decide anti-unification
 
 @dataclass(frozen=True, eq=True)
 class Imp:
@@ -376,7 +376,7 @@ def to_nameless(bound_ids : tuple[str, ...], typ : Typ) -> NL:
     elif isinstance(typ, Inter):
         return InterNL(to_nameless(bound_ids, typ.left), to_nameless(bound_ids, typ.right))
     elif isinstance(typ, Diff):
-        return DiffNL(to_nameless(bound_ids, typ.context), to_nameless(bound_ids, typ.negation))
+        return DiffNL(to_nameless(bound_ids, typ.context), to_nameless(bound_ids, typ.subtraction))
     elif isinstance(typ, Imp):
         return ImpNL(to_nameless(bound_ids, typ.antec), to_nameless(bound_ids, typ.consq))
     elif isinstance(typ, Exi):
@@ -492,7 +492,7 @@ def concretize_typ(typ : Typ) -> str:
             else:
                 plate_entry = ([control.left,control.right], lambda left, right : f"({left} & {right})")  
         elif isinstance(control, Diff):
-            plate_entry = ([control.context,control.negation], lambda context,negation : f"({context} \\ {negation})")  
+            plate_entry = ([control.context,control.subtraction], lambda context,negation : f"({context} \\ {negation})")  
         elif isinstance(control, Exi):
             constraints = concretize_constraints(control.constraints, inline=True)
             ids = concretize_ids(control.ids)
@@ -674,7 +674,7 @@ def diff_well_formed(diff : Diff) -> bool:
     '''
     restriction to avoid dealing with negating divergence (which would need to soundly fail under even negs, soundly pass under odd negs)
     '''
-    return negation_well_formed(diff.negation)
+    return negation_well_formed(diff.subtraction)
 
 
 def make_pair_typ(left : Typ, right : Typ) -> Typ:
@@ -730,7 +730,7 @@ def project_typ_recurse(t : Typ, path : Sequence[str]) -> Optional[Typ]:
         return None
     elif isinstance(t, Diff):
         context = project_typ_recurse(t.context, path)
-        negation = project_typ_recurse(t.negation, path)
[…]

### 2025-07-14 src/test_analysis_expressivity.py

@@ -1117,7 +1117,6 @@ LFP [R] <nil> @ | <succ> R
     )
     assert bool(worlds)
 
-
 ###############################################################
 ###############################################################
 
@@ -1131,7 +1130,7 @@ if __name__ == '__main__':
     # test_typing_G8A()
     # test_something_subtypes_non_decreasing_lfp()
     # test_something_subtypes_increasing_lfp()
-    test_subtyping_increasing_recursion()
+    test_two_not_three()
 
 
 #######################################################################

### 2025-07-14 src/test_analysis_rejections.py

@@ -64,9 +64,9 @@ TOP \\ (<succ> <zero> @)
 
 
 ###############################################################
-###############################################################
 
 if __name__ == '__main__':
+    test_induction_nat_is_not_one()
     pass
     ##########################
 

### 2025-07-17 src/test_analysis_expressivity.py
update stream and diff examples

@@ -50,7 +50,7 @@ def test_succ_stream():
     code = f"""
 loop([self => [seed => 
     [@ =>
-        pair;(seed, (self)(succ;seed))
+        (seed, self(succ;seed))
     ]
 ]])
     """
@@ -1130,7 +1130,8 @@ if __name__ == '__main__':
     # test_typing_G8A()
     # test_something_subtypes_non_decreasing_lfp()
     # test_something_subtypes_increasing_lfp()
-    test_two_not_three()
+    # test_two_not_three()
+    test_succ_stream()
 
 
 #######################################################################

### 2025-07-17 doc/extrinsic_type_inference_paper/paper.tex
update stream and diff examples
update notes on title and how recursive types fit in

--- 2025-07-14 extrinsic_type_inference_paper/paper.tex
@@
 
-\title{Extrinsic Type Inference (DRAFT 2025-06-09)}
+\title{Fully Extrinsic Type Inference (DRAFT 2025-06-17)}
 \author{Thomas Logan}
@@
 
-\TODO{
-I think the main novelty of my type inference design beyond previous systems is that it can infer recursive types extrinsically and then reason inductively with those types.
-Do you think that the title should reflect that?
-}
-\TODO{
-e.g. “Extrinsic Reconstruction of Recursive Types”
-}
-\TODO{
-e.g. “Inductive Subtyping over Reconstructed Recursive Types”
-}
-
-\TODO{don't need to call out extrinsic, because recursive types are the final hurdle in type inference}
+\TODO{Mention how reconstruction of recursive types is the last hurdle to fully extrinsic types}
 
@@
   \\
-  \J{])]}
+  \J{]])]}
 }
@@
 \TODO{...}
+
+\examp{
+  \J{loop([self => [x =>}
+  \\
+  \I\J{[@ => (x, self(succ;x))]}
+  \\
+  \J{]]}
+}
+
+\TODO{...}
+
+\examp{
+  \J{ALL[T] T -> @ ->}
+  \\
+  \I\J{LFP[U] (LFP[R] T | <succ> R) * (@ -> U)}
+}
+
+% (ALL[G009] (G009 -> (LFP[G007] (EXI[G010] (G010 <: G007)  : (@ -> ((LFP[G007] (<succ> G007 | G009)) * G010))))))
+
+\Par{Subtraction}
+
+\TODO{...}
+
+\examp{
+  \J{LFP[R] (<zero> @) | (<succ> <succ> R)}
+  \\
+  \subtypes
+  \\
+  \J{TOP\typdiff (<succ> <succ> <succ> <zero> @)}
+}
 

### 2025-07-20 doc/extrinsic_type_inference_paper/paper.tex
add note on terminology of plastic vs meta variable

@@
 \maketitle
+
+\TODO{lean uses the term "metavariable" for our concept of "plastic variable". Should we conform to Lean's terminology?}
+
+\TODO{Metavariable term: the variable is meta in the sense that its interpretation is chosen by the metaprogram.
+Even though it's actually still a variable in the object language.}
+
+\TODO{We use the term "plastic variable" to avoid the potentially misleading terminology of "meta variable"}
 

### 2025-07-20 doc/extrinsic_type_inference_paper/paper.tex
add notes on variable terminology
update to make locale the term for the triple of skolems, constraints, and a type

@@
 \TODO{We use the term "plastic variable" to avoid the potentially misleading terminology of "meta variable"}
+
+\TODO{Other options are "unknown", "hole", "logic variable", "placeholder", "localetic variable"}
+
+\TODO{replace "plastic" with "placeholder"}
 
@@
     \\
-    \com{schema seq} &
+    \com{locale seq} &
     \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
@@
 to infer as many types as possible
-and consolidates the results into a sequence of schemas \ms{\Pi}.
+and consolidates the results into a sequence of locales \ms{\Pi}.
 It must consolidate all of the inferred types in order to
@@
 \ms{\textbf{pack}^+(... \entails \Pi) = \tau}
-to rewrite each schema into a type,
+to rewrite each locale into a type,
 such that the skolems and plastics are converted to
@@
 First it leverages typing consolidation \ms{\Theta, \Delta, \Gamma \entails e \hastype \Pi}
-to infer as many possible schemas for the source expression \ms{e}.
-Then it packs all the schemas together and ensures that the packed type
+to infer as many possible locales for the source expression \ms{e}.
+Then it packs all the locales together and ensures that the packed type
 subtypes the type annotation.
@@
 form compatible with looping and 
-to collect as many type schemas of its paths as possible.  
-It leverage schema sequence duality 
+to collect as many type locales of its paths as possible.  
+It leverage locale sequence duality 
 \ms{
@@
   \alpha_\mu \uparrow \Pi_\mu
-}, which rewrites the implicitly co-recursive path schemas into
-implicitly recursive pair schemas. 
+}, which rewrites the implicitly co-recursive path locales into
+implicitly recursive pair locales. 
 Then, it leverages negative packing \ms{\textbf{pack}^-(... \entails \Pi_\mu) = \tau_\mu},
@@
 \begin{requirement}
-  \label{req:schema_sequence_duality_correspondence}
-  \emph{schema sequence duality correspondence}
+  \label{req:locale_sequence_duality_correspondence}
+  \emph{locale sequence duality correspondence}
   \footnotesize
@@
 \noindent
-The requirement of schema sequence duality has an ungodly amount of symbolic clutter,
+The requirement of locale sequence duality has an ungodly amount of symbolic clutter,
 but the essence of its meaning is actually quite simple and natural.
-It says that a function inhabits all schemas in some 
+It says that a function inhabits all locales in some 
 some collection if and only if for any input/ouput pair, if the pair inhabits some
-schema from the dual collection, then the function applied to the input
+locale from the dual collection, then the function applied to the input
 gives the output. 
-The requirement is essential to ensure that rewriting inferred schemas
+The requirement is essential to ensure that rewriting inferred locales
 to construct least fixed point types does not violate soundness.

### 2025-07-20 doc/extrinsic_type_inference_paper/paper.tex
switch to term placeholder

@@
 
-\title{Fully Extrinsic Type Inference (DRAFT 2025-06-17)}
+\title{Fully Extrinsic Type Inference (DRAFT 2025-06-20)}
 \author{Thomas Logan}
@@
 
-\TODO{lean uses the term "metavariable" for our concept of "plastic variable". Should we conform to Lean's terminology?}
+\TODO{lean uses the term "metavariable" for our concept of "placeholder variable". Should we conform to Lean's terminology?}
 
@@
 Even though it's actually still a variable in the object language.}
-
-\TODO{We use the term "plastic variable" to avoid the potentially misleading terminology of "meta variable"}
-
-\TODO{Other options are "unknown", "hole", "logic variable", "placeholder", "localetic variable"}
-
-\TODO{replace "plastic" with "placeholder"}
 
@@
 subsequent sub-problems.
-If a variable is not in the set of skolems, then we consider it to be plastic,
+If a variable is not in the set of skolems, then we consider it to be placeholder,
 in that it can be constrained in any way that is consistent with the other subtyping constraints
@@
 to rewrite each locale into a type,
-such that the skolems and plastics are converted to
+such that the skolems and placeholders are converted to
 existential and universal bindings,
@@
 \begin{definition} 
-  \label{def:proof_subtyping_plastic_elimination}
-  \emph{proof subtyping (plastic elimination)}
+  \label{def:proof_subtyping_placeholder_elimination}
+  \emph{proof subtyping (placeholder elimination)}
   \hfill
@@
 \noindent
-For plastic elimination \ms{\alpha \subtypes \tau} where \ms{\alpha \notin \Theta}, 
-it ensures that the supertype is consistent with all lower bounds of the plastic,
+For placeholder elimination \ms{\alpha \subtypes \tau} where \ms{\alpha \notin \Theta}, 
+it ensures that the supertype is consistent with all lower bounds of the placeholder,
 solutions against those lower bounds. Then it learns that the supertype is an upper bound 
-of the plastic.
+of the placeholder.
 
 \begin{definition} 
-  \label{def:proof_subtyping_plastic_introduction}
-  \emph{proof subtyping (plastic introduction)}
+  \label{def:proof_subtyping_placeholder_introduction}
+  \emph{proof subtyping (placeholder introduction)}
   \hfill
@@
 \noindent
-Plastic introduction is the dual of plastic elimination, so its rule
+Placeholder introduction is the dual of placeholder elimination, so its rule
 relies on analogous techniques.
@@
 there are two possibilities.
-In the first scenario, the skolem has a plastic as a lower bound.
+In the first scenario, the skolem has a placeholder as a lower bound.
 If effect, this enables learning lower bounds on the skolem,
-so the solver uses the same techniques as it did for plastic elimination.
-I the second scenario, there is a lower bound that's not a plastic,
+so the solver uses the same techniques as it did for placeholder elimination.
+I the second scenario, there is a lower bound that's not a placeholder,
 so the solver ensures that the skolem introduction is weaker
@@
 was found. In practice, the solver avoids checking the subtype against the lower bound, 
-if the lower bound is a plastic, as that would cause the solver to diverge due to
-a mutual dependence between plastic elimination and skolem elimination.
+if the lower bound is a placeholder, as that would cause the solver to diverge due to
+a mutual dependence between placeholder elimination and skolem elimination.
 

### 2025-07-29 doc/extrinsic_type_inference_paper/paper.tex
update
update model typing rules for pattern matching; note negative occurence
update rules with negative position restriction

@@
 
+\TODO{factor out definitions common to both dynamic and static systems into basic system}
+
+\TODO{flip records, functions and other collections to associate right (matching Lean design)}
+
 \noindent
@@
     \inferrule {
-      \exists e' .\ f\J{(}v\J{)} \rightsquigarrow e'
+      \nexists e' .\ f\J{(}v\J{)} \rightsquigarrow e'
       \\
@@
 \begin{definition}
+  \label{def:valuing}
+  \emph{valuing}
+  \hfill
+  \small
+  \boxed{\Downarrow e}
+  \nopad
+  \begin{mathpar}
+    \inferrule { 
+      \TODO{...}
+    } {
+      \Downarrow e 
+    } 
+  \end{mathpar}
+\end{definition}
+
+\TODO{create a seperate relation restricted negations of model typing}
+
+\TODO{use antec rules to compensate for non-self reference in negative position}
+
+\TODO{consider renaming to dynamic typing}
+
+\begin{definition}
   \label{def:model_typing}
@@
     \inferrule { 
-      \J{<}l \J{>} v \in g
-      \\
-      \delta \satisfies v \hastype \tau 
-      \\
-      \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
-    } {
-      \delta \satisfies g \hastype \J{<} l \J{>} \tau
-    }
-
-    \inferrule { 
-      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
-    } {
-      \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
-    }
-
-    \inferrule { 
-      \forall \sigma \qua
-      \delta \satisfies p[\sigma] \hastype \tau_l
-      \implies
-      \delta  \satisfies e[\sigma] \hastype \tau_r
-      \\\\
-      \forall \tau_l', \tau_r' \qua
-      \delta \satisfies f \hastype \tau_l' \J{->} \tau_r'
-      \implies
-      \delta \satisfies \tau_l \subtypes \tau_l'
-      \implies
-      \delta \satisfies \tau_r' \subtypes \tau_r
-    } {
-      \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
+      \alpha \slash \tau \in \delta
+      \\
+      \delta \satisfies e \hastype \tau
+    } {
+      \delta \satisfies e \hastype \alpha 
+    }
+
+    \inferrule { 
+      \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
+    } {
+      \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
     }
@@
       \delta \satisfies e \hastype \tau_l
-      \\
-      \neg\ (\delta \satisfies e \hastype \tau_r)
-    } {
-      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
+    } {
+      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
+    }
+
+    \inferrule { 
+      \delta \satisfies e \hastype \tau_r
+    } {
+      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
+    }
+
+    \inferrule { 
+      \textbf{dom}(\delta') \subseteq \Theta
+      \\
+      \delta \oplus \delta' \satisfies \Delta 
+      \\
+      \delta \oplus \delta' \satisfies e \hastype \tau
+    } {
+      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
     }
@@
       \delta \satisfies e \hastype \tau_l \J{\&} \tau_r
-    }
-
-    \inferrule { 
-      \delta \satisfies e \hastype \tau_l
-    } {
-      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
-    }
-
-    \inferrule { 
-      \delta \satisfies e \hastype \tau_r
-    } {
-      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
     }
@@
       \textbf{dom}(\delta') \subseteq \Theta \implies
-      \delta \oplus \delta' \satisfies \Delta \implies 
+      \delta \oplus \delta' \VDash \Delta \implies 
       \delta \oplus \delta' \satisfies e \hastype \tau
@@
     \inferrule { 
+      \J{<}l \J{>} v \in g
+      \\
+      \delta \satisfies v \hastype \tau 
+      \\
+      \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
+    } {
+      \delta \satisfies g \hastype \J{<} l \J{>} \tau
+    }
+
+    \inferrule { 
+      \TODO{diff antec}
+      \\\\
+      p \Uparrow \eta 
+      \\
+      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
+    } {
+      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \typdiff \eta \J{->} \tau_r
+    }
+
+    \inferrule { 
+      \TODO{needs separate inter antec}
+      \\\\
+      \forall \sigma \qua 
+      \delta \VDash p[\sigma] \hastype \tau_l
+      \implies
+      \delta  \satisfies e[\sigma] \hastype \tau_r
+    } {
+      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
+    }
+
+    \inferrule { 
+      \TODO{inter antec left}
+      \\\\
+      \delta \satisfies e \hastype \tau_l \J{->} \tau
+    } {
+      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
+    }
+
+    \inferrule { 
+      \TODO{inter antec right}
+      \\\\
+      \delta \satisfies e \hastype \tau_r \J{->} \tau
+    } {
+      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
+    }
+
+    \inferrule { 
+      \TODO{all antec}
+      \\\\
       \textbf{dom}(\delta') \subseteq \Theta 
@@
       \\
-      \delta \oplus \delta' \satisfies e \hastype \tau
-    } {
-      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
-    }
-
-    \inferrule { 
-      \alpha \slash \tau \in \delta
-      \\
-      \delta \satisfies e \hastype \tau
-    } {
-      \delta \satisfies e \hastype \alpha 
+      \delta \oplus \delta' \satisfies e \hastype \tau_l \J{->} \tau_r
+    } {
+      \delta \satisfies e \hastype (\J{ALL[}\Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
+    }
+
+    \inferrule { 
+      \TODO{union antec}
+      \\\\
+      \delta \satisfies e \hastype \tau_l \J{->} \tau
+      \\
+      \delta \satisfies e \hastype \tau_r \J{->} \tau
+    } {
+      \delta \satisfies e \hastype (\tau_l \J{|} \tau_r) \J{->} \tau
+    }
+
+    \inferrule { 
+      \TODO{exi antec}
+      \\\\
+      \forall \delta' \qua 
+      \textbf{dom}(\delta') \subseteq \Theta
+      \implies
+      \delta \oplus \delta' \VDash \Delta 
+      \implies
+      \delta \satisfies e \hastype \tau_l \J{->} \tau_r
+    } {
+      \delta \satisfies e \hastype (\J{EXI[} \Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
+    }
+
+    \inferrule { 
+      \TODO{lfp antec}
+      \\\\
+      \alpha \notin \textbf{ftv}(\tau_l[\alpha\slash\J{@}]^+)
+      \\
+      \delta\ \alpha \slash (\J{LFP[}\alpha \J{]} \tau_l) \satisfies e \hastype \tau_l \J{->} \tau_r
+    } {
+      \delta \satisfies e \hastype (\J{LFP[}\alpha \J{]} \tau_l) \J{->} \tau_r
+    }
+
+    \inferrule { 
+      \delta \satisfies e \hastype \tau_l
+      \\
+      \neg\ (\delta \satisfies e \hastype \tau_r)
+    } {
+      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
     }
@@
 
-    \inferrule { 
-      \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
-    } {
-      \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
-    }
+    % \inferrule { 
+    %   \TODO{remove unflipped}
+    %   \\\\
+    %   \delta \satisfies f \hastype \tau_l \J{->} \tau_r
+    % } {
+    %   \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{remove unflipped}
+    %   \\\\
+    %   \forall \sigma \qua
+    %   \delta \satisfies p[\sigma] \hastype \tau_l
+    %   \implies
+    %   \delta  \satisfies e[\sigma] \hastype \tau_r
+    %   \\\\
+    %   \forall \tau_l', \tau_r' \qua
+    %   \delta \satisfies f \hastype \tau_l' \J{->} \tau_r'
+    %   \implies
+    %   \delta \satisfies \tau_l \subtypes \tau_l'
+    %   \implies
+    %   \delta \satisfies \tau_r' \subtypes \tau_r
+    % } {
+    %   \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \delta \satisfies f \hastype \tau_l \J{->} \tau_r
+    %   \\\\
+    %   \forall \tau_l', \tau_r' \qua
+    %   \delta \satisfies \J{[}p \J{=>} e \J{]} \hastype \tau_l' \J{->} \tau_r'
+    %   \implies
+    %   \delta \satisfies \tau_l \subtypes \tau_l'
+    %   \implies
+    %   \delta \satisfies \tau_r' \subtypes \tau_r
+    % } {
+    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{how to deal with recursive negative occurence?}
+    %   \\\\
+    %   \forall v \qua 
+    %   \delta \satisfies v \hastype \tau_l
+    %   \implies
+    %   \exists \sigma \qua
+    %   v \simeq p \given \sigma 
+    %   \up
+    %   \delta  \satisfies e[\sigma] \hastype \tau_r
+    % } {
+    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{how to deal with recursive negative occurence?}
+    %   \\\\
+    %   \TODO{intersection elim is built into pattern matching}
+    %   \\\\
+    %   \forall v \qua 
+    %   \delta \VDash v \hastype \rho
+    %   \implies
+    %   \exists \sigma \qua
+    %   v \simeq p \given \sigma 
+    %   \up
+    %   \delta  \satisfies e[\sigma] \hastype \tau_r
+    % } {
+    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \rho \J{->} \tau_r
+    % }
 

### 2025-07-30 doc/extrinsic_type_inference_paper/paper.tex
begin updating model typing rules

@@
   \small
-  \boxed{\Downarrow e}
+  \boxed{\downharpoonleft e}
   \nopad
@@
     } {
-      \Downarrow e 
+      \downharpoonleft 
+      e 
     } 
@@
 
+\begin{definition} 
+  \label{def:type_sub_fold}
+  Type Substitution Fold 
+  \hfill
+  \small
+  \boxed{
+    \tau [\alpha \slash \tau']^n
+  }
+  \\
+  \[
+  \begin{array}[t]{rclcl}
+      \tau [\alpha \slash \tau']^0 
+      &=& 
+      \tau' 
+      \\
+      \tau [\alpha \slash \tau']^{(n + 1)} 
+      &=& 
+      \tau [\alpha \slash \tau[\alpha \slash \tau']^n]
+  \end{array}
+  \]
+\end{definition}
+
 \TODO{create a seperate relation restricted negations of model typing}
@@
 \TODO{consider renaming to dynamic typing}
+
+\TODO{completely redo model typing with well founded recursion}
+
+\TODO{define monotonicity/antitonicity/polarity check}
+
+\TODO{create recursive definition of least fixed point check}
+
+\TODO{use separate typing for interpreting variables as finite types}
+
+\TODO{use non-syntactic description of implication}
+
+\TODO{remove implication antecedent rules}
+
+\TODO{lift progression rule out to highest level for monotonic recursion}
+
+\begin{definition}
+  \label{def:model_core_typing}
+  \emph{model core typing}
+  \hfill
+  \small
+  \boxed{\delta \VDash e \hastype \tau}
+  \nopad
+  \begin{mathpar}
+    \inferrule { 
+      \alpha \wr^+ \tau
+      \\
+      \exists n \qua e \hastype \tau[\alpha \slash \J{BOT}]^n 
+    } {
+      \delta \VDash e \hastype \J{LFP[}\alpha\J{]}\tau 
+    } 
+  \end{mathpar}
+\end{definition}
+
+\begin{definition}
+  \label{def:model_finite_typing}
+  \emph{model finite typing}
+  \hfill
+  \small
+  \boxed{e \hastype \tau}
+  \nopad
+  \begin{mathpar}
+  \end{mathpar}
+\end{definition}
 
@@
   \begin{mathpar}
-    \inferrule { 
-      e \rightsquigarrow e'
-      \\
-      \delta \satisfies e' \hastype \tau 
-    } {
-      \delta \satisfies e \hastype \tau 
-    }
-
-    \inferrule { 
-      \alpha \slash \tau \in \delta
-      \\
-      \delta \satisfies e \hastype \tau
-    } {
-      \delta \satisfies e \hastype \alpha 
-    }
-
-    \inferrule { 
-      \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
-    } {
-      \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
-    }
-
-    \inferrule { 
-      \delta \satisfies e \hastype \tau_l
-    } {
-      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
-    }
-
-    \inferrule { 
-      \delta \satisfies e \hastype \tau_r
-    } {
-      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
-    }
-
-    \inferrule { 
-      \textbf{dom}(\delta') \subseteq \Theta
-      \\
-      \delta \oplus \delta' \satisfies \Delta 
-      \\
-      \delta \oplus \delta' \satisfies e \hastype \tau
-    } {
-      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
-    }
-
-    \inferrule { 
-      \delta \satisfies e \hastype \tau_l
-      \\
-      \delta \satisfies e \hastype \tau_r
-    } {
-      \delta \satisfies e \hastype \tau_l \J{\&} \tau_r
-    }
-
-    \inferrule { 
-      \forall \delta' \qua
-      \textbf{dom}(\delta') \subseteq \Theta \implies
-      \delta \oplus \delta' \VDash \Delta \implies 
-      \delta \oplus \delta' \satisfies e \hastype \tau
-      \\\\
-      \exists \delta' \qua 
-      \textbf{dom}(\delta') \subseteq \Theta \up 
-      \delta \oplus \delta' \satisfies \Delta
-    } {
-      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
-    }
-
-    \inferrule { 
-      \J{<}l \J{>} v \in g
-      \\
-      \delta \satisfies v \hastype \tau 
-      \\
-      \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
-    } {
-      \delta \satisfies g \hastype \J{<} l \J{>} \tau
-    }
-
-    \inferrule { 
-      \TODO{diff antec}
-      \\\\
-      p \Uparrow \eta 
-      \\
-      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
-    } {
-      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \typdiff \eta \J{->} \tau_r
-    }
-
-    \inferrule { 
-      \TODO{needs separate inter antec}
-      \\\\
-      \forall \sigma \qua 
-      \delta \VDash p[\sigma] \hastype \tau_l
-      \implies
-      \delta  \satisfies e[\sigma] \hastype \tau_r
-    } {
-      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
-    }
-
-    \inferrule { 
-      \TODO{inter antec left}
-      \\\\
-      \delta \satisfies e \hastype \tau_l \J{->} \tau
-    } {
-      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
-    }
-
-    \inferrule { 
-      \TODO{inter antec right}
-      \\\\
-      \delta \satisfies e \hastype \tau_r \J{->} \tau
-    } {
-      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
-    }
-
-    \inferrule { 
-      \TODO{all antec}
-      \\\\
-      \textbf{dom}(\delta') \subseteq \Theta 
-      \\
-      \delta \oplus \delta' \satisfies \Delta 
-      \\
-      \delta \oplus \delta' \satisfies e \hastype \tau_l \J{->} \tau_r
-    } {
-      \delta \satisfies e \hastype (\J{ALL[}\Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
-    }
-
-    \inferrule { 
-      \TODO{union antec}
-      \\\\
-      \delta \satisfies e \hastype \tau_l \J{->} \tau
-      \\
-      \delta \satisfies e \hastype \tau_r \J{->} \tau
-    } {
-      \delta \satisfies e \hastype (\tau_l \J{|} \tau_r) \J{->} \tau
-    }
-
-    \inferrule { 
-      \TODO{exi antec}
-      \\\\
-      \forall \delta' \qua 
-      \textbf{dom}(\delta') \subseteq \Theta
-      \implies
-      \delta \oplus \delta' \VDash \Delta 
-      \implies
-      \delta \satisfies e \hastype \tau_l \J{->} \tau_r
-    } {
-      \delta \satisfies e \hastype (\J{EXI[} \Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
-    }
-
-    \inferrule { 
-      \TODO{lfp antec}
-      \\\\
-      \alpha \notin \textbf{ftv}(\tau_l[\alpha\slash\J{@}]^+)
-      \\
-      \delta\ \alpha \slash (\J{LFP[}\alpha \J{]} \tau_l) \satisfies e \hastype \tau_l \J{->} \tau_r
-    } {
-      \delta \satisfies e \hastype (\J{LFP[}\alpha \J{]} \tau_l) \J{->} \tau_r
-    }
-
-    \inferrule { 
-      \delta \satisfies e \hastype \tau_l
-      \\
-      \neg\ (\delta \satisfies e \hastype \tau_r)
-    } {
-      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
-    }
-
-    \inferrule { 
-    } {
-      \delta \satisfies \J{@} \hastype \J{@}
-    }
+    % \inferrule { 
+    %   e \rightsquigarrow e'
+    %   \\
+    %   \delta \satisfies e' \hastype \tau 
+    % } {
+    %   \delta \satisfies e \hastype \tau 
+    % }
+
+    % \inferrule { 
+    %   \alpha \slash \tau \in \delta
+    %   \\
+    %   \delta \satisfies e \hastype \tau
+    % } {
+    %   \delta \satisfies e \hastype \alpha 
+    % }
+
+    % \inferrule { 
+    %   \TODO{alt def}
+    %   \\\\
+    %   \forall \tau' \qua  
+    %   \delta \satisfies \tau[\alpha\slash\tau'] \subtypes \tau'
+    %   \implies
+    %   \delta \satisfies e \hastype \tau'
+    % } {
+    %   \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
+    % }
+
+    % \inferrule { 
+    %   \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
+    % } {
+    %   \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
+    % }
+
+    % \inferrule { 
+    %   \delta \satisfies e \hastype \tau_l
+    % } {
+    %   \delta \satisfies e \hastype \tau_l \J{|} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \delta \satisfies e \hastype \tau_r
+    % } {
+    %   \delta \satisfies e \hastype \tau_l \J{|} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \textbf{dom}(\delta') \subseteq \Theta
+    %   \\
+    %   \delta \oplus \delta' \satisfies \Delta 
+    %   \\
+    %   \delta \oplus \delta' \satisfies e \hastype \tau
+    % } {
+    %   \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
+    % }
+
+    % \inferrule { 
+    %   \delta \satisfies e \hastype \tau_l
+    %   \\
+    %   \delta \satisfies e \hastype \tau_r
+    % } {
+    %   \delta \satisfies e \hastype \tau_l \J{\&} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \forall \delta' \qua
+    %   \textbf{dom}(\delta') \subseteq \Theta \implies
+    %   \delta \oplus \delta' \VDash \Delta \implies 
+    %   \delta \oplus \delta' \satisfies e \hastype \tau
+    %   \\\\
+    %   \exists \delta' \qua 
+    %   \textbf{dom}(\delta') \subseteq \Theta \up 
+    %   \delta \oplus \delta' \satisfies \Delta
+    % } {
+    %   \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
+    % }
+
+    % \inferrule { 
+    %   \J{<}l \J{>} v \in g
+    %   \\
+    %   \delta \satisfies v \hastype \tau 
+    %   \\
+    %   \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
+    % } {
+    %   \delta \satisfies g \hastype \J{<} l \J{>} \tau
+    % }
+
+    % \inferrule { 
+    %   \TODO{diff antec}
+    %   \\\\
+    %   p \Uparrow \eta 
+    %   \\
+    %   \delta \satisfies f \hastype \tau_l \J{->} \tau_r
+    % } {
+    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \typdiff \eta \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{needs separate inter antec}
+    %   \\\\
+    %   \forall \sigma \qua 
+    %   \delta \VDash p[\sigma] \hastype \tau_l
+    %   \implies
+    %   \delta  \satisfies e[\sigma] \hastype \tau_r
+    % } {
+    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{inter antec left}
+    %   \\\\
+    %   \delta \satisfies e \hastype \tau_l \J{->} \tau
+    % } {
+    %   \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
+    % }
+
+    % \inferrule { 
+    %   \TODO{inter antec right}
+    %   \\\\
+    %   \delta \satisfies e \hastype \tau_r \J{->} \tau
+    % } {
+    %   \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
+    % }
+
+    % \inferrule { 
+    %   \TODO{all antec}
+    %   \\\\
+    %   \textbf{dom}(\delta') \subseteq \Theta 
+    %   \\
+    %   \delta \oplus \delta' \satisfies \Delta 
+    %   \\
+    %   \delta \oplus \delta' \satisfies e \hastype \tau_l \J{->} \tau_r
+    % } {
+    %   \delta \satisfies e \hastype (\J{ALL[}\Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{union antec}
+    %   \\\\
+    %   \delta \satisfies e \hastype \tau_l \J{->} \tau
+    %   \\
+    %   \delta \satisfies e \hastype \tau_r \J{->} \tau
+    % } {
+    %   \delta \satisfies e \hastype (\tau_l \J{|} \tau_r) \J{->} \tau
+    % }
+
+    % \inferrule { 
+    %   \TODO{exi antec}
+    %   \\\\
+    %   \forall \delta' \qua 
+    %   \textbf{dom}(\delta') \subseteq \Theta
+    %   \implies
+    %   \delta \oplus \delta' \VDash \Delta 
+    %   \implies
+    %   \delta \satisfies e \hastype \tau_l \J{->} \tau_r
+    % } {
+    %   \delta \satisfies e \hastype (\J{EXI[} \Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \TODO{lfp antec}
+    %   \\\\
+    %   \alpha \notin \textbf{ftv}(\tau_l[\alpha\slash\J{@}]^+)
+    %   \\
+    %   \delta\ \alpha \slash (\J{LFP[}\alpha \J{]} \tau_l) \satisfies e \hastype \tau_l \J{->} \tau_r
+    % } {
+    %   \delta \satisfies e \hastype (\J{LFP[}\alpha \J{]} \tau_l) \J{->} \tau_r
+    % }
+
+    % \inferrule { 
+    %   \delta \satisfies e \hastype \tau_l
+    %   \\
+    %   \neg\ (\delta \satisfies e \hastype \tau_r)
+    % } {
+    %   \delta \satisfies e \hastype \tau_l \typdiff \tau_r
+    % }
+
+    % \inferrule { 
+    % } {
+    %   \delta \satisfies \J{@} \hastype \J{@}
+    % }
 

### 2025-07-30 src/test_analysis_expressivity.py

@@ -25,6 +25,14 @@ from tapas.slim.language import analyze, infer_typ, parse_typ, solve_subtyping
 from tapas.slim import exprlib as el, typlib as tl
 from tapas.slim.exprlib import ctx 
 
+def test_inf_loop():
+    code = f"""
+loop([self => 
+    [x => self(x)]
+])
+    """
+    assert infer_typ(code) 
+
 
 def test_succ_stream():
     ######################
@@ -1131,7 +1139,8 @@ if __name__ == '__main__':
     # test_something_subtypes_non_decreasing_lfp()
     # test_something_subtypes_increasing_lfp()
     # test_two_not_three()
-    test_succ_stream()
+    # test_succ_stream()
+    test_inf_loop()
 
 
 #######################################################################

### 2025-08-08 doc/extrinsic_type_inference_paper/paper.tex
update model typing definition in paper
update model typing lfp rule in paper
update model typing
update termination proof for dynamic typing
update notes
update static/proof subtyping in lean
update lfp elim rules in lean
update diff intro rule in lean
udpate proof subtyping rules in lean

@@
 \newcommand{\pattype}{\ \lozenge\ }
+\newcommand{\restricted}{\ \blacklozenge\ }
 \newcommand{\liftfun}{:}
@@
 
+% \begin{definition}
+%   \label{def:model_progress_typing}
+%   \emph{model progressive typing}
+%   \hfill
+%   \small
+%   \boxed{\delta \satisfies e \hastype^* \tau}
+%   \nopad
+%   \begin{mathpar}
+%     \inferrule { 
+%       \downharpoonleft e
+%       \up
+%       \delta \satisfies e \hastype \tau 
+%     } {
+%       \delta \satisfies e \hastype^* \tau 
+%     }
+
+%     \inferrule { 
+%       e \rightsquigarrow e'
+%       \\
+%       \delta \satisfies e' \hastype^* \tau 
+%     } {
+%       \delta \satisfies e \hastype^* \tau 
+%     } 
+%   \end{mathpar}
+
+% \end{definition}
+
 \begin{definition}
-  \label{def:model_core_typing}
-  \emph{model core typing}
+  \label{def:model_typing}
+  \emph{model typing}
   \hfill
   \small
-  \boxed{\delta \VDash e \hastype \tau}
-  \nopad
-  \begin{mathpar}
-    \inferrule { 
+  \boxed{\delta \satisfies e \hastype \tau}
+  \nopad
+  \[
+  \begin{array}[t]{rclcl}
+      \delta \satisfies e \hastype \J{@}
+      &\iff& 
+      e \rightsquigarrow^* \J{@}
+
+      \\
+
+      \delta \satisfies e \hastype \J{<} l \J{>} \tau
+      &\iff& 
+      \delta \satisfies e\J{.}l \hastype \tau
+
+      \\
+
+      \delta \satisfies e \hastype \tau_l \J{->} \tau_r
+      &\iff& 
+      \forall e' \qua
+      \delta \satisfies e' \hastype \tau_l 
+      \implies
+      \delta \satisfies e\J{(}e'\J{)} \hastype \tau_r
+
+      \\
+
+      \delta \satisfies e \hastype \tau_l \J{|} \tau_r 
+      &\iff& 
+      \delta \satisfies e \hastype \tau_l
+      \down
+      \delta \satisfies e \hastype \tau_r
+
+      \\
+
+      \delta \satisfies e \hastype \tau_l \J{\&} \tau_r 
+      &\iff& 
+      \delta \satisfies e \hastype \tau_l
+      \up
+      \delta \satisfies e \hastype \tau_r
+
+      \\
+
+      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
+      &\iff& 
+      \delta \satisfies e \hastype \tau_l
+      \up
+      \delta \not\satisfies e \hastype \tau_r
+
+      \\
+
+      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
+      &\iff& 
+      \exists \delta' \qua
+      \textbf{dom}(\delta') \subseteq \Theta
+      \up
+      \delta \oplus \delta' \satisfies \Delta 
+      \up
+      \delta \oplus \delta' \satisfies e \hastype \tau
+
+      \\
+
+      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
+      &\iff& 
+      \forall \delta' \qua
+      \textbf{dom}(\delta') \subseteq \Theta
+      \implies
+      \delta \oplus \delta' \satisfies \Delta 
+      \implies
+      \delta \oplus \delta' \satisfies e \hastype \tau
+
+      \\
+
+      \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau 
+      &\iff& 
       \alpha \wr^+ \tau
-      \\
-      \exists n \qua e \hastype \tau[\alpha \slash \J{BOT}]^n 
-    } {
-      \delta \VDash e \hastype \J{LFP[}\alpha\J{]}\tau 
-    } 
-  \end{mathpar}
+      \up
+      \exists n \qua \VDash e \hastype (\tau[\alpha \slash \J{BOT}]^n)[\delta]
+
+      \\
+
+      \delta \satisfies e \hastype \alpha 
+      &\iff& 
+      \alpha \slash \tau \in \delta
+      \up
+      \VDash e \hastype \tau
+  \end{array}
+  \]
 \end{definition}
@@
   \small
-  \boxed{e \hastype \tau}
-  \nopad
-  \begin{mathpar}
-  \end{mathpar}
+  \boxed{\VDash e \hastype \tau}
+  \nopad
+  \[
+  \begin{array}[t]{rclcl}
+
+      \VDash e \hastype \J{@}
+      &\iff& 
+      e \rightsquigarrow^* \J{@}
+
+      \\
+
+      \VDash e \hastype \J{<} l \J{>} \tau
+      &\iff& 
+      \VDash e\J{.}l \hastype \tau
+
+      \\
+
+      \VDash e \hastype \tau_l \J{->} \tau_r
+      &\iff& 
+      \forall e' \qua
+      \VDash e' \hastype \tau_l 
+      \implies
+      \VDash e\J{(}e'\J{)} \hastype \tau_r
+
+      \\
+
+      \VDash e \hastype \tau_l \J{|} \tau_r 
+      &\iff& 
+      \VDash e \hastype \tau_l
+      \down
+      \VDash e \hastype \tau_r
+
+      \\
+
+      \VDash e \hastype \tau_l \J{\&} \tau_r 
+      &\iff& 
+      \VDash e \hastype \tau_l
+      \up
+      \VDash e \hastype \tau_r
+
+      \\
+
+      \VDash e \hastype \tau_l \typdiff \tau_r
+      &\iff& 
+      \VDash e \hastype \tau_l
+      \up
+      \not\VDash e \hastype \tau_r
+  \end{array}
+  \]
 \end{definition}
 
-\begin{definition}
-  \label{def:model_typing}
-  \emph{model typing}
-  \hfill
-  \small
-  \boxed{\delta \satisfies e \hastype \tau}
-  \nopad
-  \begin{mathpar}
-    % \inferrule { 
-    %   e \rightsquigarrow e'
-    %   \\
-    %   \delta \satisfies e' \hastype \tau 
-    % } {
-    %   \delta \satisfies e \hastype \tau 
-    % }
-
-    % \inferrule { 
-    %   \alpha \slash \tau \in \delta
-    %   \\
-    %   \delta \satisfies e \hastype \tau
-    % } {
-    %   \delta \satisfies e \hastype \alpha 
-    % }
-
+\TODO{note that the semantics allows progam to get stuck under vacuously inhabited types}
+
+% \begin{definition}
+%   \label{def:model_scratch_typing}
+%   \emph{model scratch typing}
+%   \hfill
+%   \small
+%   \boxed{\delta \satisfies e \hastype \tau}
+%   \nopad
+%   \begin{mathpar}
     % \inferrule { 
@@
 
-  \end{mathpar}
-\end{definition}
+%   \end{mathpar}
+% \end{definition}
 
@@
       \\
-      \Theta, \Delta \given \Omega \given \Theta', \Delta'
+      \Theta, \Delta \given \Omega \entails \Theta', \Delta'
       \\
@@
       \\\\
-      (\forall \alpha' \qua \tau' = \alpha' \implies \alpha' \in \Theta) 
+      (\forall \alpha' \qua \alpha' = \tau' \implies \alpha' \in \Theta) 
       \\
@@
     \inferrule {
-      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_r' 
-      \\
-      \Theta, \Delta \entails \tau_r' \subtypes \tau_r \given \Theta', \Delta' 
+      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_f 
+      \\
+      \Theta, \Delta \entails \tau_f \subtypes \tau_r \given \Theta', \Delta' 
     } {
@@
     \inferrule {
-      \tau_l[\alpha \slash \tau_r]^+ = \tau^\dagger
+      \alpha \wr^+ \tau_l
       \\
       \Theta, \Delta \entails 
-      \tau^\dagger \subtypes \tau_r \given 
+      \tau_l[\alpha \slash \tau_r] \subtypes \tau_r \given 
       \Theta', \Delta'
-      \\
-      \alpha \not\in \textbf{ftv}(\tau^\dagger)
-      \\
-      \alpha \not\in \textbf{ftv}(\tau_r)
     } {
@@
   \small
-  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
+  \boxed{\Theta, \Delta \entails \tau \subtypes \tau_l \typdiff \tau_r \given \Theta', \Delta'}
   \nopad
@@
       \\\\
-      \Theta, \Delta \Vvdash \eta
-      \\
-      \nexists \Theta', \Delta' \qua  
+      \TODO{in lean, prove that subtyping is equivalent to decidable definition under restricted subtyping form}
+      \\\\
+      \epsilon \pattype \tau_r
+      \\
+      \Theta, \Delta \restricted \tau_r \subtypes  \tau
+      \\\\
+      \Theta, \Delta \not\Vvdash \tau \subtypes \tau_r
+      \\
+      \Theta, \Delta \not\Vvdash \tau_r \subtypes \tau
+      \\
+      \Theta, \Delta \entails \tau \subtypes \tau_l \given \Theta', \Delta' 
+    } {
       \Theta, \Delta \entails 
-      \tau_l \subtypes \eta \given \Theta', \Delta'
-      \\\\
-      \Theta, \Delta \Vvdash \eta \subtypes  \tau_l
-      \\
-      \nexists \Theta', \Delta' \qua  
-      \Theta, \Delta \entails 
-      \eta \subtypes \tau_l \given \Theta', \Delta'
-      \\\\
-      \Theta, \Delta \entails 
-      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
-    } {
-      \Theta, \Delta \entails 
-      \tau_l \subtypes \tau_r \typdiff \eta
+      \tau \subtypes \tau_l \typdiff \tau_r
       \given \Theta', \Delta'
@@
     \inferrule {
-      \Theta, \Delta \entails \tau_l \circlearrowleft \J{LFP[} \alpha \J{]} \tau_r
+      \Theta, \Delta \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
       \\
@@
     \inferrule {
-      \alpha \Vdash \tau_r \supertypes \tau_r'
-      \\
       \Theta, \Delta \entails
-      \tau_l \subtypes \tau_r' \given \Theta', \Delta'
+      \tau_l \subtypes \textbf{drop}(\alpha \entails \tau_r) \given \Theta', \Delta'
     } {
@@
 \ms{
-\tau_l \circlearrowleft \J{LFP[} \alpha \J{]} \tau_r
+... \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
 } and then solve the inflated introduction.
@@
   \small
-  \boxed{\Theta, \Delta \entails \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'}
+  \boxed{\Theta, \Delta \entails \tau_l \typdiff \tau_r \subtypes \tau \given \Theta', \Delta'}
   \nopad
@@
       \Theta, \Delta \entails 
-      \tau_l \subtypes \eta \J{|} \tau_r \given \Theta', \Delta'
+      \tau_l \subtypes \tau_r \J{|} \tau \given \Theta', \Delta'
     } {
       \Theta, \Delta \entails 
-      \tau_l \typdiff \eta \subtypes \tau_r \given \Theta', \Delta'
+      \tau_l \typdiff \tau_r \subtypes \tau \given \Theta', \Delta'
     }
@@
     \inferrule {
-      \Vdash \tau_{l}\J{\&}\tau_{r} \subtypes \tau_x \rightarrow \tau_y 
-      \\\\
-      \Theta, \Delta \entails \tau_x \J{->} \tau_y  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
+      \Theta, \Delta \entails \textbf{merge}(\tau_{l}\J{\&}\tau_{r}) \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
     } {

### 2025-08-08 doc/extrinsic_type_inference_paper/paper.tex
update notation for factoring column from lfp

@@
 
-\newcommand{\factorsinto}{\Vdash}
-
 \newcommand{\allsafe}{\ \Re\checkmark}
@@
 \newcommand{\restricted}{\ \blacklozenge\ }
+\newcommand{\dentails}{\Vdash}
 \newcommand{\liftfun}{:}
@@
       \\\\
-      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{left} \mapsto \tau_l
-      \\
-      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{right} \mapsto \tau_r
+      \textbf{factor}(\alpha' \entails \tau' \cdot \J{left}) = \tau_l
+      \\
+      \textbf{factor}(\alpha' \entails \tau' \cdot \J{right}) = \tau_r
     } {
@@
     \inferrule {
-      \Theta, \Delta \Vvdash \Omega
+      \Theta, \Delta \restricted \Omega
       \\
@@
     \inferrule {
-      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_f 
-      \\
-      \Theta, \Delta \entails \tau_f \subtypes \tau_r \given \Theta', \Delta' 
+      \Theta, \Delta \entails 
+      \textbf{factor}(\alpha \entails \tau_l \cdot l) 
+      \subtypes 
+      \tau_r \given \Theta', \Delta' 
     } {
@@
       \\\\
-      \Theta, \Delta \not\Vvdash \tau \subtypes \tau_r
-      \\
-      \Theta, \Delta \not\Vvdash \tau_r \subtypes \tau
+      \Theta, \Delta \not\dentails \tau \subtypes \tau_r
+      \\
+      \Theta, \Delta \not\dentails \tau_r \subtypes \tau
       \\
@@
 Alternatively, it can strengthen the least fixed point
-with union drop \ms{\alpha \Vdash \tau_r \supertypes \tau_r'},
+with union drop \ms{\textbf{drop}(\alpha entails \tau_r) = \tau_r'},
 which drops any parts of the union containing the self reference.
@@
 path type antecedent, yet could be stronger than union of all the path type antecedents.    
-As such, the solver may leverage function type factorization \ms{
-\Vdash \tau_l \J{\&} \tau_r \subtypes \tau_x \rightarrow \tau_y 
+As such, the solver may leverage path type merging \ms{
+\textbf{merge}(\tau_l \J{\&} \tau_r)
 } to weaken a function type
@@
     \inferrule {
-      \Vdash \J{LFP[}\alpha\J{]}\tau \subtypes l \mapsto \tau'
+      \textbf{factor}(\alpha \entails \tau \cdot l) = \tau'
     } {

### 2025-08-13 doc/extrinsic_type_inference_paper/paper.tex
begin typing rules in lean
update typing rules in lean
update rule for streaming recursion
update streaming rule in paper and lean
factor out loop subtyping
update notation for inflatable check
update loop subtyping rules in paper and lean
update terminology with zone

@@
 \newcommand{\given}{\dashv}
-\newcommand{\with}{\ \diamond\ }
-\newcommand{\notfree}{\ \#\ }
-\newcommand{\consis}{\ \star}
-\newcommand{\safe}{\ \checkmark}
-\newcommand{\relational}{\ \Re}
-
-% \newcommand{\cmark}{\ding{51}}%
-% \newcommand{\xmark}{\ding{55}}%
+
+% inflatable
+\newcommand{\inflatable}{\star}
+
+\newcommand{\cmark}{\ding{51}}%
+\newcommand{\xmark}{\ding{55}}%
 \newcommand{\pass}{\text{\scriptsize \color{Green} \ding{51}}}
@@
 \newcommand{\fail}{\text{\scriptsize \color{red} \ding{55}}}
-
-\newcommand{\unique}{\ \dagger}
-
-\newcommand{\allsafe}{\ \Re\checkmark}
 
@@
     \\
-    \com{locale seq} &
+    \com{zone seq} &
     \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
@@
     \inferrule {
-      x:\tau \in \Gamma
-    } {
-      \Theta, \Delta, \Gamma \entails x \hastype \tau \given \Theta, \Delta 
+    } {
+      \Theta, \Delta, \Gamma \entails x \hastype \Gamma(x)  \given \Theta, \Delta 
     }
@@
       \Theta, \Delta, \Gamma \entails r \hastype \tau \given \Theta', \Delta'  
-      \\\\
+      \\
       \Theta', \Delta', \Gamma \entails e \hastype \tau' \given \Theta'', \Delta'' 
@@
       \exists \Xi \qua \Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi 
+      \\
+      \textbf{pack}^+(\textbf{ftv}(\Delta) \entails \Pi) = \tau 
+    } {
+      \Theta, \Delta, \Gamma \entails f \hastype \tau \given \Theta, \Delta
+    }
+
+    \inferrule {
+      \Theta, \Delta, \Gamma \entails e_f \hastype \tau_f \given \Theta', \Delta'
+      \\
+      \Theta', \Delta', \Gamma \entails e_a \hastype \tau_a \given \Theta'', \Delta''
+      \\
+      \Theta'', \Delta'' \entails \tau_f \subtypes \tau_a\J{->}\alpha \given \Theta''', \Delta'''
+    } {
+      \Theta, \Delta, \Gamma \entails e_f\J{(}e_a \J{)} \hastype \alpha \given \Theta''', \Delta'''
+    }
+
+    \inferrule {
+      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
+      \\
+      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
+      \\
+      \textbf{ftv}(\Delta') \entails \alpha \circlearrowleft \Pi \subtypes \tau'
+    } {
+      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
+      \hastype \tau' 
+      \given \Theta', \Delta'
+    }
+
+    \inferrule {
+      \textbf{ftv}(\tau_a) \subseteq \emptyset
+      \\
+      \Theta, \Delta, \Gamma \entails e \hastype \Pi
+      \\
+      \textbf{pack}^+(\textbf{ftv}(\Delta) \entails \Pi) = \tau_e
+      \\
+      \exists \Theta', \Delta' \qua \Theta, \Delta \entails \tau_e \subtypes \tau_a \given \Theta', \Delta'
+    } {
+      \Theta, \Delta, \Gamma \entails e \J{ as } \tau_a \hastype \tau_a \given \Theta', \Delta'
+    }
+  \end{mathpar}
+\end{definition}
+
+\begin{definition} 
+  \label{def:loop_subtyping}
+  \emph{loop subtyping}
+  \hfill 
+  \small
+  \boxed{\Theta^\dagger \entails \alpha \circlearrowleft \Pi \subtypes \tau}
+  \nopad
+  \begin{mathpar}
+    \inferrule {
+      \textbf{invert}(\alpha \entails \Pi) = \Pi'
+      \\
+      \textbf{pack}^-(\alpha\ \Theta^\dagger \entails \Pi') = \tau'
       \\\\
-      \textbf{pack}^+(\Theta \cup \textbf{ftv}(\Delta) \cup \textbf{ftv}(\Gamma) \entails \Pi) = \tau 
-    } {
-      \Theta, \Delta, \Gamma \entails f \hastype \tau \given \Theta, \Delta
-    }
-
-    \inferrule {
-      \Theta, \Delta, \Gamma \entails e \hastype \tau \given \Theta', \Delta' 
+      \textbf{factor}(\alpha \entails \tau' \cdot \J{left}) = \tau_l
+      \\
+      \textbf{factor}(\alpha \entails \tau' \cdot \J{right}) = \tau_r
+    } {
+      \Theta^\dagger \entails 
+      \alpha \circlearrowleft \Pi 
+      \subtypes
+      \J{LFP[} \alpha \J{]} \tau_l \J{->} \J{LFP[} \alpha \J{]} \tau_r
+    }
+
+    \inferrule {
+      \alpha \neq \alpha_l
       \\\\
-      \Theta', \Delta' \entails 
-      \tau \subtypes  \J{<} l \J{>} \alpha 
-      \given \Theta'', \Delta'' 
-    } {
-      \Theta, \Delta, \Gamma \entails e\J{.}l \hastype \alpha \given \Theta'', \Delta''
-    }
-
-    \inferrule {
-      \Theta, \Delta, \Gamma \entails e_f \hastype \tau_f \given \Theta', \Delta'
-      \\
-      \Theta', \Delta', \Gamma \entails e_a \hastype \tau_a \given \Theta'', \Delta''
+      \textbf{invert}(\alpha \entails \Delta) = \Delta' 
+      \\
+      \textbf{pack}^-(\alpha\ \alpha_l\ \Theta^\dagger \entails \Theta, \Delta', \alpha_l \J{*} \tau_r) = \tau'
       \\\\
-      \Theta'', \Delta'', \tau_f \subtypes \tau_a\J{->}\alpha \given \Theta''', \Delta'''
-    } {
-      \Theta, \Delta, \Gamma \entails e_f\J{(}e_a \J{)} \hastype \alpha \given \Theta''', \Delta'''
-    }
-
-    \inferrule {
-      \textbf{ftv}(\tau_a) \subseteq \emptyset
-      \\
-      \Theta, \Delta, \Gamma \entails e \hastype \Pi
-      \\
-      \textbf{pack}^+(\Theta \cup \textbf{ftv}(\Delta) \cup \textbf{ftv}(\Gamma) \entails \Pi) = \tau 
+      \textbf{factor}(\alpha \entails \tau' \cdot \J{left}) = \tau_l
+      \\
+      \textbf{factor}(\alpha \entails \tau' \cdot \J{right}) = \tau_r'
       \\\\
-      \exists \Theta^\dagger, \Delta^\dagger \qua \Theta', \Delta' \entails \tau \subtypes \tau_a \given \Theta^\dagger, \Delta^\dagger
-      \\
-      \Theta', \Delta', \Gamma, x:\tau_a \entails e' \hastype \tau' \given \Theta'', \Delta''
-    } {
-      \Theta, \Delta, \Gamma \entails \J{def } x \J{:} \tau_a \J{ = } e \J{ in } e' \hastype \tau' \given \Theta'', \Delta''
-    }
-
-    \inferrule {
-      \TODO{modify with case for non-recursive parameter type}
-      \\\\
-      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
-      \\
-      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
-      \\
-      \alpha \notin \textbf{ftv}(\Delta')
-      \\\\
-      \alpha \downarrow \Pi
-      \fallingdotseq 
-      \alpha' \uparrow \Pi'
-      \\
-      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha' \entails \Pi') = \tau'
-      \\\\
-      \textbf{factor}(\alpha' \entails \tau' \cdot \J{left}) = \tau_l
-      \\
-      \textbf{factor}(\alpha' \entails \tau' \cdot \J{right}) = \tau_r
-    } {
-      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
-      \hastype 
-      \tau_l \J{->} \tau_r
-      \given \Theta', \Delta'
+      \alpha_l \wr^+ \tau_r'
+      \\
+      \textbf{found}(\alpha \entails \tau_l) = \tau_l'
+      \\
+      \tau_r'[\alpha_l\slash \J{LFP[} \alpha \J{]} \tau_l'] = \tau_r''
+    } {
+      \Theta^\dagger \entails \alpha \circlearrowleft (\Theta, \Delta, \alpha_l \J{->}\tau_r) \subtypes \alpha_l \J{->} \J{LFP[} \alpha \J{]} \tau_r''
     }
@@
 to infer as many types as possible
-and consolidates the results into a sequence of locales \ms{\Pi}.
+and consolidates the results into a sequence of zones \ms{\Pi}.
 It must consolidate all of the inferred types in order to
@@
 \ms{\textbf{pack}^+(... \entails \Pi) = \tau}
-to rewrite each locale into a type,
+to rewrite each zone into a type,
 such that the skolems and placeholders are converted to
@@
 First it leverages typing consolidation \ms{\Theta, \Delta, \Gamma \entails e \hastype \Pi}
-to infer as many possible locales for the source expression \ms{e}.
-Then it packs all the locales together and ensures that the packed type
+to infer as many possible zones for the source expression \ms{e}.
+Then it packs all the zones together and ensures that the packed type
 subtypes the type annotation.
@@
 form compatible with looping and 
-to collect as many type locales of its paths as possible.  
-It leverage locale sequence duality 
+to collect as many type zones of its paths as possible.  
+It leverage zone sequence duality 
 \ms{
@@
   \alpha_\mu \uparrow \Pi_\mu
-}, which rewrites the implicitly co-recursive path locales into
-implicitly recursive pair locales. 
+}, which rewrites the implicitly co-recursive path zones into
+implicitly recursive pair zones. 
 Then, it leverages negative packing \ms{\textbf{pack}^-(... \entails \Pi_\mu) = \tau_\mu},
@@
     \inferrule {
-      \Theta, \Delta \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
+      % inflatable iff founded
+      \Theta, \Delta \inflatable \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
       \\
@@
 \ms{
-... \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
+... \inflatable \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
 } and then solve the inflated introduction.
@@
 \begin{requirement}
-  \label{req:locale_sequence_duality_correspondence}
-  \emph{locale sequence duality correspondence}
+  \label{req:zone_sequence_duality_correspondence}
+  \emph{zone sequence duality correspondence}
   \footnotesize
@@
 \noindent
-The requirement of locale sequence duality has an ungodly amount of symbolic clutter,
+The requirement of zone sequence duality has an ungodly amount of symbolic clutter,
 but the essence of its meaning is actually quite simple and natural.
-It says that a function inhabits all locales in some 
+It says that a function inhabits all zones in some 
 some collection if and only if for any input/ouput pair, if the pair inhabits some
-locale from the dual collection, then the function applied to the input
+zone from the dual collection, then the function applied to the input
 gives the output. 
-The requirement is essential to ensure that rewriting inferred locales
+The requirement is essential to ensure that rewriting inferred zones
 to construct least fixed point types does not violate soundness.

### 2025-08-31 src/test_analysis_expressivity.py
update upper founded

@@ -1142,7 +1142,8 @@ if __name__ == '__main__':
     # test_succ_stream()
     # test_inf_loop()
     # test_succ_stream()
-    test_induction_even_is_nat_not_three()
+    # test_induction_even_is_nat_not_three()
+    test_succ_stream()
 
 
 #######################################################################

### 2025-09-04 doc/extrinsic_type_inference_paper/paper.tex
update checking interpretation in model typing
add notes on decidable check procedure
update subtyping check and tidy and interpret definitions in lean
update definitions to ensure completeness of negated predicates
update upper founded
define height and structural wrap
update struct less than
in paper, rename experiements to instances, since these are proved, not just externally observed

@@
 
-\newtheorem{experiment}{Experiment}[section]
+\newtheorem{instances}{Instances}[section]
 
@@
       \up
-      \delta \oplus \delta' \satisfies \Delta 
+      \delta' \oplus \delta \satisfies \Delta 
       \up
-      \delta \oplus \delta' \satisfies e \hastype \tau
+      \delta' \oplus \delta \satisfies e \hastype \tau
 
@@
       \implies
-      \delta \oplus \delta' \satisfies \Delta 
+      \delta' \oplus \delta \satisfies \Delta 
       \implies
-      \delta \oplus \delta' \satisfies e \hastype \tau
+      \delta' \oplus \delta \satisfies e \hastype \tau
 
@@
       &\iff& 
-      \alpha \slash \tau \in \delta
-      \up
-      \VDash e \hastype \tau
+      \VDash e \hastype \delta(\alpha) 
   \end{array}
@@
       \\
-      \textbf{found}(\alpha \entails \tau_l) = \tau_l'
+      % founded upper bound
+      \alpha \downarrow \tau_l \subtypes \tau_l'
       \\
@@
       \\\\
+      \TODO{the pattern restriction on both sides is more restrictive than subtyping restriction}
+      \\\\
       \epsilon \pattype \tau_r
       \\
-      \Theta, \Delta \restricted \tau_r \subtypes  \tau
-      \\\\
       \Theta, \Delta \not\dentails \tau \subtypes \tau_r
@@
       \given \Theta', \Delta'
+    }
+
+    \inferrule {
+      \TODO{proof rules should not require induction to construct premises}
+      \\\\
+      \epsilon \pattype \tau_r
+      \\
+      \alpha \prec \tau
+      \\
+      \alpha \wr^+ \tau
+      \\\\
+      \Theta, \Delta \not\dentails \tau[\alpha\slash\J{BOT}]^1 \subtypes \tau_r
+      \\
+      \Theta, \Delta \not\dentails \tau_r \subtypes \tau[\alpha\slash\J{BOT}]^{\textbf{height}(\tau_r)}
+      \\
+      \Theta, \Delta \entails \tau \subtypes \tau_l \given \Theta', \Delta' 
+    } {
+      \Theta, \Delta \entails 
+      \J{LFP[} \alpha \J{]} \tau \subtypes \tau_l \typdiff \tau_r
+      \given \Theta', \Delta'
+    }
+
+    \inferrule {
+      \TODO{meta-theory}
+      \\\\
+      \epsilon \pattype \tau_r
+      \\
+      \alpha \prec \tau % struct-less-than
+      \\
+      \alpha \wr^+ \tau
+      \\
+    } {
+      \Theta, \Delta \not\dentails \tau[\alpha\slash\J{BOT}]^1 \subtypes \tau_r
+      \implies
+      \nexists n \qua \Theta, \Delta \dentails \tau[\alpha\slash\J{BOT}]^n \subtypes \tau_r
+    }
+
+    \inferrule {
+      \TODO{meta-theory}
+      \\\\
+      \epsilon \pattype \tau_r
+      \\
+      \alpha \prec \tau
+      \\
+      \alpha \wr^+ \tau
+      \\
+    } {
+      \Theta, \Delta \not\dentails \tau_r \subtypes \tau[\alpha\slash\J{BOT}]^{\textbf{height}(\tau_r)}
+      \implies
+      \nexists n \qua \Theta, \Delta \dentails \tau_r \subtypes \tau[\alpha\slash\J{BOT}]^n
     }
@@
 
-\section{Theory}
-\label{sec:theory}
+\section{General Theory}
+\label{sec:general_theory}
 
@@
 
-\section{Experiments}
-\label{sec:experiments}
-
-\noindent
-We performed experiments to see how type inference performs on both functions and constructors. 
+\section{Instance Theory}
+\label{sec:}
+
+\noindent
+From the declarative definitions, we extracted procedures to infer types and 
+solve subtyping.
+We constructed tactics to automatically prove 
+static typing on instances of expressions and their inferred types.
+
+We constructed and proved instances with both functions and constructors. 
 These two categories have historically been tackled independently of each other, where parametric types
@@
 that SuperF was evaluated on, as presented in the paper on SuperF \cite{}.
-Based on these experiments, our system all programs admitted by SuperF, as well as some programs 
+Based on these instances, our system admits all reported be admitted by SuperF \cite{}, as well as some programs 
 that SuperF is unable to admit.
@@
 \begin{definition} 
-  \label{def:experimental_expressions_and_types}
-  \emph{experimental expressions and types} 
+  \label{def:expressions_and_type_instances}
+  \emph{expression and type instances} 
   \scriptsize
@@
 \begin{definition} 
-  \label{def:experimental_typing_context}
-  \emph{experimental typing context}
+  \label{def:tying_context_instance}
+  \emph{tying context instance}
   \tiny
@@
 \noindent
-Additionally, for the purpose of experiment, it is useful to assume that   
+Additionally, it is illustrative to assume that   
 the typing of some identifiers have already been learned or verified.
 
-\begin{experiment}
+\begin{instances}
   \label{exp:part_1}
@@
   \end{center}
-\end{experiment}
-
-\noindent
-In the experiment on functions, we 
+\end{instances}
+
+\noindent
+In the instances with functions, we 
 are interested to observe if our implementation (\textbf{Extrinsic})
@@
 
-\begin{experiment}
+\begin{instances}
   \label{experi:part_2}
@@
   \end{center}
-\end{experiment}
-
-\begin{experiment}
+\end{instances}
+
+\begin{instances}
   \label{experi:part_3}
@@
   \end{center}
-\end{experiment}
-
-\begin{experiment}
+\end{instances}
+
+\begin{instances}
   \label{experi:part_4}
@@
   \end{center}
-\end{experiment}
-
-\begin{experiment}
+\end{instances}
+
+\begin{instances}
   \label{experi:part_5}
@@
   \end{center}
-\end{experiment}
+\end{instances}
 
@@
 \noindent
-The experiment on programs with constructors demonstrate
+The instances with constructors demonstrate
 that our system is roughly as expressive as untyped languages

### 2025-12-23 doc/extrinsic_type_inference_paper/paper.tex
update static subtyping proof automation
update static subtyping solver for lfp elim
add proofs for union intro, inter elim, and universal elim
add soundness statement and update naming convetion
update lfp case to use full dynamic recursive call to ensure expressiveness
update annotation case for typing soundness
update stream case of loop typing soundness
restructure loop typing soudness with substance lemma
update notes on completeness nomenclature
add note on semantics-directed typing

@@
       &\iff& 
-      \alpha \wr^+ \tau
+      \alpha \wr \tau
       \up
-      \exists n \qua \VDash e \hastype (\tau[\alpha \slash \J{BOT}]^n)[\delta]
+      \exists \tau' \qua
+        \delta \satisfies \tau' \subtypes \tau \cdot \alpha \slash \tau'
+        \up
+        \alpha \slash \tau'\ \delta \satisfies e \hastype \tau
 
@@
 \end{definition}
+
+\TODO{
+RATIONALE -- SEMANTICS-DIRECTED TYPING:
+note that we use semantics-directed typing (i.e. dynamic) rather than syntax-directed typing (i.e. static), in order
+to avoid overly restricting the types of constructions like functions or records.
+That is, in a syntax-directed semantics, we strive to have functions 
+introduce the strongest type possible, which requires combining numerous type forms,
+and parsing the structures of expressions.
+To ensure the semantics of type forms is clear and intuitive, 
+we use sementics-directed typing, which allows
+each type form to be specified separately.
+In essence, the semantics-directed typing is a denotation
+from type syntax to predicate logic.
+By contrast, simple type systems are so simple that the syntax-directed typing definitions
+are intuitive already, and so there is no need for a separate semantics-directed definition.
+}
 
@@
       \\
-      \Theta, \Delta, \Gamma \entails e \hastype \Pi
-      \\
-      \textbf{pack}^+(\textbf{ftv}(\Delta) \entails \Pi) = \tau_e
-      \\
-      \exists \Theta', \Delta' \qua \Theta, \Delta \entails \tau_e \subtypes \tau_a \given \Theta', \Delta'
-    } {
-      \Theta, \Delta, \Gamma \entails e \J{ as } \tau_a \hastype \tau_a \given \Theta', \Delta'
+      \Theta, \Delta, \Gamma \entails e \hastype \tau_e \given \Theta', \Delta'
+      \\
+      \Theta', \Delta' \entails \tau_e \subtypes \tau_a \given \Theta'', \Delta''
+    } {
+      \Theta, \Delta, \Gamma \entails e \J{ as } \tau_a \hastype \tau_a \given \Theta'', \Delta''
     }
@@
       \\\\
-      \alpha_l \wr^+ \tau_r'
+      \alpha_l \wr^+ \J{LFP[} \alpha \J{]}\tau_r'
       \\
@@
       \\
-      \tau_r'[\alpha_l\slash \J{LFP[} \alpha \J{]} \tau_l'] = \tau_r''
-    } {
-      \Theta^\dagger \entails \alpha \circlearrowleft (\Theta, \Delta, \alpha_l \J{->}\tau_r) \subtypes \alpha_l \J{->} \J{LFP[} \alpha \J{]} \tau_r''
+      (\J{LFP[} \alpha \J{]}\tau_r')[\alpha_l\slash \J{LFP[} \alpha \J{]} \tau_l'] = \tau_r''
+    } {
+      \Theta^\dagger \entails \alpha \circlearrowleft (\Theta, \Delta, \alpha_l \J{->}\tau_r) \subtypes \alpha_l \J{->} \tau_r''
     }
@@
     \inferrule {
+      \textbf{factor}(\alpha \entails \tau_l \cdot l) = \tau
+      \\
       \Theta, \Delta \entails 
-      \textbf{factor}(\alpha \entails \tau_l \cdot l) 
+      \J{LFP[}\alpha\J{]} \tau
       \subtypes 
@@
     \inferrule {
+      \TODO{maybe learned subtypings on bound variabes can be removed}
+      \\\\
       \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
@@
 
-    \inferrule {
-      \Theta, \Delta \entails \textbf{merge}(\tau_{l}\J{\&}\tau_{r}) \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
-    } {
-      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
-    }
+    %% TODO: remove merge rule. Should be subsumed by antecedent union
+    % \inferrule {
+    %   \Theta, \Delta \entails \textbf{merge}(\tau_{l}\J{\&}\tau_{r}) \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
+    % } {
+    %   \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
+    % }
 
@@
 
+\TODO{consider replacing "substance" with some variation on "conditional completemness"}
 \begin{theorem}
@@
     } {
-      \forall \delta \qua \textbf{dom}(\delta) \subseteq \Theta \implies
-      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta
+      \delta \satisfies \Delta
       \implies
@@
     \inferrule {
+      \TODO{switch ordering of assignment maps}
+      \\\\
+      \TODO{look up from left to right}
+      \\\\
       \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
